### 2018-10-31 abstract/paper.tex
starting

# Towards practical probabilistic programming

We present our view on practical probabilistic programming, as
well as an inference facility for Go programs following the
view. Probabilistic programming promises to free the
practitioner from burden of re-implementing inference paradigms
and algorithms. There are dozens of probabilistic programming
languages currently in development. Yet, integrating
probabilistic programming inference into a software system
requires greater effort than writing a custom simulation and
inference algorithm. Probabilistic programming remains, at best,
a tool for modeling and prototyping.

We believe that the probabilistic programming approach can be
made practical. However, this requires a shift from design of
languages best suited for probabilistic inference to design
of tools and libraries which make probabilistic programming
in modern general-purpose languages accessible and convenient.

### 2018-10-31 abstract/paper.tex
starting

@@
-# Towards practical probabilistic programming
+\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
 
-We present our view on practical probabilistic programming, as
+\acmConference[LAFI'2019]{Languages for Inference}{January 15,
+2019}{Lisbon, Portugal}
+
+\setcopyright{none}
+
+\bibliographystyle{ACM-Reference-Format}
+
+\begin{abstract}
+We present a view on practical probabilistic programming, as
 well as an inference facility for Go programs following the
@@
 made practical. However, this requires a shift from design of
-languages best suited for probabilistic inference to design
-of tools and libraries which make probabilistic programming
-in modern general-purpose languages accessible and convenient.
+languages best suited for probabilistic inference to design of
+tools and libraries which make probabilistic programming in
+modern general-purpose languages accessible and convenient.
+Here, we first discuss the challenges in practical use of
+existing probabilistic programming systems, and then outline the
+guidelines to implementation of a practical probabilistic
+programming facility. We illustrate these guidelines on an
+early-stage implementation of probabilistic programming facility
+for the Go programming language. 
+\end{abstract}
+
+\begin{document}
+
+\title{Towards practical probabilistic programming}
+\author{David Tolpin}
+\affiliation{
+    \institution{PUB+}
+    \country{Israel}
+}
+\email{david.tolpin@gmail.com}
+
+\maketitle
+
+\section{Challenges in Practice of Probabilistic Programming}
+
+\section{Guidelines to a Practical Probabilistic Programming
+Facility}
+
+\section{Case Study: Probabilistic Programming Facility for
+Go}
+
+\bibliography{refs}
+
+\end{document}

### 2018-10-31 abstract/paper.tex
in progres

@@
 \documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\usepackage{url}
+\usepackage{hyperref}
 
@@
 Here, we first discuss the challenges in practical use of
-existing probabilistic programming systems, and then outline the
+existing probabilistic programming systems, and then suggest 
 guidelines to implementation of a practical probabilistic
-programming facility. We illustrate these guidelines on an
+programming facility. We illustrate the guidelines on an
 early-stage implementation of probabilistic programming facility
@@
 
-\title{Towards practical probabilistic programming}
+\title{Towards Practical Probabilistic Programming}
 \author{David Tolpin}
@@
 
+Incorporating a probabilistic program, or rather a probabilistic
+procedure, within a larger code body appears to be rather
+straightforward: one implements the model in the probabilistic
+programming language, fetches and preprocesses the data in the
+`regular' host programming language, passes the data and the
+model to an inference algorithm, and post-processes the results
+in the `regular' programming language again to make algorithmic
+decisions based on inference outcomes. The reality is more
+complicated though, and the practitioner may face obstacles on
+each step of the above routine.
+
+\paragraph{Implementing the model} Probabilistic programming is
+supposed to let the programmer use the expressive power of a
+general programming language to specify probabilistic models.
+However, the probabilistic language is often either a small
+language lacking advanced control and data manipulation
+capabilities (Anglican~\cite{TMY+16}, Stan~\cite{Stan17},
+WebPPL~\cite{GS15}),
+or an API to the representation of the probabilistic model
+(Edward~\cite{THS+17}, Figaro~\cite{P09}). When the probabilistic
+language is the same or as rich as the host language, it's use
+is still constrained by available data structures
+(Pyro~\cite{Pyro18}), or, the host language is itself uncommon as a
+general-purpose language (Turing~\cite{GXG18}). Often, the
+probabilistic programming language is interpreted, making
+algorithm execution of the model prohibitively slow.
+
+\paragraph{Passing the data} The case when all observations are
+handed over to inference in bulk, and a short summary of
+inference results returned, can mostly be implemented
+efficiently. This use case is well addressed by most
+environments even when the host language differs from the
+probabilistic language. A difficulty arises when
+the data arrives incrementally, and inference outcomes are
+processed in any-time manner. Data has to be marshalled when the
+host and probabilistic language platforms differ (e.g.
+Stan), or transferred between data structures used in
+the host language and the data structures required by the
+probabilistic language (e.g. Pyro).
+
+\paragraph{Simulation-inference dichotomy} Many probabilistic
+models follow a design pattern of simulation-inference: a
+significant part of the model is a simulator, running an
+algorithm with fixed parameter; the optimal parameters, or their
+distribution, are to be inferred. The simulator may be quite
+complicated, however in addition to the apparent difficulty of
+implementing the simulator in a less expressive language,  the
+simulator has to be implemented twice: first in the
+probabilistic language for the inference, and then in the host
+environment to make use of the inferred parameters. This not
+just involves extra effort but is also prone to introducing
+programming mistakes, when the two versions of the simulator
+behave differently for the same parameter values. One possible
+solution is to treat the simulator as a black box~\cite{LBW17},
+however better inference algorithms can be used if the structure
+of the simulator is revealed to the probabilistic programming
+facility, e.g. through the use of automatic differentiation.
+
+\paragraph{Deployment} A software system composed of
+heterogeneous languages and environments is harder to integrate
+and deploy. Many software solutions are monolithic, comprising
+of just a single programming language and with a conservative
+set of dependencies. Adding a probabilistic programming system,
+often depending on a different runtime and libraries, makes
+integration and deployment more complicated.
+
 \section{Guidelines to a Practical Probabilistic Programming
 Facility}
+
+Based on the experience of implementing  probabilistic
+programming languages, as well as developing and deploying solutions
+using different probabilistic environments, we suggest
+the following guidelines to implementation of a probabilistic
+programming facility. We believe that these guidelines, if
+followed, will lead to eventual wide adoption of probabilistic
+programming in software development practice.
+
+\begin{enumerate}
+\item A probabilistic model should be programmed in the host
+programming language. The facility may impose a discipline on
+model implementation, such as through interface constraints, but
+otherwise supporting unrestricted use of the host language for
+implementation of the model.
+
+\item Built-in and user-defined data structures and libraries
+should be accessible in the probabilistic programming model.
+Inference techniques relying on the code structure, such as
+those based on automatic differentiation, should support the
+use of common data structures of the host language.
+
+\item The model code should be reusable between inference and
+simulation. The code which is not required solely for inference
+(simulation) should be written once for both inference of
+parameters and use of the parameters in the host environment.
+It should be possible to simulation outside the probabilistic
+model without runtime or memory overhead imposed by inference
+needs.
+\end{enumerate}
 
@@
 
+Following the guidelines, we have implemented a probabilistic
+programming facility for the Go programming language,
+\texttt{infergo} (\url{http://infergo.ml/}). At the time of
+writing, the implementation is still at an early stage. However,
+this facility is already used in production environment to infer
+the website visitor's value using a hierarchical mixed effects
+model.
+
 \bibliography{refs}

### 2018-10-31 abstract/paper.tex
in progress

@@
 simulation. The code which is not required solely for inference
-(simulation) should be written once for both inference of
-parameters and use of the parameters in the host environment.
-It should be possible to simulation outside the probabilistic
-model without runtime or memory overhead imposed by inference
-needs.
+should be written once for both inference of parameters and use
+of the parameters in the host environment.  It should be
+possible to run simulation outside the probabilistic model without
+runtime or memory overhead imposed by inference needs.
 \end{enumerate}

### 2018-10-31 abstract/paper.tex
finished and squeezed

@@
 programming facility. We illustrate the guidelines on an
-early-stage implementation of probabilistic programming facility
-for the Go programming language. 
+early-stage implementation of such facility for the Go
+programming language. 
 \end{abstract}
@@
 straightforward: one implements the model in the probabilistic
-programming language, fetches and preprocesses the data in the
+language, fetches and preprocesses the data in the
 `regular' host programming language, passes the data and the
@@
 capabilities (Anglican~\cite{TMY+16}, Stan~\cite{Stan17},
-WebPPL~\cite{GS15}),
-or an API to the representation of the probabilistic model
-(Edward~\cite{THS+17}, Figaro~\cite{P09}). When the probabilistic
-language is the same or as rich as the host language, it's use
-is still constrained by available data structures
-(Pyro~\cite{Pyro18}), or, the host language is itself uncommon as a
-general-purpose language (Turing~\cite{GXG18}). Often, the
-probabilistic programming language is interpreted, making
-algorithm execution of the model prohibitively slow.
+WebPPL~\cite{GS15}), or an API to the representation of the
+probabilistic model (Edward~\cite{THS+17}, Figaro~\cite{P09}).
+When the probabilistic language is the same or as rich as the
+host language, its use is still constrained by available data
+structures (Pyro~\cite{Pyro18}), or, the host language is 
+uncommon as a general-purpose language (Turing~\cite{GXG18}).
+Often, the probabilistic language is interpreted, making
+execution of the model too slow.
 
@@
 simulator has to be implemented twice: first in the
-probabilistic language for the inference, and then in the host
+probabilistic language for inference, and then in the host
 environment to make use of the inferred parameters. This not
@@
 Based on the experience of implementing  probabilistic
-programming languages, as well as developing and deploying solutions
-using different probabilistic environments, we suggest
-the following guidelines to implementation of a probabilistic
-programming facility. We believe that these guidelines, if
-followed, will lead to eventual wide adoption of probabilistic
-programming in software development practice.
+programming languages, as well as developing and deploying
+solutions using different probabilistic environments, we suggest
+guidelines to implementation of a probabilistic programming
+facility. We believe that these guidelines, if followed, will
+lead to eventual wide adoption of probabilistic programming in
+software engineering.
 
@@
 
-\section{Case Study: Probabilistic Programming Facility for
+\section{Probabilistic Programming Facility for
 Go}
@@
 programming facility for the Go programming language,
-\texttt{infergo} (\url{http://infergo.ml/}). At the time of
-writing, the implementation is still at an early stage. However,
-this facility is already used in production environment to infer
-the website visitor's value using a hierarchical mixed effects
-model.
+\texttt{infergo} (\url{http://infergo.ml/}). We have chosen Go
+because Go is a relatively small programming language with
+efficient implementation, which has recently become quite
+popular for computation-intensive server-side programming. At
+the time of writing, the implementation is still at an early
+stage. However, this facility is already used in production
+environment for inference of mission-critical algorithm
+parameters.
+
+A probabilistic model in \texttt{infergo} is an implementation
+of the \texttt{Model} interface requiring a single method
+\texttt{Observe} which accepts a vector (a Go \textit{slice}) of
+floats, the parameters to infer, and returns a single float,
+interpreted as unnormalized log-likelihood of the posterior
+distribution. The implementation of the methods can be written
+in virtually unrestricted Go and use any Go libraries.
+
+For inference, \texttt{infergo} relies on automatic
+differentiation. The source code of the model is
+translated by a command-line tool provided by \texttt{infergo}
+into an equivalent model with reverse-mode automatic
+differentiation of the log-likelihood with respect 
+to the parameters applied. The differentiation operates
+on the built-in floating-point type and incurs only a small
+computational overhead. However, even this overhead is avoided
+when the model code is executed outside of inference algorithms:
+both the original and the differentiated model are
+simultaneously available to the rest of the program code, so
+the methods can be called on the differentiated model for
+inference, and on the original model for the most efficient
+execution with the inferred parameters.
+
+While we enjoy the ease and efficiency of
+\texttt{infergo}-based inference in our production environment,
+we view \texttt{infergo} as a reference implementation rather than
+as the only solution, and encourage the community to equip other
+general-purpose languages with similar inference facilities,
+which, we believe, will make wide adoption of probabilistic
+programming more likely.
+be
 

### 2018-10-31 abstract/paper.tex
another word

@@
 
-While we enjoy the ease and efficiency of
-\texttt{infergo}-based inference in our production environment,
-we view \texttt{infergo} as a reference implementation rather than
-as the only solution, and encourage the community to equip other
-general-purpose languages with similar inference facilities,
-which, we believe, will make wide adoption of probabilistic
-programming more likely.
-be
+While we enjoy the ease and efficiency of \texttt{infergo}-based
+inference in our production environment, we view
+\texttt{infergo} as a reference implementation rather than as
+the only recommended solution, and encourage the community to
+equip other general-purpose languages with similar inference
+facilities, which, we believe, will make wide adoption of
+probabilistic programming more likely.
 

### 2018-11-01 abstract/paper.tex
positive

--- 2018-10-31 abstract/paper.tex
@@
 
-\begin{abstract}
-We present a view on practical probabilistic programming, as
-well as an inference facility for Go programs following the
-view. Probabilistic programming promises to free the
-practitioner from burden of re-implementing inference paradigms
-and algorithms. There are dozens of probabilistic programming
-languages currently in development. Yet, integrating
-probabilistic programming inference into a software system
-requires greater effort than writing a custom simulation and
-inference algorithm. Probabilistic programming remains, at best,
-a tool for modeling and prototyping.
-
-We believe that the probabilistic programming approach can be
-made practical. However, this requires a shift from design of
-languages best suited for probabilistic inference to design of
-tools and libraries which make probabilistic programming in
-modern general-purpose languages accessible and convenient.
-Here, we first discuss the challenges in practical use of
-existing probabilistic programming systems, and then suggest 
-guidelines to implementation of a practical probabilistic
-programming facility. We illustrate the guidelines on an
-early-stage implementation of such facility for the Go
-programming language. 
-\end{abstract}
-
-\begin{document}
-
-\title{Towards Practical Probabilistic Programming}
+\title{Server-Side Probabilistic Programming}
 \author{David Tolpin}
@@
 
+\begin{abstract}
+	We share our experiences from integrating probabilistic
+	programming into a server-side software system and
+	present our ongoing work on probabilistic programming
+	facility for Go, a modern programming language of choice for
+	server-side software development. We demonstrate how a
+	lightweight efficient probabilistic programming facility can
+	be added to an existing programming language.  Server-side
+	application of probabilistic programming poses challenges
+	for a probabilistic programming system. We discuss the
+	challenges and our experience in overcoming them, and
+	suggest guidelines which can help in a wider adoption of
+	probabilistic programming in server-side software systems.
+\end{abstract}
+
+\begin{document}
 \maketitle
 
-\section{Challenges in Practice of Probabilistic Programming}
+\section{Challenges of Server-Side Probabilistic Programming}
 
@@
 straightforward: one implements the model in the probabilistic
-language, fetches and preprocesses the data in the
-`regular' host programming language, passes the data and the
-model to an inference algorithm, and post-processes the results
-in the `regular' programming language again to make algorithmic
-decisions based on inference outcomes. The reality is more
-complicated though, and the practitioner may face obstacles on
-each step of the above routine.
+programming language, fetches and preprocesses the data in the
+host programming language, passes the data and the model to an
+inference algorithm, and post-processes the results in the
+host programming language again to make algorithmic
+decisions based on inference outcomes. However, complex
+server-side software systems make integration of probabilistic
+inference challenging.
 
-\paragraph{Implementing the model} Probabilistic programming is
-supposed to let the programmer use the expressive power of a
-general programming language to specify probabilistic models.
-However, the probabilistic language is often either a small
-language lacking advanced control and data manipulation
-capabilities (Anglican~\cite{TMY+16}, Stan~\cite{Stan17},
-WebPPL~\cite{GS15}), or an API to the representation of the
-probabilistic model (Edward~\cite{THS+17}, Figaro~\cite{P09}).
-When the probabilistic language is the same or as rich as the
-host language, its use is still constrained by available data
-structures (Pyro~\cite{Pyro18}), or, the host language is 
-uncommon as a general-purpose language (Turing~\cite{GXG18}).
-Often, the probabilistic language is interpreted, making
-execution of the model too slow.
+\paragraph{Simulation-inference dichotomy} Probabilistic models
+often follow the design pattern of simulation-inference: a
+significant part of the model is a simulator, running an
+algorithm with fixed parameters; the optimal parameters, or
+their distribution, are to be inferred. The inferred parameters
+are then used by the software system to execute the simulation
+independently of inference for forecasting and decision making.
 
-\paragraph{Passing the data} The case when all observations are
-handed over to inference in bulk, and a short summary of
-inference results returned, can mostly be implemented
-efficiently. This use case is well addressed by most
-environments even when the host language differs from the
-probabilistic language. A difficulty arises when
-the data arrives incrementally, and inference outcomes are
-processed in any-time manner. Data has to be marshalled when the
-host and probabilistic language platforms differ (e.g.
-Stan), or transferred between data structures used in
-the host language and the data structures required by the
-probabilistic language (e.g. Pyro).
+This pattern suggest re-use of the simulator: instead of
+implementing the simulator twice, in the probabilistic model and
+in the host system, the same code can serve both purposes.
+However to achieve this, the host language must coincide with
+the implementation language of the probabilistic model, on one
+hand, and allow a computationally efficient implementation of
+the simulation, on the other hand. Some probabilistic systems
+(Figaro~\cite{P09}, Anglican~\cite{TMY+16}, Turing~\cite{GXG18})
+are built with tight integration with the host environment in
+mind; more often than not though the probabilistic code is
+not trivial to re-use.
 
-\paragraph{Simulation-inference dichotomy} Many probabilistic
-models follow a design pattern of simulation-inference: a
-significant part of the model is a simulator, running an
-algorithm with fixed parameter; the optimal parameters, or their
-distribution, are to be inferred. The simulator may be quite
-complicated, however in addition to the apparent difficulty of
-implementing the simulator in a less expressive language,  the
-simulator has to be implemented twice: first in the
-probabilistic language for inference, and then in the host
-environment to make use of the inferred parameters. This not
-just involves extra effort but is also prone to introducing
-programming mistakes, when the two versions of the simulator
-behave differently for the same parameter values. One possible
-solution is to treat the simulator as a black box~\cite{LBW17},
-however better inference algorithms can be used if the structure
-of the simulator is revealed to the probabilistic programming
-facility, e.g. through the use of automatic differentiation.
+\paragraph{Data interface} In a server-side application data
+for inference comes from a variety of sources: network,
+databases, distributed file systems, and in many different
+formats. Efficient inference depends on fast data access and
+updating. Libraries for data access and manipulation are
+available in the host environment. While the host environment
+can be used as a proxy retrieving and transforming the data,
+such as in the case of Stan~\cite{Stan17} integrations,
+sometimes direct access from the probabilistic code is the
+preferred option, for example when the data is streamed or
+retrieved conditionally.
 
-\paragraph{Deployment} A software system composed of
-heterogeneous languages and environments is harder to integrate
-and deploy. Many software solutions are monolithic, comprising
-of just a single programming language and with a conservative
-set of dependencies. Adding a probabilistic programming system,
-often depending on a different runtime and libraries, makes
-integration and deployment more complicated.
+\paragraph{Integration and deployment} Deployment of server-side
+software systems is a delicate process involving automatic
+builds and maintenance of dependencies. Adding  a component
+which possibly introduces additional software dependencies or
+even a separate runtime complicates deployment. Minimizing the
+burden of probabilistic programming on the integration
+and deployment process should be a major consideration in
+design or selection of probabilistic programming tools.
+Probabilistic programming systems which are implemented or 
+provide an interface in a popular programming language, e.g.
+Python (Edward~\cite{THS+17}, Pyro~\cite{Pyro18}) are easier
+to integrate and deploy, however the smaller the footprint
+of a probabilistic system, the easier is the adoption.
 
-\section{Guidelines to a Practical Probabilistic Programming
-Facility}
+\section{Probabilistic Programming Facility for Go}
 
-Based on the experience of implementing  probabilistic
-programming languages, as well as developing and deploying
-solutions using different probabilistic environments, we suggest
+Based on the experience of developing and deploying solutions
+using different probabilistic environments, we came up with
 guidelines to implementation of a probabilistic programming
-facility. We believe that these guidelines, if followed, will
-lead to eventual wide adoption of probabilistic programming in
-software engineering.
+facility for server-side applications. We believe that these
+guidelines, when followed, help easier integration of
+probabilistic programming inference into large-scale server-side
+software systems.
 
@@
 
-\section{Probabilistic Programming Facility for
-Go}
-
-Following the guidelines, we have implemented a probabilistic
+In line with the guidelines, we have implemented a probabilistic
 programming facility for the Go programming language,
 \texttt{infergo} (\url{http://infergo.ml/}). We have chosen Go
-because Go is a relatively small programming language with
+because Go is a small but expressive programming language with
 efficient implementation, which has recently become quite
@@
 
-While we enjoy the ease and efficiency of
-\texttt{infergo}-based inference in our production environment,
-we view \texttt{infergo} as a reference implementation rather than
-as the only solution, and encourage the community to equip other
-general-purpose languages with similar inference facilities,
-which, we believe, will make wide adoption of probabilistic
-programming more likely.
-be
+The Go programming language and development environment offer
+capabilities which made implementation of \texttt{infergo}
+affordable.
+
+\begin{enumerate}
+	\item The Go parser and abstract syntax tree serializer are a
+		part of the standard library. Reading, transforming, and
+		writing Go source code is straightforward and effortless.
+	\item Type inference (or \textit{type checking} as it is
+		called in the Go ecosystem), also provided in the
+		standard library, augments parsing and allows to
+		selectively apply transformation-based automatic
+		differentiation  based on static expression types. 
+	\item Go compiles and runs fast. Fast compilation and
+		execution speeds allow to use the same facility for both
+		exploratory design of probabilistic models and for
+		inference in production environment.
+	\item Go offers efficient parallel execution as a
+		first-class feature, via so-called \textit{goroutines}.
+		Goroutines streamline implementation of sampling-based
+		inference algorithms. Sample generators and consumers
+		are run in parallel, communicating through channels. 
+		Inference is easy to parallelize in order to exploit
+		hardware multi-processing, and samples are retrieved
+		lazily for postprocessing. 
+\end{enumerate}
+
+A lightweight probabilistic programming facility similar to
+\texttt{infergo} can be added to most modern general-purpose
+programming languages, in particular those used in implementing
+large-scale software systems, making probabilistic
+programming inference more accessible in server-side
+applications.
 

### 2018-11-01 abstract/paper.tex
cosmetic

@@
 	challenges and our experience in overcoming them, and
-	suggest guidelines which can help in a wider adoption of
+	suggest guidelines that can help in a wider adoption of
 	probabilistic programming in server-side software systems.
@@
 \paragraph{Simulation-inference dichotomy} Probabilistic models
-often follow the design pattern of simulation-inference: a
+often follow a design pattern of simulation-inference: a
 significant part of the model is a simulator, running an
@@
 implementing the simulator twice, in the probabilistic model and
-in the host system, the same code can serve both purposes.
+in the host environment, the same code can serve both purposes.
 However to achieve this, the host language must coincide with
@@
 software systems is a delicate process involving automatic
-builds and maintenance of dependencies. Adding  a component
+builds and maintenance of dependencies. Adding  a component,
 which possibly introduces additional software dependencies or
-even a separate runtime complicates deployment. Minimizing the
+even a separate runtime, complicates deployment. Minimizing the
 burden of probabilistic programming on the integration
@@
 design or selection of probabilistic programming tools.
-Probabilistic programming systems which are implemented or 
+Probabilistic programming systems that are implemented or 
 provide an interface in a popular programming language, e.g.
@@
 
-\begin{enumerate}
+\begin{itemize}
 	\item The Go parser and abstract syntax tree serializer are a
@@
 		lazily for postprocessing. 
-\end{enumerate}
+\end{itemize}
 

### 2018-11-01 abstract/paper.tex
cosmetic

@@
 
-This pattern suggest re-use of the simulator: instead of
+This pattern suggests re-use of the simulator: instead of
 implementing the simulator twice, in the probabilistic model and

### 2018-11-03 abstract/paper.tex
language edited

@@
 interpreted as unnormalized log-likelihood of the posterior
-distribution. The implementation of the methods can be written
+distribution. Implementation of model methods can be written
 in virtually unrestricted Go and use any Go libraries.
@@
 \begin{itemize}
-	\item The Go parser and abstract syntax tree serializer are a
-		part of the standard library. Reading, transforming, and
-		writing Go source code is straightforward and effortless.
+    \item The Go parser and abstract syntax tree serializer are
+        a part of the standard library. Parsing, transforming,
+        and generating Go source code is straightforward and
+        effortless.
 	\item Type inference (or \textit{type checking} as it is

### 2018-11-06 abstract/paper.tex
measurements

--- 2018-11-01 abstract/paper.tex
@@
 
+Table~\ref{tab:memory-runtime} provides memory consumption and
+running time measurements on basic models to
+illustrate  \texttt{infergo}'s performance.
+The measurements were obtained on a 2.3GHz Intel Core 5
+CPU with 8GB of memory for 1000 iterations of Hamiltonian Monte
+Carlo with 10 leapfrog steps. Note that computation of log-likelihood
+for standard distributions is not optimized yet. Quite the
+opposite: since models in \texttt{infergo} are fully composable,
+primitive distributions are themselves implemented as
+\texttt{infergo} models and automatically differentiated.
+
+\begin{table}[H]
+	\caption{Memory and running times for 1000 iterations
+	of HMC with 10 leapfrog steps.}
+	\label{tab:memory-runtime}
+	\begin{tabular}{r | c |  c | c}
+		{\it model}  & \multicolumn{2}{c|}{\it time} & {\it memory} \\ 
+		 & {\it compilation} & {\it execution} & \\\hline
+		8 schools & 0.15s & 0.6s & 5.5MB \\
+		10D normal, 100 points & 0.15s & 3.0s & 5.7MB \\
+		50D normal, 100 points & 0.15s & 13.0s & 5.8MB 
+	\end{tabular}
+\end{table}
+
 A lightweight probabilistic programming facility similar to

### 2018-11-06 abstract/paper.tex
formatted

@@
 \usepackage{hyperref}
+
+\linespread{1.03}
 
@@
 
+\begin{sloppypar}
 \section{Challenges of Server-Side Probabilistic Programming}
@@
 
-\paragraph{Simulation-inference dichotomy} Probabilistic models
+\paragraph{Simulation vs. inference} Probabilistic models
 often follow a design pattern of simulation-inference: a
@@
 CPU with 8GB of memory for 1000 iterations of Hamiltonian Monte
-Carlo with 10 leapfrog steps. Note that computation of log-likelihood
+Carlo with 10 leapfrog steps. Note that log-likelihood computation 
 for standard distributions is not optimized yet. Quite the
@@
 
+{\smaller
 \begin{table}[H]
@@
 	\end{tabular}
-\end{table}
+\end{table}}
 
@@
 applications.
+\end{sloppypar}
 

### 2019-04-15 paper/paper.tex
in progress, overall structure - fill it!

--- 2018-11-06 abstract/paper.tex
@@
-\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[sigplan,review,10pt,anonymous]{acmart}
 \usepackage{url}
@@
 
-\linespread{1.03}
+\acmConference[Onward! 2019]{SPLASH Onward! 2019}{October 20-25, 2019}{Athens, Greece}
 
-\acmConference[LAFI'2019]{Languages for Inference}{January 15,
-2019}{Lisbon, Portugal}
-
+% Copyright
 \setcopyright{none}
+%\setcopyright{acmcopyright}
+%\setcopyright{acmlicensed}
+%\setcopyright{rightsretained}
+%\setcopyright{usgov}
+%\setcopyright{usgovmixed}
+%\setcopyright{cagov}
+%\setcopyright{cagovmixed}
 
@@
 	We share our experiences from integrating probabilistic
-	programming into a server-side software system and
-	present our ongoing work on probabilistic programming
-	facility for Go, a modern programming language of choice for
-	server-side software development. We demonstrate how a
-	lightweight efficient probabilistic programming facility can
-	be added to an existing programming language.  Server-side
+	programming into a server-side software system and introduce
+	a probabilistic programming facility for Go, a modern
+	programming language of choice for server-side software
+	development. We demonstrate how a lightweight but powerful
+	and efficient probabilistic programming facility can be
+	added to an existing programming language.  Server-side
 	application of probabilistic programming poses challenges
@@
 \begin{sloppypar}
+
+\section{Introduction}
+
+The promise of probabilistic programming.
+
+The language war.
+
+A solution: probabilistic programming without a DSL.
+
+Organization of the paper.
+
+\subsection*{Contributions}
+
+\section{Related Work}
+
 \section{Challenges of Server-Side Probabilistic Programming}
@@
 
-\section{Probabilistic Programming Facility for Go}
+\section{Guidelines to Implementation of a Probabilistic Programming Facility}
 
@@
 
+\section{Probabilistic Programming in Go}
+
 In line with the guidelines, we have implemented a probabilistic
 programming facility for the Go programming language,
-\texttt{infergo} (\url{http://infergo.ml/}). We have chosen Go
-because Go is a small but expressive programming language with
-efficient implementation, which has recently become quite
-popular for computation-intensive server-side programming. At
-the time of writing, the implementation is still at an early
-stage. However, this facility is already used in production
-environment for inference of mission-critical algorithm
-parameters.
+Infergo. %(\url{http://infergo.ml/}).
+We have chosen Go because Go is a small but expressive
+programming language with efficient implementation, which has
+recently become quite popular for computation-intensive
+server-side programming.  the time of writing, the
+implementation is still at an early stage.  This facility
+ialready used in production environment for inference of
+mission-critical algorithm parameters.
 
-A probabilistic model in \texttt{infergo} is an implementation
-of the \texttt{Model} interface requiring a single method
-\texttt{Observe} which accepts a vector (a Go \textit{slice}) of
-floats, the parameters to infer, and returns a single float,
-interpreted as unnormalized log-likelihood of the posterior
-distribution. Implementation of model methods can be written
-in virtually unrestricted Go and use any Go libraries.
+\subsection{An Overview of Infergo}
 
-For inference, \texttt{infergo} relies on automatic
-differentiation. The source code of the model is
-translated by a command-line tool provided by \texttt{infergo}
-into an equivalent model with reverse-mode automatic
-differentiation of the log-likelihood with respect 
-to the parameters applied. The differentiation operates
-on the built-in floating-point type and incurs only a small
+A probabilistic model in Infergo is an implementation of an
+interface requiring a single method \texttt{Observe} which
+accepts a vector (a Go \textit{slice}) of floats, the parameters
+to infer, and returns a single float, interpreted as
+unnormalized log-likelihood of the posterior distribution.
+Implementation of model methods can be written in virtually
+unrestricted Go and use any Go libraries.
+
+For inference, Infergo relies on automatic differentiation. The
+source code of the model is translated by a command-line tool
+provided by Infergo into an equivalent model with reverse-mode
+automatic differentiation of the log-likelihood with respect to
+the parameters applied. The differentiation operates on the
+built-in floating-point type and incurs only a small
 computational overhead. However, even this overhead is avoided
@@
 both the original and the differentiated model are
-simultaneously available to the rest of the program code, so
-the methods can be called on the differentiated model for
-inference, and on the original model for the most efficient
-execution with the inferred parameters.
+simultaneously available to the rest of the program code, so the
+methods can be called on the differentiated model for inference,
+and on the original model for the most efficient execution with
+the inferred parameters.
 
 The Go programming language and development environment offer
-capabilities which made implementation of \texttt{infergo}
-affordable.
+capabilities which made implementation of Infergo affordable.
 
@@
 
+\subsection{Probabilistic ``Hello, World!''}
+
+TODO: Example
+
+\section{Model Syntax and Rules of Differentiation}
+
+\section{Design and Implementation Highlights}
+
+\subsection{Choice of Programming Language}
+
+\subsection{Automatic Differentiation}
+
+\subsection{Inference}
+
+\subsection{Composability of Models}
+
+\subsection{Stochasticity and Streaming}
+
+\section{Case Studies}
+
+\subsection{Performance}
+
 Table~\ref{tab:memory-runtime} provides memory consumption and
-running time measurements on basic models to
-illustrate  \texttt{infergo}'s performance.
-The measurements were obtained on a 2.3GHz Intel Core 5
-CPU with 8GB of memory for 1000 iterations of Hamiltonian Monte
-Carlo with 10 leapfrog steps. Note that log-likelihood computation 
-for standard distributions is not optimized yet. Quite the
-opposite: since models in \texttt{infergo} are fully composable,
-primitive distributions are themselves implemented as
-\texttt{infergo} models and automatically differentiated.
+running time measurements on basic models to illustrate
+Infergo's performance.  The measurements were obtained on a
+2.3GHz Intel Core 5 CPU with 8GB of memory for 1000 iterations
+of Hamiltonian Monte Carlo with 10 leapfrog steps. Note that
+log-likelihood computation for standard distributions is not
+optimized yet. Quite the opposite: since models in Infergo are
+fully composable, primitive distributions are themselves
+implemented as Infergo models and automatically differentiated.
 
@@
 
+\section{Conclusion}
+
 A lightweight probabilistic programming facility similar to
-\texttt{infergo} can be added to most modern general-purpose
-programming languages, in particular those used in implementing
-large-scale software systems, making probabilistic
-programming inference more accessible in server-side
-applications.
+Infergo can be added to most modern general-purpose programming
+languages, in particular those used in implementing large-scale
+software systems, making probabilistic programming inference
+more accessible in server-side applications.
+
+\section*{Acknowledgements}
+
 \end{sloppypar}

### 2019-04-15 paper/paper.tex
in progress

@@
 
-\title{Server-Side Probabilistic Programming}
+\title{Deployable Probabilistic Programming}
 \author{David Tolpin}
@@
 
-\section{Challenges of Server-Side Probabilistic Programming}
+\section{Challenges}
 
@@
 
-\section{Guidelines to Implementation of a Probabilistic Programming Facility}
+\section{Guidelines}
 
@@
 In line with the guidelines, we have implemented a probabilistic
-programming facility for the Go programming language,
-Infergo. %(\url{http://infergo.ml/}).
-We have chosen Go because Go is a small but expressive
-programming language with efficient implementation, which has
-recently become quite popular for computation-intensive
-server-side programming.  the time of writing, the
-implementation is still at an early stage.  This facility
-ialready used in production environment for inference of
-mission-critical algorithm parameters.
+programming facility for the Go programming language, Infergo.
+%(\url{http://infergo.ml/}). 
+We have chosen Go because Go is a
+small but expressive programming language with efficient
+implementation, which has recently become quite popular for
+computation-intensive server-side programming. Infergo is
+used in production environment for inference of mission-critical
+algorithm parameters.
 
@@
 
+\subsubsection{Optimization}
+
+\subsubsection{Full Posterior}
+
 \subsection{Composability of Models}
@@
 
+The author owes a debt of gratitude to Frank Wood who introduced
+the author to probabilistic programming and inspired to
+pursue probabilistic programming paradigms and applications. The
+author also would like to thank Jan-Willem van de Meent, with
+whom the author had fruitful discussions of motives, ideas, and
+implementation choices behind Infergo, and whose thoughts and
+recommendations significantly influenced Infergo design.
+Finally, the author thanks PUB+, the company the author works
+for, for supporting development of Infergo and letting the
+author experiment with applying probabilistic programming to
+critical decision-making in production environment.
+
 \end{sloppypar}

### 2019-04-16 paper/paper.tex
introduction in progress

@@
 
-The promise of probabilistic programming.
+	Probabilistic programming \cite{GMR+08,MSP14,WVM14,GS15} represents
+	statistical models as programs written in an otherwise
+	general programming language that provides syntax for the
+	definition and conditioning of random variables. Inference can be
+	performed on probabilistic programs to obtain the posterior
+	distribution or point estimates of the variables. Inference
+	algorithms are provided by the probabilistic programming system,
+	and each algorithm is usually applicable to a wide class of
+	probabilistic programs in a black box manner.  Inference
+	techniques provided by the system back end, such as
+	Metropolis-Hastings \cite{WSG11,MSP14,YHG14}, Hamiltonian Monte
+	Carlo \cite{SDT14}, expectation propagation \cite{MWG+10},
+	extensions of Sequential Monte Carlo \cite{WVM14,MYM+15,PWD+14},
+	variational inference~\cite{}, gradient-based
+	optimization~\cite{}, and others.
+
+	There are two polar views on the role of probabilistic programming.
+	One view is that probabilistic programming is a flexible tool for
+	specification of statistical models. Proponents of this view
+	probabilistic programming as a tool for data science
+	practitioners. A typical work flow consists of data acquisition
+	and pre-processing, followed by several iterations of exploratory
+	model design and testing of inference algorithms. Once a
+	sufficiently robust statistical model is obtained, analysis
+	results are post-processed, visualized, and occasionally
+	integrated into algorithms of a production software system.
+
+	The other, emerging, view is that probabilistic programming is an
+	extension to a regular programming toolbox allowing
+	algorithms implemented in general programming languages to
+	have learnable parameters. In this view, statistical 
+	models are integrated into the software system, and inference
+	and optimization take place during data processing, prediction,
+	and algorithmic decision making in production.  Probabilistic
+	programming code runs unattended, and inference results are an
+	integral part of the algorithms.
+
+	In this work we are concerned with the later view on
+	probabilistic programming. Our objective is to pave a path to
+	deployment of probabilistic programs in production systems,
+	promoting a wider adoption of probabilistic programming as a
+	flexible tool for ubiquitous statistical inference.
 
@@
 \subsection*{Contributions}
+
+\begin{itemize}
+	\item Guidelines for design and implementation of a
+		probabilistic programming facility deployable as a part 
+		of production software systems.
+	\item A reference implementation of deployable probabilistic
+		programming facility for the Go programming language.
+	\item Design and implementation highlights 
+\end{itemize}
 

### 2019-04-16 paper/paper.tex
better language
code in paper

@@
 \documentclass[sigplan,review,10pt,anonymous]{acmart}
+\usepackage[utf8]{inputenc}
 \usepackage{url}
 \usepackage{hyperref}
+\usepackage{courier}
+\usepackage{listings}[language=Golang]
+
+\lstdefinelanguage{Golang}%
+  {morekeywords=[1]{package,import,func,type,struct,return,defer,panic,%
+     recover,select,var,const,iota,},%
+   morekeywords=[2]{string,uint,uint8,uint16,uint32,uint64,int,int8,int16,%
+     int32,int64,bool,float32,float64,complex64,complex128,byte,rune,uintptr,%
+     error,interface},%
+   morekeywords=[3]{map,slice,make,new,nil,len,cap,copy,close,true,false,%
+     delete,append,real,imag,complex,chan,},%
+   morekeywords=[4]{for,break,continue,range,go,goto,switch,case,fallthrough,if,%
+     else,default,},%
+   morekeywords=[5]{Println,Printf,Error,Print,},%
+   sensitive=true,%
+   morecomment=[l]{//},%
+   morecomment=[s]{/*}{*/},%
+   morestring=[b]',%
+   morestring=[b]",%
+   morestring=[s]{`}{`},%
+}
+
+\lstset{ % add your own preferences
+    frame=none,
+    basicstyle=\footnotesize\ttfamily,
+    numbers=left,
+    numbersep=5pt,
+    showstringspaces=false,
+    tabsize=4,
+    language=Golang % this is it !
+}
 
@@
 
-	Probabilistic programming \cite{GMR+08,MSP14,WVM14,GS15} represents
-	statistical models as programs written in an otherwise
-	general programming language that provides syntax for the
-	definition and conditioning of random variables. Inference can be
-	performed on probabilistic programs to obtain the posterior
-	distribution or point estimates of the variables. Inference
-	algorithms are provided by the probabilistic programming system,
-	and each algorithm is usually applicable to a wide class of
-	probabilistic programs in a black box manner.  Inference
-	techniques provided by the system back end, such as
-	Metropolis-Hastings \cite{WSG11,MSP14,YHG14}, Hamiltonian Monte
-	Carlo \cite{SDT14}, expectation propagation \cite{MWG+10},
-	extensions of Sequential Monte Carlo \cite{WVM14,MYM+15,PWD+14},
-	variational inference~\cite{}, gradient-based
-	optimization~\cite{}, and others.
+	Probabilistic programming \cite{GMR+08,MSP14,WVM14,GS15}
+	represents statistical models as programs written in an
+	otherwise general programming language that provides syntax
+	for the definition and conditioning of random variables.
+	Inference can be performed on probabilistic programs to
+	obtain the posterior distribution or point estimates of the
+	variables. Inference algorithms are provided by the
+	probabilistic programming system, and each algorithm is
+	usually applicable to a wide class of probabilistic programs
+	in a black box manner.  Inference techniques provided by the
+	system back end, such as Metropolis-Hastings
+	\cite{WSG11,MSP14,YHG14}, Hamiltonian Monte Carlo
+	\cite{SDT14}, expectation propagation \cite{MWG+10},
+	extensions of Sequential Monte Carlo
+	\cite{WVM14,MYM+15,PWD+14}, variational inference~\cite{},
+	gradient-based optimization~\cite{}, and others.
 
-	There are two polar views on the role of probabilistic programming.
-	One view is that probabilistic programming is a flexible tool for
-	specification of statistical models. Proponents of this view
+	There are two polar views on the role of probabilistic
+	programming.  One view is that probabilistic programming is
+	a flexible framework for specification and analysis of
+	statistical models.  Proponents of this view perceive
 	probabilistic programming as a tool for data science
-	practitioners. A typical work flow consists of data acquisition
-	and pre-processing, followed by several iterations of exploratory
-	model design and testing of inference algorithms. Once a
-	sufficiently robust statistical model is obtained, analysis
-	results are post-processed, visualized, and occasionally
-	integrated into algorithms of a production software system.
+	practitioners. A typical work flow consists of data
+	acquisition and pre-processing, followed by several
+	iterations of exploratory model design and testing of
+	inference algorithms. Once a sufficiently robust statistical
+	model is obtained, analysis results are post-processed,
+	visualized, and occasionally integrated into algorithms of a
+	production software system.
 
-	The other, emerging, view is that probabilistic programming is an
-	extension to a regular programming toolbox allowing
+	The other, emerging, view is that probabilistic programming
+	is an extension to a regular programming toolbox allowing
 	algorithms implemented in general programming languages to
-	have learnable parameters. In this view, statistical 
-	models are integrated into the software system, and inference
-	and optimization take place during data processing, prediction,
-	and algorithmic decision making in production.  Probabilistic
-	programming code runs unattended, and inference results are an
-	integral part of the algorithms.
+	have learnable parameters. In this view, statistical models
+	are integrated into the software system, and inference and
+	optimization take place during data processing, prediction,
+	and algorithmic decision making in production.
+	Probabilistic programming code runs unattended, and
+	inference results are an integral part of the algorithms.
 
 	In this work we are concerned with the later view on
-	probabilistic programming. Our objective is to pave a path to
-	deployment of probabilistic programs in production systems,
-	promoting a wider adoption of probabilistic programming as a
-	flexible tool for ubiquitous statistical inference.
+	probabilistic programming. Our objective is to pave a path
+	to deployment of probabilistic programs in production
+	systems, promoting a wider adoption of probabilistic
+	programming as a flexible tool for ubiquitous statistical
+	inference. Most probabilistic programming systems are
+	suited, to a certain extent, to be used in either the
+	exploratory or the production scenario. However, the
+	proliferation of probabilistic programming languages and
+	systems~\cite{} one one hand, and scarcity of success
+	stories about production use of probabilistic programming on
+	the other hand, suggest that there is a need for new ideas
+	and approaches to make probabilistic programming models
+	available for production use.
 
-The language war.
-
-A solution: probabilistic programming without a DSL.
-
-Organization of the paper.
+	Our attitude differs from the established approach in that
+	instead of proposing yet another probabilistic programming
+	language, either genuinely different~\cite{} or derived from
+	an existing general programming language by extending the
+	syntax or changing the semantics~\cite{}, we advocate
+	enabling probabilistic inference on models written in a
+	general probabilistic programming language, the same one as
+	the language used to program the bulk of the system. We
+	formulate guidelines for such implementation, and based on
+	the guidelines, introduce a probabilistic programming
+	facility for the Go programming language. By explaining our
+	design choices and implementation techniques, we argue that
+	a similar facility can be added to any modern general
+	purpose programming language, giving a production system
+	modeling and inference capabilities which do not fall short
+	from and sometimes exceed those of probabilistic programming
+	centric systems with custom languages and runtimes.
 
 	\subsection*{Contributions}
+
+	This work brings the following major contributions:
 
@@
 			programming facility for the Go programming language.
-	\item Design and implementation highlights 
+		\item Design and implementation highlights outlining
+			important choices we made in implementation of the
+			probabilistic programming facility, and providing a
+			basis for implementation of a similar facility for
+			other general programming languages.
 	\end{itemize}
@@
 A probabilistic model in Infergo is an implementation of an
-interface requiring a single method \texttt{Observe} which
+interface requiring a single method \lstinline{Observe} which
 accepts a vector (a Go \textit{slice}) of floats, the parameters
@@
 \section{Model Syntax and Rules of Differentiation}
+
+A model is defined in it's own package. The model must
+implement interface \lstinline{Model} containing a single method
+\lstinline{Observe}. \lstinline{Observe} accepts receives a slice of
+\lstinline{float64} --- the model parameters --- and returns a
+\lstinline{float64} scalar. For probabilistic inference, the
+returned value is interpreted as the log-likelihood of the model
+parameters. In the model's source code:
+
+\begin{enumerate}
+	\item Methods on the type implementing \lstinline{Model} returning a
+		single \lstinline{float64} or nothing are differentiated.
+	\item Within the methods, the following is differentiated:
+		\begin{itemize}
+			\item assignments to \lstinline{float64} (including parallel
+				assignments if all values are of type
+				\lstinline{float64});
+			\item returns of \lstinline{float64};
+			\item standalone calls to methods on the type implementing
+				\lstinline{Model} (apparently called for side  effects on
+				the model).
+		\end{itemize}
+\end{enumerate}
+
+Derivatives do not propagate through a function that is not
+an elemental or a call to a model method. If a derivative is
+not registered for an elemental, calling the elemental in a
+differentiated context will cause a run-time error.
+
+\subsection{Elementals}
+
+Functions are considered elementals (and must have a
+registered derivative) if their signature is of kind
+\lstinline{func (float64, float64*) float64}; that is, 
+one or more non-variadic \lstinline{float64} arguments and
+\lstinline{float64} return value. For example, function
+\lstinline{func (float64, float64, float64) float64}
+is considered elemental, while functions
+
+\begin{itemize}
+	\item \lstinline{func (...float64) float64}
+	\item \lstinline{func ([]float64) float64}
+	\item \lstinline{func (int, float64) float64}
+\end{itemize}
+
+are not. Gradients for selected functions from the \lstinline{math}
+package are pre-defined (\lstinline{Sqrt}, \lstinline{Exp}, \lstinline{Log}, \lstinline{Pow}, \lstinline{Sin},
+\lstinline{Cos}, \lstinline{Tan}). Auxiliary elemental functions with pre-defined
+gradients are provided in a separate package.
 

### 2019-04-16 paper/paper.tex
making things look like real

@@
     frame=none,
-    basicstyle=\footnotesize\ttfamily,
+    basicstyle=\ttfamily,
     numbers=left,

### 2019-04-16 paper/paper.tex
better abstract

@@
 \usepackage{hyperref}
-\usepackage{courier}
 \usepackage{listings}[language=Golang]
+\usepackage{GoMono}
 
@@
     frame=none,
-    basicstyle=\ttfamily,
+    basicstyle=\ttfamily\footnotesize,
     numbers=left,
@@
 \begin{abstract}
-	We share our experiences from integrating probabilistic
-	programming into a server-side software system and introduce
-	a probabilistic programming facility for Go, a modern
+	We propose design guidelines for a probabilistic programming
+	facility suitable for deployment as a part of a production
+	software system. As a reference implementation, we introduce
+	Infergo, a probabilistic programming facility for Go, a modern
 	programming language of choice for server-side software
-	development. We demonstrate how a lightweight but powerful
-	and efficient probabilistic programming facility can be
-	added to an existing programming language.  Server-side
-	application of probabilistic programming poses challenges
-	for a probabilistic programming system. We discuss the
-	challenges and our experience in overcoming them, and
-	suggest guidelines that can help in a wider adoption of
-	probabilistic programming in server-side software systems.
+	development. We argue that a similar probabilistic
+	programming facility can be added to most modern general
+	programming languages.
+
+	Probabilistic programming enables automatic tuning of
+	program parameters and algorithmic decision making through
+	probabilistic inference based on the data. To facilitate
+	addition of probabilistic programming capabilities to other
+	general programming languages, we share design choices and
+	implementation techniques employed in development of
+	Infergo. We illustrate applicability of Infergo to various
+	use cases on case studies, and evaluate Infergo's
+	performance on several benchmarks, comparing Infergo to
+	dedicated inference-centric probabilistic programming
+	systems.
 \end{abstract}
@@
 	usually applicable to a wide class of probabilistic programs
-	in a black box manner.  Inference techniques provided by the
-	system back end, such as Metropolis-Hastings
-	\cite{WSG11,MSP14,YHG14}, Hamiltonian Monte Carlo
-	\cite{SDT14}, expectation propagation \cite{MWG+10},
-	extensions of Sequential Monte Carlo
-	\cite{WVM14,MYM+15,PWD+14}, variational inference~\cite{},
-	gradient-based optimization~\cite{}, and others.
+	in a black-box manner. The algorithms include
+	Metropolis-Hastings\cite{WSG11,MSP14,YHG14}, Hamiltonian
+	Monte Carlo~\cite{Stan17}, expectation
+	propagation~\cite{MWG+10}, extensions of Sequential Monte
+	Carlo~\cite{WVM14,MYM+15,PWD+14}, variational
+	inference~\cite{WW13,KTR+17}, gradient-based
+	optimization~\cite{Stan17,BCJ+19}, and others.
 
@@
 	proliferation of probabilistic programming languages and
-	systems~\cite{} one one hand, and scarcity of success
+	systems\footnote{There are 45 probabilistic programming
+	languages listed on
+	Wikipedia~\cite{wiki:Probabilistic_programming_language}. 18
+	probabilistic programming languages were presented during
+	the developer meetup at PROBPROG'2018, the inaugural
+	conference on probabilistic
+	programming,~\url{http://probprog.cc/}, half of which are
+	not on the Wikipedia list.} on one hand, and scarcity of success
 	stories about production use of probabilistic programming on
@@
 	instead of proposing yet another probabilistic programming
-	language, either genuinely different~\cite{} or derived from
+	language, either genuinely different~\cite{MMR+07,Stan17} or derived from
 	an existing general programming language by extending the
-	syntax or changing the semantics~\cite{}, we advocate
+	syntax or changing the semantics~\cite{GMR+08,TMY+16,GXG18}, we advocate
 	enabling probabilistic inference on models written in a
@@
 provide an interface in a popular programming language, e.g.
-Python (Edward~\cite{THS+17}, Pyro~\cite{Pyro18}) are easier
+Python (Edward~\cite{THS+17}, Pyro~\cite{BCJ+19}) are easier
 to integrate and deploy, however the smaller the footprint
@@
 
-TODO: Example
+Model:
+
+\begin{lstlisting}
+type Model struct {
+    Data []float64
+}
+
+// x[0] is the mean,
+// x[1] is the log stddev of the distribution
+func (m *Model) Observe(x []float64) float64 {
+    // Our prior is a unit normal ...
+    ll := Normal.Logps(0, 1, x...)
+    // ... but the posterior is based 
+	// on data observations.
+	ll += Normal.Logps(x[0], math.Exp(x[1]),
+	                   m.Data...)
+    return ll
+}
+\end{lstlisting}
+
+Inference:
+
+\begin{lstlisting}
+// Data
+m := &Model{[]float64{
+	-0.854, 1.067, -1.220, 0.818, -0.749,
+	0.805, 1.443, 1.069, 1.426, 0.308}}
+
+// Parameters
+mean, logs := 0, 0
+x := []float64{mean, logs}
+\end{lstlisting}
+
+Optimization:
+
+\begin{lstlisting}
+opt := &infer.Momentum{
+    Rate:  0.01,
+    Decay: 0.998,
+}
+for iter := 0; iter != 1000; iter++ {
+    opt.Step(m, x)
+}
+mean, logs = x[0], x[1]
+\end{lstlisting}
+
+Posterior:
+
+\begin{lstlisting}
+hmc := &infer.HMC{
+	Eps: 0.1,
+}
+samples := make(chan []float64)
+hmc.Sample(m, x, samples)
+for i := 0; i != 1000; i++ {
+	x = <-samples
+}
+hmc.Stop()
+\end{lstlisting}
 
@@
 
-A model is defined in it's own package. The model must
+In infergo, the inference is performed on a model. An Infergo
+model is a data type and methods defined on the data type.
+Inference algorithms accept a model object as an argument.
+A model object usually encapsulates observations --- the data 
+to which the parameter values are adapted.
+
+A model is implemented in Go, virtually
+unrestricted\footnote{There are insignificant implementation
+limitations which are gradually lifted as Infergo is being
+developed.}. Because of this, the rules of model specification
+and differentiation are quite straightforward. Nonetheless, the
+differentiation rules ensure that provided that the model's
+\lstinline{Observe} method (including calls to other model methods)
+is differentiable, the gradient is properly computed~\cite{GW08}.
+
+A model is defined in its own package. The model must
 implement interface \lstinline{Model} containing a single method
@@
 returned value is interpreted as the log-likelihood of the model
-parameters. In the model's source code:
+parameters. Inference models rely on computing the gradient of
+the returned value with respect to the model parameters through
+automatic differentiation. In the model's source code:
 
@@
 
-Derivatives do not propagate through a function that is not
-an elemental or a call to a model method. If a derivative is
-not registered for an elemental, calling the elemental in a
-differentiated context will cause a run-time error.
-
-\subsection{Elementals}
+Functions for which the gradient is provided rather than derived
+through automatic differentiation are called
+\textit{elementals}~\cite{GW08}.  Derivatives do not propagate
+through a function that is not an elemental or a call to a model
+method. If a derivative is not registered for an elemental,
+calling the elemental in a differentiated context will cause a
+run-time error.
 
@@
 is considered elemental, while functions
-
 \begin{itemize}
@@
 \end{itemize}
+are not. Gradients for selected functions from the
+\lstinline{math} package are pre-defined (\lstinline{Sqrt},
+\lstinline{Exp}, \lstinline{Log}, \lstinline{Pow},
+\lstinline{Sin}, \lstinline{Cos}, \lstinline{Tan}). Auxiliary
+elemental functions with pre-defined gradients are provided in a
+separate package.
 
-are not. Gradients for selected functions from the \lstinline{math}
-package are pre-defined (\lstinline{Sqrt}, \lstinline{Exp}, \lstinline{Log}, \lstinline{Pow}, \lstinline{Sin},
-\lstinline{Cos}, \lstinline{Tan}). Auxiliary elemental functions with pre-defined
-gradients are provided in a separate package.
+Any user-defined function of appropriate kind can be used as an
+elemental inside model methods. An elemental must be declared
+for the function using a call to
+\begin{lstlisting}
+func RegisterElemental(
+	f interface{}, 
+	g func(value float64,
+		   params ...float64) []float64).
+\end{lstlisting} For
+example, the call
+\begin{lstlisting}
+RegisterElemental(Sigm,
+	func(value float64,
+	     _ ...float64) []float64 {
+		return []float64{value*(1-value)}
+	})
+\end{lstlisting}
+registers the gradient for function \lstinline{Sigm}.
 
@@
 
-The author owes a debt of gratitude to Frank Wood who introduced
-the author to probabilistic programming and inspired to
-pursue probabilistic programming paradigms and applications. The
-author also would like to thank Jan-Willem van de Meent, with
-whom the author had fruitful discussions of motives, ideas, and
-implementation choices behind Infergo, and whose thoughts and
-recommendations significantly influenced Infergo design.
-Finally, the author thanks PUB+, the company the author works
-for, for supporting development of Infergo and letting the
-author experiment with applying probabilistic programming to
-critical decision-making in production environment.
+% The author owes a debt of gratitude to Frank Wood who introduced
+% the author to probabilistic programming and inspired to
+% pursue probabilistic programming paradigms and applications. The
+% author also would like to thank Jan-Willem van de Meent, with
+% whom the author had fruitful discussions of motives, ideas, and
+% implementation choices behind Infergo, and whose thoughts and
+% recommendations significantly influenced Infergo design.
+% Finally, the author thanks PUB+, the company the author works
+% for, for supporting development of Infergo and letting the
+% author experiment with applying probabilistic programming to
+% critical decision-making in production environment.
 

### 2019-04-16 paper/paper.tex
in progress

@@
 	probabilistic programming as a tool for data science
-	practitioners. A typical work flow consists of data
+	practitioners. A typical workflow consists of data
 	acquisition and pre-processing, followed by several
@@
 	The other, emerging, view is that probabilistic programming
-	is an extension to a regular programming toolbox allowing
+	is an extension of a regular programming toolbox allowing
 	algorithms implemented in general programming languages to
@@
 			probabilistic programming facility deployable as a part 
-			of production software systems.
+			of a production software system.
 		\item A reference implementation of deployable probabilistic
@@
 
-\subsection{Probabilistic ``Hello, World!''}
+\subsection{A Basic Example}
 
@@
 implement interface \lstinline{Model} containing a single method
-\lstinline{Observe}. \lstinline{Observe} accepts receives a slice of
+\lstinline{Observe}. \lstinline{Observe} accepts a slice of
 \lstinline{float64} --- the model parameters --- and returns a

### 2019-04-16 paper/paper.tex
more citations

@@
 	a flexible framework for specification and analysis of
-	statistical models.  Proponents of this view perceive
-	probabilistic programming as a tool for data science
-	practitioners. A typical workflow consists of data
-	acquisition and pre-processing, followed by several
+	statistical models~\cite{WSG11,TMY+16,GXG18}. Proponents of
+	this view perceive probabilistic programming as a tool for
+	data science practitioners. A typical workflow consists of
+	data acquisition and pre-processing, followed by several
 	iterations of exploratory model design and testing of
@@
 	the guidelines, introduce a probabilistic programming
-	facility for the Go programming language. By explaining our
-	design choices and implementation techniques, we argue that
-	a similar facility can be added to any modern general
-	purpose programming language, giving a production system
-	modeling and inference capabilities which do not fall short
-	from and sometimes exceed those of probabilistic programming
-	centric systems with custom languages and runtimes.
+	facility for the Go programming language~\cite{Golang}. By
+	explaining our design choices and implementation techniques,
+	we argue that a similar facility can be added to any modern
+	general purpose programming language, giving a production
+	system modeling and inference capabilities which do not fall
+	short from and sometimes exceed those of probabilistic
+	programming centric systems with custom languages and
+	runtimes.
 

### 2019-04-16 paper/paper.tex
removed listings-golang, included inline in the paper

@@
 
+Performing inference as an integral part of algorithms of a
+production software system poses challenges to the probabilistic
+programming facility. To address the challenges, we proposed
+guidelines for implementation of a deployable probabilistic
+programming facility. Infergo, a probabilistic programming
+facility in Go, is a reference implementation following 
+the guidelines.
+
 A lightweight probabilistic programming facility similar to

### 2019-04-16 paper/paper.tex
conclusion

--- 2019-04-16 paper/paper.tex
@@
 
-A lightweight probabilistic programming facility similar to
-Infergo can be added to most modern general-purpose programming
-languages, in particular those used in implementing large-scale
-software systems, making probabilistic programming inference
-more accessible in server-side applications.
+Building a production software system which employs
+probabilistic inference as an integral part of prediction and
+decision making algorithms is challenging. To address the
+challenges, we proposed guidelines for implementation of a
+deployable probabilistic programming facility. Further on, we
+introduced Infergo, a probabilistic programming facility in Go,
+as a reference implementation following the guidelines. Infergo
+allows to program inference models in virtually unrestricted Go,
+and to perform inference on the model using either supplied or
+third-party algorithms. We demonstrated on case studies how
+Infergo overcomes the challenges by following the guidelines.
+
+A probabilistic programming facility similar to Infergo can be
+added to most modern general programming languages, in
+particular, those used in implementing large-scale software
+systems, making probabilistic programming inference more
+accessible in server-side applications. We described Infergo
+design choices and implementation techniques in intent to
+encourage broader adoption of probabilistic programming in
+production software systems and ease implementation of similar
+facilities in other languages and software environments.
+
+Infergo is an ongoing project. Development of Infergo is
+influenced both by feedback from the open-source Go community
+and by the needs of the software system in which Infergo was
+initially deployed, and which still supplies the most demanding
+use cases in terms of flexibility, scalability, and ease of
+integration and maintenance. Deploying probabilistic programming
+as a part of production software system has a mutual benefit of
+both improving the core algorithms of the system and of helping
+to shape and improve probabilistic modeling and inference, thus
+further advancing the field of probabilistic programming.
 

### 2019-04-16 paper/paper.tex
in progress

@@
     basicstyle=\ttfamily\footnotesize,
-    numbers=left,
+    numbers=none,
     numbersep=5pt,
@@
 
-Model:
-
+Let us illustrate the use of Infergo on a basic example serving
+as a probabilistic programming equivalent of the "Hello, World!"
+program --- inferring parameters of a unidimensional Normal
+distribution. In this example, the model holds the set of
+samples from which the distribution parameters are to be
+inferred:
 \begin{lstlisting}
@@
 }
+\end{lstlisting}
 

### 2019-04-16 paper/paper.tex
removed acknowledgements altogether

--- 2019-04-16 paper/paper.tex
@@
 
-Performing inference as an integral part of algorithms of a
-production software system poses challenges to the probabilistic
-programming facility. To address the challenges, we proposed
-guidelines for implementation of a deployable probabilistic
-programming facility. Infergo, a probabilistic programming
-facility in Go, is a reference implementation following 
-the guidelines.
-
 A lightweight probabilistic programming facility similar to
@@
 
-\section*{Acknowledgements}
-
-% The author owes a debt of gratitude to Frank Wood who introduced
-% the author to probabilistic programming and inspired to
-% pursue probabilistic programming paradigms and applications. The
-% author also would like to thank Jan-Willem van de Meent, with
-% whom the author had fruitful discussions of motives, ideas, and
-% implementation choices behind Infergo, and whose thoughts and
-% recommendations significantly influenced Infergo design.
-% Finally, the author thanks PUB+, the company the author works
-% for, for supporting development of Infergo and letting the
-% author experiment with applying probabilistic programming to
-% critical decision-making in production environment.
-
 \end{sloppypar}

### 2019-04-16 paper/paper.tex
hello world in progress

--- 2019-04-16 paper/paper.tex
@@
 program --- inferring parameters of a unidimensional Normal
-distribution. In this example, the model holds the set of
-samples from which the distribution parameters are to be
+distribution. In this example, the model object holds the set of
+observations from which the distribution parameters are to be
 inferred:
@@
 
+The \lstinline{Observe} method computes the log-likelihood of
+the distribution parameters. The Normal distribution has two
+parameters: mean $\mu$ and standard deviation
+$\sigma$. To ease the inference, positive $\sigma$ is transformed to
+unrestricted $\log \sigma$, and the parameter vector is $[\mu,
+\log \sigma]$. \lstinline{Observe} first imposes a prior on the
+parameters, and then conditions the posterior distribution on
+the observations:
+
+\begin{lstlisting}
 // x[0] is the mean,

### 2019-04-17 paper/paper.tex
hello world example, more references

@@
 \lstset{ % add your own preferences
-    frame=none,
+    frame=single,
     basicstyle=\ttfamily\footnotesize,
@@
 $\sigma$. To ease the inference, positive $\sigma$ is transformed to
-unrestricted $\log \sigma$, and the parameter vector is $[\mu,
+unrestricted $\log \sigma$, and the parameter vector \lstinline{x} is $[\mu,
 \log \sigma]$. \lstinline{Observe} first imposes a prior on the
@@
 
-Inference:
+For inference, we supply the data and optionally initialize the
+parameter vector. Here the data is hard-coded, but in general
+can be read from a file, a database, a network connection,
+or dynamically generated by another part of the program:
 
@@
 // Parameters
-mean, logs := 0, 0
-x := []float64{mean, logs}
+x := []float64{}
 \end{lstlisting}
 
-Optimization:
+The fastest and most straightforward inference is the maximum
+\textit{a posteriori} (MAP) estimation, which can be done using
+a gradient descent algorithm (Adam~\cite{KB15} in this case).
 
 \begin{lstlisting}
-opt := &infer.Momentum{
+opt := &infer.Adam{
     Rate:  0.01,
-    Decay: 0.998,
 }
@@
 }
-mean, logs = x[0], x[1]
 \end{lstlisting}
 
-Posterior:
+To recover the full posterior distribution of
+the parameters we use Hamiltonian Monte Carlo~\cite{N12}:
 
@@
 \end{lstlisting}
+
+Models are of course not restricted to linear code, and may
+comprise multiple methods, contain loops, conditional
+statements, function and method calls, and recursion. Case
+studies~(Section~\ref{sec:case-studies}) provide elaborated
+model examples.
 
@@
 \section{Case Studies}
+\label{sec:case-studies}
 

### 2019-04-17 paper/paper.tex
small edits, more references

@@
 Models are of course not restricted to linear code, and may
-comprise multiple methods, contain loops, conditional
+comprise multiple methods containing loops, conditional
 statements, function and method calls, and recursion. Case
@@
 
-In infergo, the inference is performed on a model. An Infergo
+In Infergo, the inference is performed on a model. An Infergo
 model is a data type and methods defined on the data type.
@@
 introduced Infergo, a probabilistic programming facility in Go,
-as a reference implementation following the guidelines. Infergo
-allows to program inference models in virtually unrestricted Go,
-and to perform inference on the model using either supplied or
-third-party algorithms. We demonstrated on case studies how
-Infergo overcomes the challenges by following the guidelines.
+as a reference implementation in accordance with the guidelines.
+Infergo allows to program inference models in virtually
+unrestricted Go, and to perform inference on the model using
+either supplied or third-party algorithms. We demonstrated on
+case studies how Infergo overcomes the challenges by following
+the guidelines.
 
@@
 added to most modern general programming languages, in
-particular, those used in implementing large-scale software
+particular, those used for implementation of large-scale software
 systems, making probabilistic programming inference more

### 2019-04-17 paper/paper.tex
more references

@@
 
+\lstdefinelanguage{Stan}{
+  morekeywords=[1]{functions,data,parameters,transformed,model,generated,quantities,%
+    for,in,while,print,if,else,lower,upper,increment_log_prob,T,return,%
+    reject,integrate_ode,integrate_ode_bdf,integrate_ode_rk45,target},%
+  morekeywords=[2]{int,real,vector,%
+    ordered,positive_ordered,simplex,unit_vector,%
+    row_vector,matrix,%
+    cholesky_factor_corr,cholesky_factor_cov,%
+    coor_matrix,cov_matrix,%
+    void},%
+  morekeywords=[3]{%
+    Phi,%
+    Phi_approx,%
+    abs,%
+    acos,%
+    acosh,%
+    append_col,%
+    append_row,%
+    asin,%
+    asinh,%
+    atan,%
+    atan2,%
+    atanh,%
+    bernoulli_cdf,%
+    bernoulli_cdf_log,%
+    bernoulli_lccdf,%
+    bernoulli_lcdf,%
+    bernoulli_logit_lpmf,%
+    bernoulli_logit_lpmf,%
+    bernoulli_lpmf,%
+    bernoulli_lpmf,%
+    bernoulli_rng,%
+    bessel_first_kind,%
+    bessel_second_kind,%
+    beta_binomial_cdf,%
+    beta_binomial_cdf_log,%
+    beta_binomial_lccdf,%
+    beta_binomial_lcdf,%
+    beta_binomial_lpmf,%
+    beta_binomial_lpmf,%
+    beta_binomial_rng,%
+    beta_cdf,%
+    beta_cdf_log,%
+    beta_lccdf,%
+    beta_lcdf,%
+    beta_lpdf,%
+    beta_lpdf,%
+    beta_rng,%
+    binary_log_loss,%
+    binomial_cdf,%
+    binomial_cdf_log,%
+    binomial_coefficient_log,%
+    binomial_lccdf,%
+    binomial_lcdf,%
+    binomial_logit_lpmf,%
+    binomial_logit_lpmf,%
+    binomial_lpmf,%
+    binomial_lpmf,%
+    binomial_rng,%
+    block,%
+    categorical_logit_lpmf,%
+    categorical_logit_lpmf,%
+    categorical_lpmf,%
+    categorical_lpmf,%
+    categorical_rng,%
+    cauchy_cdf,%
+    cauchy_cdf_log,%
+    cauchy_lccdf,%
+    cauchy_lcdf,%
+    cauchy_lpdf,%
+    cauchy_lpdf,%
+    cauchy_rng,%
+    cbrt,%
+    ceil,%
+    chi_square_cdf,%
+    chi_square_cdf_log,%
+    chi_square_lccdf,%
+    chi_square_lcdf,%
+    chi_square_lpdf,%
+    chi_square_lpdf,%
+    chi_square_rng,%
+    cholesky_decompose,%
+    col,%
+    cols,%
+    columns_dot_product,%
+    columns_dot_self,%
+    cos,%
+    cosh,%
+    crossprod,%
+    csr_extract_u,%
+    csr_extract_v,%
+    csr_extract_w,%
+    csr_matrix_times_vector,%
+    csr_to_dense_matrix,%
+    cumulative_sum,%
+    determinant,%
+    diag_matrix,%
+    diag_post_multiply,%
+    diag_pre_multiply,%
+    diagonal,%
+    digamma,%
+    dims,%
+    dirichlet_lpdf,%
+    dirichlet_lpdf,%
+    dirichlet_rng,%
+    distance,%
+    dot_product,%
+    dot_self,%
+    double_exponential_cdf,%
+    double_exponential_cdf_log,%
+    double_exponential_lccdf,%
+    double_exponential_lcdf,%
+    double_exponential_lpdf,%
+    double_exponential_lpdf,%
+    double_exponential_rng,%
+    e,%
+    eigenvalues_sym,%
+    eigenvectors_sym,%
+    erf,%
+    erfc,%
+    exp,%
+    exp2,%
+    exp_mod_normal_cdf,%
+    exp_mod_normal_cdf_log,%
+    exp_mod_normal_lccdf,%
+    exp_mod_normal_lcdf,%
+    exp_mod_normal_lpdf,%
+    exp_mod_normal_lpdf,%
+    exp_mod_normal_rng,%
+    expm1,%
+    exponential_cdf,%
+    exponential_cdf_log,%
+    exponential_lccdf,%
+    exponential_lcdf,%
+    exponential_lpdf,%
+    exponential_lpdf,%
+    exponential_rng,%
+    fabs,%
+    falling_factorial,%
+    fdim,%
+    floor,%
+    fma,%
+    fmax,%
+    fmin,%
+    fmod,%
+    frechet_cdf,%
+    frechet_cdf_log,%
+    frechet_lccdf,%
+    frechet_lcdf,%
+    frechet_lpdf,%
+    frechet_lpdf,%
+    frechet_rng,%
+    gamma_cdf,%
+    gamma_cdf_log,%
+    gamma_lccdf,%
+    gamma_lcdf,%
+    gamma_lpdf,%
+    gamma_lpdf,%
+    gamma_p,%
+    gamma_q,%
+    gamma_rng,%
+    gaussian_dlm_obs_lpdf,%
+    gaussian_dlm_obs_lpdf,%
+    get_lp,%
+    gumbel_cdf,%
+    gumbel_cdf_log,%
+    gumbel_lccdf,%
+    gumbel_lcdf,%
+    gumbel_lpdf,%
+    gumbel_lpdf,%
+    gumbel_rng,%
+    head,%
+    hypergeometric_lpmf,%
+    hypergeometric_lpmf,%
+    hypergeometric_rng,%
+    hypot,%
+    if_else,%
+    inc_beta,%
+    int_step,%
+    inv,%
+    inv_chi_square_cdf,%
+    inv_chi_square_cdf_log,%
+    inv_chi_square_lccdf,%
+    inv_chi_square_lcdf,%
+    inv_chi_square_lpdf,%
+    inv_chi_square_lpdf,%
+    inv_chi_square_rng,%
+    inv_cloglog,%
+    inv_gamma_cdf,%
+    inv_gamma_cdf_log,%
+    inv_gamma_lccdf,%
+    inv_gamma_lcdf,%
+    inv_gamma_lpdf,%
+    inv_gamma_lpdf,%
+    inv_gamma_rng,%
+    inv_logit,%
+    inv_phi,%
+    inv_sqrt,%
+    inv_square,%
+    inv_wishart_lpdf,%
+    inv_wishart_lpdf,%
+    inv_wishart_rng,%
+    inverse,%
+    inverse_spd,%
+    is_inf,%
+    is_nan,%
+    lbeta,%
+    lchoose,%
+    lgamma,%
+    lkj_corr_cholesky_lpdf,%
+    lkj_corr_cholesky_lpdf,%
+    lkj_corr_cholesky_rng,%
+    lkj_corr_lpdf,%
+    lkj_corr_lpdf,%
+    lkj_corr_rng,%
+    lmgamma,%
+    lmultiply,%
+    log,%
+    log10,%
+    log1m,%
+    log1m_exp,%
+    log1m_inv_logit,%
+    log1p,%
+    log1p_exp,%
+    log2,%
+    log_determinant,%
+    log_diff_exp,%
+    log_falling_factorial,%
+    log_inv_logit,%
+    log_mix,%
+    log_rising_factorial,%
+    log_softmax,%
+    log_sum_exp,%
+    logistic_cdf,%
+    logistic_cdf_log,%
+    logistic_lccdf,%
+    logistic_lcdf,%
+    logistic_lpdf,%
+    logistic_lpdf,%
+    logistic_rng,%
+    logit,%
+    lognormal_cdf,%
+    lognormal_cdf_log,%
+    lognormal_lccdf,%
+    lognormal_lcdf,%
+    lognormal_lpdf,%
+    lognormal_lpdf,%
+    lognormal_rng,%
+    machine_precision,%
+    max,%
+    mdivide_left_tri_low,%
+    mdivide_right_tri_low,%
+    mean,%
+    min,%
+    modified_bessel_first_kind,%
+    modified_bessel_second_kind,%
+    multi_gp_cholesky_lpdf,%
+    multi_gp_cholesky_lpdf,%
+    multi_gp_lpdf,%
+    multi_gp_lpdf,%
+    multi_normal_cholesky_lpdf,%
+    multi_normal_cholesky_lpdf,%
+    multi_normal_cholesky_rng,%
+    multi_normal_lpdf,%
+    multi_normal_lpdf,%
+    multi_normal_prec_lpdf,%
+    multi_normal_prec_lpdf,%
+    multi_normal_rng,%
+    multi_student_t_lpdf,%
+    multi_student_t_lpdf,%
+    multi_student_t_rng,%
+    multinomial_lpmf,%
+    multinomial_lpmf,%
+    multinomial_rng,%
+    multiply_log,%
+    multiply_lower_tri_self_transpose,%
+    neg_binomial_2_cdf,%
+    neg_binomial_2_cdf_log,%
+    neg_binomial_2_lccdf,%
+    neg_binomial_2_lcdf,%
+    neg_binomial_2_log_lpmf,%
+    neg_binomial_2_log_lpmf,%
+    neg_binomial_2_log_rng,%
+    neg_binomial_2_lpmf,%
+    neg_binomial_2_lpmf,%
+    neg_binomial_2_rng,%
+    neg_binomial_cdf,%
+    neg_binomial_cdf_log,%
+    neg_binomial_lccdf,%
+    neg_binomial_lcdf,%
+    neg_binomial_lpmf,%
+    neg_binomial_lpmf,%
+    neg_binomial_rng,%
+    negative_infinity,%
+    normal_cdf,%
+    normal_cdf_log,%
+    normal_lccdf,%
+    normal_lcdf,%
+    normal_lpdf,%
+    normal_lpdf,%
+    normal_rng,%
+    not_a_number,%
+    num_elements,%
+    ordered_logistic_lpmf,%
+    ordered_logistic_lpmf,%
+    ordered_logistic_rng,%
+    owens_t,%
+    pareto_cdf,%
+    pareto_cdf_log,%
+    pareto_lccdf,%
+    pareto_lcdf,%
+    pareto_lpdf,%
+    pareto_lpdf,%
+    pareto_rng,%
+    pareto_type_2_cdf,%
+    pareto_type_2_cdf_log,%
+    pareto_type_2_lccdf,%
+    pareto_type_2_lcdf,%
+    pareto_type_2_lpdf,%
+    pareto_type_2_lpdf,%
+    pareto_type_2_rng,%
+    pi,%
+    poisson_cdf,%
+    poisson_cdf_log,%
+    poisson_lccdf,%
+    poisson_lcdf,%
+    poisson_log_lpmf,%
+    poisson_log_lpmf,%
+    poisson_log_rng,%
+    poisson_lpmf,%
+    poisson_lpmf,%
+    poisson_rng,%
+    positive_infinity,%
+    pow,%
+    prod,%
+    qr_Q,%
+    qr_R,%
+    quad_form,%
+    quad_form_diag,%
+    quad_form_sym,%
+    rank,%
+    rayleigh_cdf,%
+    rayleigh_cdf_log,%
+    rayleigh_lccdf,%
+    rayleigh_lcdf,%
+    rayleigh_lpdf,%
+    rayleigh_lpdf,%
+    rayleigh_rng,%
+    rep_array,%
+    rep_matrix,%
+    rep_row_vector,%
+    rep_vector,%
+    rising_factorial,%
+    round,%
+    row,%
+    rows,%
+    rows_dot_product,%
+    rows_dot_self,%
+    scaled_inv_chi_square_cdf,%
+    scaled_inv_chi_square_cdf_log,%
+    scaled_inv_chi_square_lccdf,%
+    scaled_inv_chi_square_lcdf,%
+    scaled_inv_chi_square_lpdf,%
+    scaled_inv_chi_square_lpdf,%
+    scaled_inv_chi_square_rng,%
+    sd,%
+    segment,%
+    sin,%
+    singular_values,%
+    sinh,%
+    size,%
+    skew_normal_cdf,%
+    skew_normal_cdf_log,%
+    skew_normal_lccdf,%
+    skew_normal_lcdf,%
+    skew_normal_lpdf,%
+    skew_normal_lpdf,%
+    skew_normal_rng,%
+    softmax,%
+    sort_asc,%
+    sort_desc,%
+    sort_indices_asc,%
+    sort_indices_desc,%
+    sqrt,%
+    sqrt2,%
+    square,%
+    squared_distance,%
+    step,%
+    student_t_cdf,%
+    student_t_cdf_log,%
+    student_t_lccdf,%
+    student_t_lcdf,%
+    student_t_lpdf,%
+    student_t_lpdf,%
+    student_t_rng,%
+    sub_col,%
+    sub_row,%
+    sum,%
+    tail,%
+    tan,%
+    tanh,%
+    tcrossprod,%
+    tgamma,%
+    to_array_1d,%
+    to_array_2d,%
+    to_matrix,%
+    to_row_vector,%
+    to_vector,%
+    trace,%
+    trace_gen_quad_form,%
+    trace_quad_form,%
+    trigamma,%
+    trunc,%
+    uniform_cdf,%
+    uniform_cdf_log,%
+    uniform_lccdf,%
+    uniform_lcdf,%
+    uniform_lpdf,%
+    uniform_lpdf,%
+    uniform_rng,%
+    variance,%
+    von_mises_lpdf,%
+    von_mises_lpdf,%
+    von_mises_rng,%
+    weibull_cdf,%
+    weibull_cdf_log,%
+    weibull_lccdf,%
+    weibull_lcdf,%
+    weibull_lpdf,%
+    weibull_lpdf,%
+    weibull_rng,%
+    wiener_lpdf,%
+    wiener_lpdf,%
+    wishart_lpdf,%
+    wishart_lpdf,%
+    wishart_rng
+  },%
+  otherkeywords={<-,~,+=,=},%
+  sensitive=true,%
+  morecomment=[l]{\#},%
+  morecomment=[l]{//},%
+  morecomment=[n]{/*}{*/},%
+  string=[d]"%,
+  literate={<-}{{$\leftarrow$}}1 {~}{{$\sim$}}1%
+}
+
 \lstset{ % add your own preferences
@@
     numbers=none,
-    numbersep=5pt,
+    numbersep=4pt,
+	framexleftmargin=10pt,
     showstringspaces=false,
@@
 \section{Related Work}
+
+Integrated probabilistic programming (Figaro, Anglican, Turing).
+
+Differentiable programming for machine learning (fundamentals,
+Stan, Tangent, Flux, Zygote, Pyro, Edward).
 
@@
 
+\subsection{8 schools}
+
+\begin{figure*}
+	\begin{minipage}{0.45\textwidth}
+\begin{lstlisting}[language=Stan,numbers=left]
+data {
+  int<lower=0> J;
+  vector[J] y;
+  vector<lower=0>[J] sigma;
+}
+
+parameters {
+  real mu;
+  real<lower=0> tau;
+  vector[J] eta;
+}
+
+transformed parameters {
+  vector[J] theta;
+  theta = mu + tau * eta;
+}
+
+model {
+  eta ~ normal(0, 1);
+  y ~ normal(theta, sigma);
+}
+\end{lstlisting}
+
+\centering
+a. Stan
+	\end{minipage}
+	\hfill
+	\begin{minipage}{0.45\textwidth}
+\begin{lstlisting}[numbers=left]
+type Model struct {
+  J          int
+  Y          []float64
+  Sigma      []float64
+}
+
+func (m *Model) Observe(x []float64) float64 {
+  mu := x[0]
+  tau := math.Exp(x[1])
+  eta := x[2:]
+
+  ll := Normal.Logp(0, 1, eta)
+
+  for i, y := range m.Y {
+    theta := mu + tau*eta[i]
+    ll += Normal.Logp(theta, m.Sigma[i], y)
+  }
+
+  return ll
+}
+\end{lstlisting}
+\centering
+b. Infergo
+	\end{minipage}
+	\caption{8 schools: Stan vs. Infergo. The Go implementation
+	has a similar length and structure to the Stan model.}
+	\label{fig:8-schools}
+\end{figure*}
+
+\subsection{Linear regression}
+
+\begin{figure}
+\begin{lstlisting}[numbers=left]
+type Model struct {
+  Data  [][]float64
+}
+
+func (m *Model) Observe(x []float64)
+    float64 {
+  ll := 0.
+
+  alpha, beta := x[0], x[1]
+  sigma := math.Exp(x[2])
+
+  for i := range m.Data {
+    ll += Normal.Logp(
+      m.Simulate(m.Data[i][0], alpha, beta),
+      sigma, m.Data[i][1])
+  }
+  return ll
+
+// Simulate predicts y for x based
+// on inferred parameters.
+func (m *Model) Simulate(x, alpha, beta float64)
+    float64 {
+  y := alpha +beta*x
+  return y
+}
+\end{lstlisting}
+\caption{Linear regression: the simulator is re-used in both
+	inference and prediction.}
+\label{fig:linear-regression}
+\end{figure}
+
+\subsection{Gaussian mixture}
+
+\begin{figure}
+\begin{lstlisting}[numbers=left]
+type Model struct {
+  Data  []float64 // samples
+  NComp int       // number of components
+}
+
+func (m *Model) Observe(x []float64) float64 {
+  ll := 0.0
+  mu := make([]float64, m.NComp)
+  sigma := make([]float64, m.NComp)
+
+  // Fetch component parameters
+  for j := 0; j != m.NComp; j++ {
+    mu[j] = x[2*j]
+    sigma[j] = math.Exp(x[2*j+1])
+  }
+
+  // Compute log likelihood of mixture
+  // given the data
+  for i := 0; i != len(m.Data); i++ {
+    var l float64
+    for j := 0; j != m.NComp; j++ {
+      lj := Normal.Logp(mu[j], sigma[j], m.Data[i])
+      if j == 0 {
+        l = lj
+      } else {
+        l = mathx.LogSumExp(l, lj)
+      }
+    }
+    ll += l
+  }
+  return ll
+}
+\end{lstlisting}
+\caption{Gaussian mixture: Infergo can handle models with
+	dynamic control structure.}
+\label{fig:linear-regression}
+\end{figure}
+
 \subsection{Performance}

### 2019-04-17 paper/paper.tex
more references, plan for related work
better rendering of listings

@@
     numbersep=4pt,
-	framexleftmargin=10pt,
     showstringspaces=false,
@@
 \section{Related Work}
-
-Integrated probabilistic programming (Figaro, Anglican, Turing).
+Probabilistic programming languages \cite{GS15}, \cite{MPY+18},
+\cite{Stan17}, \cite{GGS19}.
+
+Integrated probabilistic programming (Figaro~\cite{P09},
+Anglican~\cite{TMY+16}, Turing~\cite{GXG18}).
 
@@
 	\begin{minipage}{0.45\textwidth}
-\begin{lstlisting}[language=Stan,numbers=left]
+	\begin{lstlisting}[language=Stan,framexleftmargin=10pt,numbers=left]
 data {
@@
 	\begin{minipage}{0.45\textwidth}
-\begin{lstlisting}[numbers=left]
+\begin{lstlisting}[framexleftmargin=10pt,numbers=left]
 type Model struct {
@@
 \begin{figure}
-\begin{lstlisting}[numbers=left]
+\begin{lstlisting}[framexleftmargin=10pt,numbers=left]
 type Model struct {
@@
 	dynamic control structure.}
-\label{fig:linear-regression}
+\label{fig:gaussian-mixture}
 \end{figure}

### 2019-04-17 paper/paper.tex
in progress, broken

@@
 \section{Related Work}
-Probabilistic programming languages \cite{GS15}, \cite{MPY+18},
-\cite{Stan17}, \cite{GGS19}.
+
+Our work is related to three interconnected areas of research:
+\begin{enumerate}
+	\item Design and implementation of probabilistic programming
+		languages. 
+	\item Integration and interoperability between probabilistic
+		programming models and the surrounding software
+		systems.
+	\item Differentiable programming for machine learning.
+\end{enumerate}
+
+Probabilistic programming is traditionally associated with
+design and implemenetation of probabilistic programming
+languages. An apparent justification for a new programming
+language is that a probabilistic program has a different
+semantics~\cite{GHNR14} than a `regular' program of similar
+structure. Goodman and Stuhlm\"uller~\cite{GS15} offer a
+systematic approach to design and implementation of
+probabilistic programming languages based on extension of the
+syntax of a (subset of) regular programming language and
+transformational compilation to continuation-passing form. Van
+de Meent et al~\cite{MPY+18} give a comprehensive overview of
+available choices of design and implementation of first-order
+and higher-order probabilistic programming languages.
+Stan~\cite{Stan17} is built around an imperative probabilistic
+programming language with program structure tuned for most
+efficient execution of inference on the model.
+SlicStan~\cite{GGS19} is built on top of Stan as a
+source-to-source compiler, providing the model developer with
+a more intuitive language while retaining performance benefits
+of Stan.
 

### 2019-04-17 paper/paper.tex
in progress related work

@@
 Probabilistic programming is traditionally associated with
-design and implemenetation of probabilistic programming
+design and implementation of probabilistic programming
 languages. An apparent justification for a new programming
@@
 a more intuitive language while retaining performance benefits
-of Stan.
-
-Integrated probabilistic programming (Figaro~\cite{P09},
-Anglican~\cite{TMY+16}, Turing~\cite{GXG18}).
+of Stan. Our work differs from existing approaches in that we
+advocate enabling probabilistic programming in a general purpose
+programming language by leveraging existing capabilities of the
+language, rather than building a new language on top or besides
+an existing one.
+
+The need for integration between probabilistic programs and the
+surrounding software environment is well understood in the
+literature~\cite{TMY+16,Pyro18}. Probabilistic programming
+languages are often implemented as embedded domain-specific
+languages~\cite{SGG15,GXG18,TMY+16} and include a mechanism of calling
+functions from libraries of the host languages. Our work goes a
+step further in this direction: since the language of
+implementation of probabilistic models coincides with the host
+language, any library or user-defined function of the host
+language can be directly called from the probabilistic model,
+and model methods can be directly called from the host language.
 

### 2019-04-17 paper/paper.tex
fixed bibliography to look reasonably well

@@
 of Stan. Our work differs from existing approaches in that we
-advocate enabling probabilistic programming in a general purpose
+advocate enabling probabilistic programming in a general 
 programming language by leveraging existing capabilities of the
@@
 surrounding software environment is well understood in the
-literature~\cite{TMY+16,Pyro18}. Probabilistic programming
+literature~\cite{TMY+16,BCJ+19}. Probabilistic programming
 languages are often implemented as embedded domain-specific
@@
 
-Differentiable programming for machine learning (fundamentals,
-Stan, Tangent, Flux, Zygote, Pyro, Edward).
+Automatic differentiation is widely employed in machine
+learning~\cite{BPR+17,MBB+18} where it is also known as
+`differentiable programming', and is responsible for enabling
+efficient inference in many probabilistic programming
+systems~\cite{Stan17,GXG18,ISF+18,BCJ+19,THS+17}. Different
+automatic differentiation techniques~\cite{GW08} allow 
+different compromises between flexibility, efficiency, and
+feature-richness~\cite{SDT14,PGC+17,ISF+18}.  Automatic
+differentiation is usually implemented through either
+operator overloading~\cite{Stan17,PCG+17,GXG18} or source
+code transformation~\cite{ISF+18,MMW18}. Our work, too, relies
+on automatic differentiation, implemented as source code
+transformation. However, a novelty of our approach is that
+instead of using explicit directives to denote parts of code
+which needs to be differentiated, we rely on the type system of
+Go to selectively differentiate the code relevant for inference,
+thus combining advantages of both operator overloading and
+source code transformation.
 
@@
 hmc.Sample(m, x, samples)
-for i := 0; i != 1000; i++ {
+for i := 0; i != 5000; i++ {
 	x = <-samples

### 2019-04-17 paper/paper.tex
in progress

--- 2019-04-17 paper/paper.tex
@@
 decisions based on inference outcomes. However, complex
-server-side software systems make integration of probabilistic
+software systems make integration of probabilistic
 inference challenging.
@@
 
-\paragraph{Data interface} In a server-side application data
-for inference comes from a variety of sources: network,
-databases, distributed file systems, and in many different
-formats. Efficient inference depends on fast data access and
-updating. Libraries for data access and manipulation are
-available in the host environment. While the host environment
-can be used as a proxy retrieving and transforming the data,
-such as in the case of Stan~\cite{Stan17} integrations,
-sometimes direct access from the probabilistic code is the
-preferred option, for example when the data is streamed or
-retrieved conditionally.
-
-\paragraph{Integration and deployment} Deployment of server-side
-software systems is a delicate process involving automatic
-builds and maintenance of dependencies. Adding  a component,
-which possibly introduces additional software dependencies or
-even a separate runtime, complicates deployment. Minimizing the
-burden of probabilistic programming on the integration
-and deployment process should be a major consideration in
-design or selection of probabilistic programming tools.
-Probabilistic programming systems that are implemented or 
-provide an interface in a popular programming language, e.g.
-Python (Edward~\cite{THS+17}, Pyro~\cite{BCJ+19}) are easier
-to integrate and deploy, however the smaller the footprint
-of a probabilistic system, the easier is the adoption.
+\paragraph{Data interface} The data for inference may come from
+a variety of sources: network, databases, distributed file
+systems, and in many different formats. Efficient inference
+depends on fast data access and updating. Libraries for data
+access and manipulation are available in the host environment.
+While the host environment can be used as a proxy retrieving and
+transforming the data, such as in the case of Stan~\cite{Stan17}
+integrations, sometimes direct access from the probabilistic
+code is the preferred option, for example when the data is
+streamed or retrieved conditionally.
+
+\paragraph{Integration and deployment} Deployment of software
+systems is a delicate process involving automatic builds and
+maintenance of dependencies. Adding  a component, which possibly
+introduces additional software dependencies or even a separate
+runtime, complicates deployment. Minimizing the burden of
+probabilistic programming on the integration and deployment
+process should be a major consideration in design or selection
+of probabilistic programming tools.  Probabilistic programming
+systems that are implemented or provide an interface in a
+popular programming language, e.g.  Python
+(Edward~\cite{THS+17}, Pyro~\cite{BCJ+19}) are easier to
+integrate and deploy, however the smaller the footprint of a
+probabilistic system, the easier is the adoption.
 
@@
 guidelines, when followed, help easier integration of
-probabilistic programming inference into large-scale server-side
+probabilistic programming inference into large-scale 
 software systems.
@@
 implementation, which has recently become quite popular for
-computation-intensive server-side programming. Infergo is
-used in production environment for inference of mission-critical
+computation-intensive server-side programming. Infergo is used
+in production environment for inference of mission-critical
 algorithm parameters.
@@
 
+\subsection{Inference}
+
 \section{Design and Implementation Highlights}
@@
 \subsection{Automatic Differentiation}
-
-\subsection{Inference}
-
-\subsubsection{Optimization}
-
-\subsubsection{Full Posterior}
 
@@
 systems, making probabilistic programming inference more
-accessible in server-side applications. We described Infergo
-design choices and implementation techniques in intent to
-encourage broader adoption of probabilistic programming in
+accessible in large scale server-side applications. We described
+Infergo design choices and implementation techniques in intent
+to encourage broader adoption of probabilistic programming in
 production software systems and ease implementation of similar

### 2019-04-17 paper/paper.tex
in progress, slightly changed the program sstructure and consistently use general-purpose

@@
 	development. We argue that a similar probabilistic
-	programming facility can be added to most modern general
+	programming facility can be added to most modern general-purpose
 	programming languages.
@@
 	addition of probabilistic programming capabilities to other
-	general programming languages, we share design choices and
+	programming languages, we share design choices and
 	implementation techniques employed in development of
@@
 	is an extension of a regular programming toolbox allowing
-	algorithms implemented in general programming languages to
+	algorithms implemented in general-purpose programming languages to
 	have learnable parameters. In this view, statistical models
@@
 	instead of proposing yet another probabilistic programming
-	language, either genuinely different~\cite{MMR+07,Stan17} or derived from
-	an existing general programming language by extending the
-	syntax or changing the semantics~\cite{GMR+08,TMY+16,GXG18}, we advocate
-	enabling probabilistic inference on models written in a
-	general probabilistic programming language, the same one as
-	the language used to program the bulk of the system. We
-	formulate guidelines for such implementation, and based on
-	the guidelines, introduce a probabilistic programming
+	language, either genuinely different~\cite{MMR+07,Stan17} or
+	derived from an existing general-purpose programming
+	language by extending the syntax or changing the
+	semantics~\cite{GMR+08,TMY+16,GXG18}, we advocate enabling
+	probabilistic inference on models written in a
+	general-purpose probabilistic programming language, the same
+	one as the language used to program the bulk of the system.
+	We formulate guidelines for such implementation, and based
+	on the guidelines, introduce a probabilistic programming
 	facility for the Go programming language~\cite{Golang}. By
@@
 	we argue that a similar facility can be added to any modern
-	general purpose programming language, giving a production
-	system modeling and inference capabilities which do not fall
-	short from and sometimes exceed those of probabilistic
-	programming centric systems with custom languages and
-	runtimes.
+	general-purpose purpose programming language, giving a
+	production system modeling and inference capabilities which
+	do not fall short from and sometimes exceed those of
+	probabilistic programming-centric systems with custom
+	languages and runtimes.
 
@@
 			basis for implementation of a similar facility for
-			other general programming languages.
+			other general-purpose programming languages.
 	\end{itemize}
@@
 design and implementation of probabilistic programming
-languages. An apparent justification for a new programming
-language is that a probabilistic program has a different
-semantics~\cite{GHNR14} than a `regular' program of similar
-structure. Goodman and Stuhlm\"uller~\cite{GS15} offer a
+\textit{languages}. An apparent justification for a new
+programming language is that a probabilistic program has a
+different semantics~\cite{GHNR14} than a `regular' program of
+similar structure. Goodman and Stuhlm\"uller~\cite{GS15} offer a
 systematic approach to design and implementation of
 probabilistic programming languages based on extension of the
-syntax of a (subset of) regular programming language and
+syntax of a (subset of) general-purpose programming language and
 transformational compilation to continuation-passing form. Van
@@
 SlicStan~\cite{GGS19} is built on top of Stan as a
-source-to-source compiler, providing the model developer with
-a more intuitive language while retaining performance benefits
-of Stan. Our work differs from existing approaches in that we
-advocate enabling probabilistic programming in a general 
+source-to-source compiler, providing the model developer with a
+more intuitive language while retaining performance benefits of
+Stan. Our work differs from existing approaches in that we
+advocate enabling probabilistic programming in a general-purpose
 programming language by leveraging existing capabilities of the
@@
 languages are often implemented as embedded domain-specific
-languages~\cite{SGG15,GXG18,TMY+16} and include a mechanism of calling
-functions from libraries of the host languages. Our work goes a
-step further in this direction: since the language of
+languages~\cite{SGG15,GXG18,TMY+16} and include a mechanism of
+calling functions from libraries of the host languages. Our work
+goes a step further in this direction: since the language of
 implementation of probabilistic models coincides with the host
@@
 transformation. However, a novelty of our approach is that
-instead of using explicit directives to denote parts of code
-which needs to be differentiated, we rely on the type system of
-Go to selectively differentiate the code relevant for inference,
-thus combining advantages of both operator overloading and
-source code transformation.
+instead of using explicit calls or directives to denote parts of
+code which needs to be differentiated, we rely on the type
+system of Go to selectively differentiate the code relevant for
+inference, thus combining advantages of both operator
+overloading and source code transformation.
 
@@
 implementing the simulator twice, in the probabilistic model and
-in the host environment, the same code can serve both purposes.
-However to achieve this, the host language must coincide with
-the implementation language of the probabilistic model, on one
-hand, and allow a computationally efficient implementation of
-the simulation, on the other hand. Some probabilistic systems
-(Figaro~\cite{P09}, Anglican~\cite{TMY+16}, Turing~\cite{GXG18})
-are built with tight integration with the host environment in
-mind; more often than not though the probabilistic code is
-not trivial to re-use.
+in the host environment, one can invoke the same code for both
+purposes.  However to achieve this, the host language must
+coincide with the implementation language of the probabilistic
+model, on one hand, and allow a computationally efficient
+implementation of the simulation, on the other hand. Some
+probabilistic systems (Figaro~\cite{P09},
+Anglican~\cite{TMY+16}, Turing~\cite{GXG18}) are built with
+tight integration with the host environment in mind; more often
+than not though the probabilistic code is not trivial to re-use.
 
@@
 
-\subsection{Inference}
+\section{Inference}
 
@@
 A probabilistic programming facility similar to Infergo can be
-added to most modern general programming languages, in
+added to most modern general-purpose programming languages, in
 particular, those used for implementation of large-scale software

### 2019-04-17 paper/paper.tex
cleanups

@@
 code is the preferred option, for example when the data is
-streamed or retrieved conditionally.
+streamed or retrieved conditionally, as in active
+learning~\cite{SLB15}.
 
@@
 the distribution parameters. The Normal distribution has two
-parameters: mean $\mu$ and standard deviation
-$\sigma$. To ease the inference, positive $\sigma$ is transformed to
-unrestricted $\log \sigma$, and the parameter vector \lstinline{x} is $[\mu,
+parameters: mean $\mu$ and standard deviation $\sigma$. To ease
+the inference, positive $\sigma$ is transformed to unrestricted
+$\log \sigma$, and the parameter vector \lstinline{x} is $[\mu,
 \log \sigma]$. \lstinline{Observe} first imposes a prior on the
@@
 
+\cite{R16} \cite{KB15} \cite{N12} \cite{HG11} \cite{LN89}
+
 \section{Design and Implementation Highlights}

### 2019-04-17 paper/paper.tex
more on challenges

@@
 
+A flexible data interface is also benefitial for support of
+mini-batch optimization~\cite{LZC+14}. Mini-batch optimization
+is used when the data set is too big to fit in memory, and the
+objective function gradient is estimated based on mini-batches
+--- small portions of data. Different models and inference
+objectives may require different mini-batch loading schemes.
+For best performance of mini-batch optimization, it should be
+possible to program loading of mini-batches on case-by-case
+basis.
+
 \paragraph{Integration and deployment} Deployment of software

### 2019-04-18 paper/paper.tex
cited onward

@@
 	probabilistic programming-centric systems with custom
-	languages and runtimes.
+	languages and runtimes. Availability of such facilities
+	in major general-purpose programming languages would free
+	probabilistic programming researchers and practitioners
+	alike from language wars~\cite{SH14} and foster practical
+	applications of probabilistic programming.
 
@@
 different semantics~\cite{GHNR14} than a `regular' program of
-similar structure. Goodman and Stuhlm\"uller~\cite{GS15} offer a
+similar structure. \citet{GS15} offer a
 systematic approach to design and implementation of
@@
 syntax of a (subset of) general-purpose programming language and
-transformational compilation to continuation-passing form. Van
-de Meent et al~\cite{MPY+18} give a comprehensive overview of
+transformational compilation to continuation-passing form. 
+\citet{MPY+18} give a comprehensive overview of
 available choices of design and implementation of first-order
@@
 differentiation is usually implemented through either
-operator overloading~\cite{Stan17,PCG+17,GXG18} or source
+operator overloading~\cite{Stan17,PGC+17,GXG18} or source
 code transformation~\cite{ISF+18,MMW18}. Our work, too, relies
@@
 probabilistic system, the easier is the adoption.
+An obstacle in integration of probabilistic programming lies
+also in adoption of the probabilistic programming
+language~\cite{MR12}, if the latter differs from the
+implementation language of the rest of the system.
 
@@
 
-In Infergo, the inference is performed on a model. An Infergo
-model is a data type and methods defined on the data type.
-Inference algorithms accept a model object as an argument.
-A model object usually encapsulates observations --- the data 
-to which the parameter values are adapted.
+In Infergo, the inference is performed on a model. Just like in
+Stan~\cite{Stan17} models, latent random variables in Infergo
+models are continuous. Additionally, Infergo can express and
+perform inference on non-deterministic models by including
+stochastic choices in the model code.
+
+An Infergo model is a data type and methods defined on the data
+type.  Inference algorithms accept a model object as an
+argument.  A model object usually encapsulates observations ---
+the data to which the parameter values are adapted.
 

### 2019-04-18 paper/paper.tex
citations of dissertations and a note of comparison of probabilisti languages

@@
 host programming language, passes the data and the model to an
-inference algorithm, and post-processes the results in the
-host programming language again to make algorithmic
-decisions based on inference outcomes. However, complex
-software systems make integration of probabilistic
-inference challenging.
+inference algorithm, and post-processes the results in the host
+programming language again to make algorithmic decisions based
+on inference outcomes. To choose the best one for each particular
+application, probabilistic programming languages and systems are
+customarily compared by the expressive power (classes of
+probabilistic models that can be represented), conciseness (e.g.
+the number of lines of code to describe a particular model), and
+time and space efficiency of inference
+algorithms~\cite{WVM14,P16,R17,GXG18}.  However, complex
+software systems make integration of probabilistic inference
+challenging, and considerations between beyond expressiveness
+and performance come into play.
 

### 2019-04-18 paper/paper.tex
small language fix

@@
 programming language again to make algorithmic decisions based
-on inference outcomes. To choose the best one for each particular
-application, probabilistic programming languages and systems are
-customarily compared by the expressive power (classes of
-probabilistic models that can be represented), conciseness (e.g.
-the number of lines of code to describe a particular model), and
-time and space efficiency of inference
+on inference outcomes. To choose the best option for each
+particular application, probabilistic programming languages and
+systems are customarily compared by the expressive power
+(classes of probabilistic models that can be represented),
+conciseness (e.g.  the number of lines of code to describe a
+particular model), and time and space efficiency of inference
 algorithms~\cite{WVM14,P16,R17,GXG18}.  However, complex

### 2019-04-18 paper/paper.tex
inference in progress

@@
 software systems make integration of probabilistic inference
-challenging, and considerations between beyond expressiveness
+challenging, and considerations beyond expressiveness
 and performance come into play.
@@
 
-A flexible data interface is also benefitial for support of
+A flexible data interface is also beneficial for support of
 mini-batch optimization~\cite{LZC+14}. Mini-batch optimization
@@
 
+Out of the box, Infergo offers 
+\begin{itemize}
+	\item maximum \textit{a posteriori} estimation by stochastic
+		gradient descent~\cite{R16}; 
+	\item full posterior inference by Hamiltonian Monte
+		Carlo~\cite{N12,HG11}.
+\end{itemize}
+
+When only a point estimate of the model parameters is required,
+as, for example, in the simulation-inference case, stochastic
+gradient descent is a fast and robust inference family. Infergo
+offers stochastic gradient descent with momentum and
+Adam~\cite{KB15}.  The inference is performed by calling the
+\lstinline{Step} method of the optimizer until a termination
+condition is met.
+
+When the full posterior must be recovered for integration,
+Hamiltonian Monte Carlo (HMC) can be used instead. Vanilla HMC
+as well as the No-U-Turn sampler (NUTS)~\cite{HG11} are
+provided. To support lazy any-time computation, the inference
+runs in a separate \textit{goroutine}, connected by a channel to
+the caller, and asynchronously writes samples to the channel.
+The caller reads as many samples from the channel as needed, and
+further post-processes them.
+\begin{lstlisting}
+\end{lstlising}
+
 \cite{R16} \cite{KB15} \cite{N12} \cite{HG11} \cite{LN89}

### 2019-04-18 paper/paper.tex
inference in progress, code snippets due

@@
 \begin{itemize}
-	\item maximum \textit{a posteriori} estimation by stochastic
+	\item maximum \textit{a posteriori} (MAP) estimation by stochastic
 		gradient descent~\cite{R16}; 
@@
 condition is met.
+\begin{lstlisting}
+\end{lstlisting}
 
@@
 \begin{lstlisting}
-\end{lstlising}
-
-\cite{R16} \cite{KB15} \cite{N12} \cite{HG11} \cite{LN89}
+\end{lstlisting}
+
+Other inference schemes, most notably automatic differentiation
+variational inference~\cite{KTR+17}, are planned for addition
+in future versions of Infergo. However, an inference algorithm
+does not have to be a part of Infergo to work with Infergo
+models. Since infergo models operate on the built-in Go floating
+point type, \lstinline{float64}, any third-party inference or
+optimization library can be used as well.
+
+As an example, the Gonum~\cite{Gonum} library for numerical
+computations contains an optimization package offering several
+algorithms for gradient-based optimization. A Gonum optimization
+algorithm require the function to optimize as well as the
+function's gradient as parameters. An Infergo model can be
+trivially wrapped for use with Gonum, and Infergo provides a
+convenience function
+\begin{lstlisting}
+func FuncGrad(m Model) (
+	Func func(x []float64) float64,
+	Grad func(grad []float64, x []float64)
+		[]float64)
+\end{lstlisting}
+which accepts a model and returns the model's
+\lstinline{Observe} method and its gradient suitable for passing
+to Gonum optimization algorithms. Among available algorithms are
+advanced versions of stochastic gradient descent, as well as
+algorithms of BFGS family, in particular LBFGS~\cite{LN89},
+providing for a more efficient MAP estimation than stochastic
+gradient descent at the cost of higher memory consumption.
+
+Other third-party implementations of optimization algorithms can
+be used just as easily, thanks to Infergo models being pure Go
+code operating on built-in Go data types.
 

### 2019-04-18 paper/paper.tex
inference drafted

@@
 \usepackage{GoMono}
+
+%\linespread{1.0}
 
@@
 	g func(value float64,
-		   params ...float64) []float64).
+		   params ...float64) []float64)
 \end{lstlisting} For
@@
 \lstinline{Step} method of the optimizer until a termination
-condition is met.
+condition is met:
 \begin{lstlisting}
+for !<termination condition> {
+    optimizer.Step(model, parameters)
+}
 \end{lstlisting}
@@
 The caller reads as many samples from the channel as needed, and
-further post-processes them.
+further postprocesses them:
 \begin{lstlisting}
+samples := make(chan []float64)
+hmc.Sample(model, parameters, samples)
+for !<termination condition> {
+	sample = <-samples
+	// postprocess the sample
+}
+hmc.Stop()
 \end{lstlisting}
@@
 does not have to be a part of Infergo to work with Infergo
-models. Since infergo models operate on the built-in Go floating
+models. Since Infergo models operate on a built-in Go floating
 point type, \lstinline{float64}, any third-party inference or
@@
 algorithms for gradient-based optimization. A Gonum optimization
-algorithm require the function to optimize as well as the
-function's gradient as parameters. An Infergo model can be
+algorithm requires the function to optimize, as well as the
+function's gradient, as parameters. An Infergo model can be
 trivially wrapped for use with Gonum, and Infergo provides a
@@
 providing for a more efficient MAP estimation than stochastic
-gradient descent at the cost of higher memory consumption.
+gradient descent at the cost of higher memory consumption:
+\begin{lstlisting}
+function, gradient := FuncGrad(model)
+problem := gonum.Problem{Func: function, 
+                         Grad: gradient}
+result, err := gonum.Minimize(problem, parameters)
+\end{lstlisting}
 
@@
 
+\clearpage
+\clearpage
+
 \bibliography{refs}

### 2019-04-19 paper/paper.tex
half way to the  paper, pagewise
language choice

@@
 	addition of probabilistic programming capabilities to other
-	programming languages, we share design choices and
-	implementation techniques employed in development of
-	Infergo. We illustrate applicability of Infergo to various
-	use cases on case studies, and evaluate Infergo's
-	performance on several benchmarks, comparing Infergo to
-	dedicated inference-centric probabilistic programming
-	systems.
+	programming languages, we share implementation choices and
+	techniques employed in development of Infergo. We illustrate
+	applicability of Infergo to various use cases on case
+	studies, and evaluate Infergo's performance on several
+	benchmarks, comparing Infergo to dedicated inference-centric
+	probabilistic programming systems.
 \end{abstract}
@@
 	propagation~\cite{MWG+10}, extensions of Sequential Monte
-	Carlo~\cite{WVM14,MYM+15,PWD+14}, variational
+	Carlo~\cite{WVM14,MYM+15,PWD+14,RNL+2016}, variational
 	inference~\cite{WW13,KTR+17}, gradient-based
@@
 	facility for the Go programming language~\cite{Golang}. By
-	explaining our design choices and implementation techniques,
+	explaining our implementation choices and techniques,
 	we argue that a similar facility can be added to any modern
@@
 			programming facility for the Go programming language.
-		\item Design and implementation highlights outlining
-			important choices we made in implementation of the
+		\item Implementation highlights outlining important
+			choices we made in implementation of the
 			probabilistic programming facility, and providing a
@@
 programming language is that a probabilistic program has a
-different semantics~\cite{GHNR14} than a `regular' program of
-similar structure. \citet{GS15} offer a
-systematic approach to design and implementation of
-probabilistic programming languages based on extension of the
-syntax of a (subset of) general-purpose programming language and
-transformational compilation to continuation-passing form. 
-\citet{MPY+18} give a comprehensive overview of
-available choices of design and implementation of first-order
+different semantics~\cite{GHNR14,SYH+16} than a `regular'
+program of similar structure. \citet{GS15} offer a systematic
+approach to design and implementation of probabilistic
+programming languages based on extension of the syntax of a
+(subset of) general-purpose programming language and
+transformational compilation to continuation-passing
+style~\cite{AJ89,A07}.  \citet{MPY+18} give a comprehensive overview
+of available choices of design and implementation of first-order
 and higher-order probabilistic programming languages.
@@
 
-The Go programming language and development environment offer
-capabilities which made implementation of Infergo affordable.
-
-\begin{itemize}
-\item The Go parser and abstract syntax tree serializer are
-	a part of the standard library. Parsing, transforming,
-	and generating Go source code is straightforward and
-	effortless.
-\item Type inference (or \textit{type checking} as it is
-	called in the Go ecosystem), also provided in the
-	standard library, augments parsing and allows to
-	selectively apply transformation-based automatic
-	differentiation  based on static expression types. 
-\item Go compiles and runs fast. Fast compilation and
-	execution speeds allow to use the same facility for both
-	exploratory design of probabilistic models and for
-	inference in production environment.
-\item Go offers efficient parallel execution as a
-	first-class feature, via so-called \textit{goroutines}.
-	Goroutines streamline implementation of sampling-based
-	inference algorithms. Sample generators and consumers
-	are run in parallel, communicating through channels. 
-	Inference is easy to parallelize in order to exploit
-	hardware multi-processing, and samples are retrieved
-	lazily for postprocessing. 
-\end{itemize}
-
 \subsection{A Basic Example}
@@
 Any user-defined function of appropriate kind can be used as an
-elemental inside model methods. An elemental must be declared
+elemental inside model methods. The gradient must be registered
 for the function using a call to
@@
 \end{lstlisting}
-registers the gradient for function \lstinline{Sigm}.
+registers the gradient for function
+\lstinline{Sigm}$(x)\coloneq\left(1+e^{-x}\right)^{-1}$.
 
@@
 
-\section{Design and Implementation Highlights}
+\section{Implementation Highlights}
+
+In our journey towards implementation and release of Infergo,
 
 \subsection{Choice of Programming Language}
+
+Python~\cite{Python95}, Scala~\cite{Scala06}, and
+Julia~\cite{Julia14} have been popular choices for development
+of probabilistic programming languages and systems integrated
+with general-purpose programming
+languages~\cite{S16,MMW18,BCJ+19,P09,Rainier,GXG18,ISF+18}.
+Each of these languages has advantages for implementing a
+probabilistic programming facility. However, for a reference
+implementation of a facility best suited for probabilistic
+inference in a porduction software system, we were looking
+for a language that \begin{itemize}
+	\item compiles into fast and compact code;
+	\item does not incur dependency on a heavy runtime; and
+	\item is a popular choice for building server-side software.
+\end{itemize}
+In addition, since we were going to automatically differentiate
+the language's source code, we were looking for a relatively
+small language, preferably one for which parsing and program
+analysis tools are readily available.
+
+We chose Go~\cite{Golang}. Go is a small but expressive
+programming language widely used for implementing
+computationally-intensive server-side software systems. 
+The standard libraries and development environment offer
+capabilities which made implementation of Infergo affordable
+and streamlined:
+
+\begin{itemize}
+	\item The Go parser and abstract syntax tree serializer are
+		a part of the standard library. Parsing, transforming,
+		and generating Go source code is straightforward and
+		effortless.
+	\item Type inference (or \textit{type checking} as it is
+		called in the Go ecosystem), also provided in the
+		standard library, augments parsing and allows to
+		selectively apply transformation-based automatic
+		differentiation  based on static expression types. 
+	\item Go provides reflection capabilities for almost any aspect
+		of the running program, without compromising efficiency.
+		Reflection gives the flexibility highly desirable to
+		implement features such as calling Go library
+		functions from the model code.
+	\item Go compiles and runs fast. Fast compilation and
+		execution speeds allow to use the same facility for both
+		exploratory design of probabilistic models and for
+		inference in production environment.
+	\item Go is available on many platforms and for a wide range
+		of operating systems and environments, from portable
+		devices to supercomputers. Go programs can be
+		cross-compiled, and go programs can be reliably
+		developed and deployed in heterogeneous environments.
+	\item Go offers efficient parallel execution as a
+		first-class feature, via so-called \textit{goroutines}.
+		Goroutines streamline implementation of sampling-based
+		inference algorithms. Sample generators and consumers
+		are run in parallel, communicating through channels. 
+		Inference is easy to parallelize in order to exploit
+		hardware multi-processing, and samples are retrieved
+		lazily for postprocessing. 
+\end{itemize}
+
+Considering and choosing Go for development of what later became
+Infergo also affected the shape of the software system in which
+Infergo was deployed first. Formerly implemented mostly in
+Python, and both enjoying a wide choice of libraries for machine
+learning and data processing and suffering from limitations of
+Python as a language for implementation of computation-intensive
+algorithms, the system has been gradually drifting towards
+adopting Go, with performance-critical parts re-implemented and
+running faster, with better hardware and network utilization,
+and simpler and more maintainable code for data processing,
+analysis, and decision making.
 
@@
     for j := 0; j != m.NComp; j++ {
-      lj := Normal.Logp(mu[j], sigma[j], m.Data[i])
+      lj := Normal.Logp(mu[j], sigma[j],
+	                    m.Data[i])
       if j == 0 {
@@
 
-\subsection{Performance}
+\section{Performance}
 
@@
 accessible in large scale server-side applications. We described
-Infergo design choices and implementation techniques in intent
+Infergo implementation choices and techniques in intent
 to encourage broader adoption of probabilistic programming in

### 2019-04-19 paper/paper.tex
insights prologue

@@
 
-In our journey towards implementation and release of Infergo,
+In our journey towards implementation and deployment of Infergo, we
+considered options and made implementation decisions which we
+believe to have been crucial for addressing the challenges and
+following the guidelines we formulated for implementation of a
+deployable probabilistic programming facility. While different
+choices may be as well suited for the purpose as the ones we
+made, we feel it important to share our contemplations and
+justify decisions, to which the rest of this section is
+dedicated, from choosing the implementation language, through
+details of automatic differentiation, to support for inference
+on streaming and stochastic observations.
 
@@
 		devices to supercomputers. Go programs can be
-		cross-compiled, and go programs can be reliably
+		cross-compiled, and Go programs can be reliably
 		developed and deployed in heterogeneous environments.

### 2019-04-19 paper/paper.tex
restructured slightly
plan for autodiff
autodiff in progress

@@
 
-\section{Model Syntax and Rules of Differentiation}
+\subsection{Model Syntax and Rules of Differentiation}
 
@@
 
-\section{Inference}
+\subsection{Inference}
 
@@
 
-\subsection{Composability of Models}
+Infergo employs reverse-mode autodifferentiation via source code
+transformation~\cite{GW08}. The reverse mode is the default
+choice in machine learning applications, because it works best
+with differentiating a scalar return value with respect to
+multiple parameters. Automatic differentiation is usually
+implemented using either operator overloading or source code
+transformation. Operator overloading is often the method of
+choice~\cite{Stan17,GXG18,PGC+17} due to relative simplicity,
+but the differentiated code must use a special type (on which
+the operators are overloaded) for representing floating point
+numbers. Regardless of the issue of computational overhead, this
+restricts the language used to specify the model to operations on
+that special type and impairs interoperability: library
+functions cannot be directly called but have to use an adaptor,
+or a `lifted' version of the standard library must be supplied.
+Similarly, the data must be cast into the special type before it
+is passed to the model for inference.
+
+Source code transformation allows to automatically differentiate
+program code operating on the built-in floating point type. This
+method involves generation of new code, for the function
+gradient or for the function itself, ahead of time or on the
+fly. To achieve the best performance, some implementations
+generate static code for computing the function
+gradient~\cite{MMW18,I18}. This requires a rather elaborated
+source code analysis, and although a significant progress has
+been made in this direction, not all language constructs are
+supported. Alternatively, the code of the function itself can be
+transformed by augmenting arithmetics and function calls on the
+floating point type by function calls that record the
+computations on the \textit{tape}, just like in the case of
+operator overloading. Then, the backward pass is identical to
+that of operator overloading. This method allows the
+use of the built-in floating point type along with
+straightforward virtually unrestricted language for model
+specification, at the cost of slight performance loss.
+
+This last method is used in Infergo: the model is written in Go
+and is placed in a separate Go package. A command-line utility
+is run on the package to generate a sub-package containing a
+differentiated version of the model.  Both the original and the
+differentiated model are available simultaneously in the calling
+code, so that invocations of model methods which do not require
+computation of gradients can be performed without the overhead
+of automatic differentiation.  However, the performance loss is
+rarely a major issue --- differentiated versions of model
+methods are only 2--4 times slower on models from the case
+studies~\ref{sec:case-studies}, and the backward pass, which
+computes the gradient, is roughly as fast as the original,
+undifferentiated version. At least partially, fast backward pass
+is due to careful implementation and thorough profiling of
+the tape.
+
+% Elementals. Differentiation
+%of calls. Distributions. Composability. Nested inference.
 
@@
 
-\section{Case Studies}
-\label{sec:case-studies}
+\section{Case Studies} \label{sec:case-studies}
 
@@
 
-\begin{figure*}
+\begin{figure*} \begin{minipage}{0.45\textwidth}
+	\begin{lstlisting}[language=Stan,framexleftmargin=10pt,numbers=left]
+	data { int<lower=0> J; vector[J] y; vector<lower=0>[J]
+	sigma; }
+
+parameters { real mu; real<lower=0> tau; vector[J] eta; }
+
+transformed parameters { vector[J] theta; theta = mu + tau *
+	eta; }
+
+model { eta ~ normal(0, 1); y ~ normal(theta, sigma); }
+	\end{lstlisting}
+
+\centering a. Stan \end{minipage} \hfill
 \begin{minipage}{0.45\textwidth}
-	\begin{lstlisting}[language=Stan,framexleftmargin=10pt,numbers=left]
-data {
-  int<lower=0> J;
-  vector[J] y;
-  vector<lower=0>[J] sigma;
-}
-
-parameters {
-  real mu;
-  real<lower=0> tau;
-  vector[J] eta;
-}
-
-transformed parameters {
-  vector[J] theta;
-  theta = mu + tau * eta;
-}
-
-model {
-  eta ~ normal(0, 1);
-  y ~ normal(theta, sigma);
-}
-\end{lstlisting}
-
-\centering
-a. Stan
-	\end{minipage}
-	\hfill
-	\begin{minipage}{0.45\textwidth}
-\begin{lstlisting}[framexleftmargin=10pt,numbers=left]
-type Model struct {
-  J          int
-  Y          []float64
-  Sigma      []float64
-}
-
-func (m *Model) Observe(x []float64) float64 {
-  mu := x[0]
-  tau := math.Exp(x[1])
-  eta := x[2:]
+	\begin{lstlisting}[framexleftmargin=10pt,numbers=left] type
+	Model struct { J          int Y          []float64 Sigma
+	[]float64 }
+
+func (m *Model) Observe(x []float64) float64 { mu := x[0] tau :=
+	math.Exp(x[1]) eta := x[2:]
 
@@
 
-  for i, y := range m.Y {
-    theta := mu + tau*eta[i]
-    ll += Normal.Logp(theta, m.Sigma[i], y)
-  }
-
-  return ll
-}
-\end{lstlisting}
-\centering
-b. Infergo
-	\end{minipage}
-	\caption{8 schools: Stan vs. Infergo. The Go implementation
-	has a similar length and structure to the Stan model.}
-	\label{fig:8-schools}
-\end{figure*}
+  for i, y := range m.Y { theta := mu + tau*eta[i] ll +=
+	Normal.Logp(theta, m.Sigma[i], y) }
+
+  return ll } \end{lstlisting} \centering b. Infergo
+\end{minipage} \caption{8 schools: Stan vs. Infergo. The Go
+implementation has a similar length and structure to the Stan
+model.} \label{fig:8-schools} \end{figure*}
 
@@
 \begin{figure}
-\begin{lstlisting}[framexleftmargin=10pt,numbers=left]
-type Model struct {
-  Data  [][]float64
-}
-
-func (m *Model) Observe(x []float64)
-    float64 {
-  ll := 0.
-
-  alpha, beta := x[0], x[1]
-  sigma := math.Exp(x[2])
-
-  for i := range m.Data {
-    ll += Normal.Logp(
-      m.Simulate(m.Data[i][0], alpha, beta),
-      sigma, m.Data[i][1])
-  }
-  return ll
-
-// Simulate predicts y for x based
-// on inferred parameters.
-func (m *Model) Simulate(x, alpha, beta float64)
-    float64 {
-  y := alpha +beta*x
-  return y
-}
-\end{lstlisting}
-\caption{Linear regression: the simulator is re-used in both
-	inference and prediction.}
-\label{fig:linear-regression}
-\end{figure}
+	\begin{lstlisting}[framexleftmargin=10pt,numbers=left] type
+	Model struct { Data  [][]float64 }
+
+func (m *Model) Observe(x []float64) float64 { ll := 0.
+
+  alpha, beta := x[0], x[1] sigma := math.Exp(x[2])
+
+  for i := range m.Data { ll += Normal.Logp(
+	m.Simulate(m.Data[i][0], alpha, beta), sigma, m.Data[i][1])
+	} return ll
+
+// Simulate predicts y for x based // on inferred parameters.
+	func (m *Model) Simulate(x, alpha, beta float64) float64 { y
+	:= alpha +beta*x return y } \end{lstlisting} \caption{Linear
+	regression: the simulator is re-used in both inference and
+prediction.} \label{fig:linear-regression} \end{figure}
 
@@
 \begin{figure}
-\begin{lstlisting}[framexleftmargin=10pt,numbers=left]
-type Model struct {
-  Data  []float64 // samples
-  NComp int       // number of components
-}
-
-func (m *Model) Observe(x []float64) float64 {
-  ll := 0.0
-  mu := make([]float64, m.NComp)
-  sigma := make([]float64, m.NComp)
-
-  // Fetch component parameters
-  for j := 0; j != m.NComp; j++ {
-    mu[j] = x[2*j]
-    sigma[j] = math.Exp(x[2*j+1])
-  }
-
-  // Compute log likelihood of mixture
-  // given the data
-  for i := 0; i != len(m.Data); i++ {
-    var l float64
-    for j := 0; j != m.NComp; j++ {
-      lj := Normal.Logp(mu[j], sigma[j],
-	                    m.Data[i])
-      if j == 0 {
-        l = lj
-      } else {
-        l = mathx.LogSumExp(l, lj)
-      }
-    }
-    ll += l
-  }
-  return ll
-}
-\end{lstlisting}
-\caption{Gaussian mixture: Infergo can handle models with
-	dynamic control structure.}
-\label{fig:gaussian-mixture}
-\end{figure}
+	\begin{lstlisting}[framexleftmargin=10pt,numbers=left] type
+	Model struct { Data  []float64 // samples NComp int       //
+	number of components }
+
+func (m *Model) Observe(x []float64) float64 { ll := 0.0 mu :=
+	make([]float64, m.NComp) sigma := make([]float64, m.NComp)
+
+  // Fetch component parameters for j := 0; j != m.NComp; j++ {
+	  mu[j] = x[2*j] sigma[j] = math.Exp(x[2*j+1]) }
+
+  // Compute log likelihood of mixture // given the data for i
+	:= 0; i != len(m.Data); i++ { var l float64 for j := 0; j !=
+	m.NComp; j++ { lj := Normal.Logp(mu[j], sigma[j], m.Data[i])
+	if j == 0 { l = lj } else { l = mathx.LogSumExp(l, lj) } }
+	ll += l } return ll } \end{lstlisting} \caption{Gaussian
+	mixture: Infergo can handle models with dynamic control
+	structure.} \label{fig:gaussian-mixture} \end{figure}
 
@@
 
-{\smaller
-\begin{table}[H]
-\caption{Memory and running times for 1000 iterations
-of HMC with 10 leapfrog steps.}
-\label{tab:memory-runtime}
-\begin{tabular}{r | c |  c | c}
-	{\it model}  & \multicolumn{2}{c|}{\it time} & {\it memory} \\ 
-	 & {\it compilation} & {\it execution} & \\\hline
-	8 schools & 0.15s & 0.6s & 5.5MB \\
-	10D normal, 100 points & 0.15s & 3.0s & 5.7MB \\
-	50D normal, 100 points & 0.15s & 13.0s & 5.8MB 
-\end{tabular}
+{\smaller \begin{table}[H] \caption{Memory and running times for
+	1000 iterations of HMC with 10 leapfrog steps.}
+	\label{tab:memory-runtime} \begin{tabular}{r | c |  c | c}
+		{\it model}  & \multicolumn{2}{c|}{\it time} & {\it
+		memory} \\ & {\it compilation} & {\it execution} &
+		\\\hline 8 schools & 0.15s & 0.6s & 5.5MB \\ 10D normal,
+		100 points & 0.15s & 3.0s & 5.7MB \\ 50D normal, 100
+		points & 0.15s & 13.0s & 5.8MB \end{tabular}
 \end{table}}
@@
 added to most modern general-purpose programming languages, in
-particular, those used for implementation of large-scale software
-systems, making probabilistic programming inference more
-accessible in large scale server-side applications. We described
-Infergo implementation choices and techniques in intent
-to encourage broader adoption of probabilistic programming in
-production software systems and ease implementation of similar
-facilities in other languages and software environments.
+particular, those used for implementation of large-scale
+software systems, making probabilistic programming inference
+more accessible in large scale server-side applications. We
+described Infergo implementation choices and techniques in
+intent to encourage broader adoption of probabilistic
+programming in production software systems and ease
+implementation of similar facilities in other languages and
+software environments.
 
@@
 
-\clearpage
-\clearpage
+\clearpage \clearpage
 

### 2019-04-19 paper/paper.tex
autodiff in progress

@@
 
-\section{Case Studies} \label{sec:case-studies}
+\section{Case Studies}
+\label{sec:case-studies}
 
@@
 
-\begin{figure*} \begin{minipage}{0.45\textwidth}
+\begin{figure*}
+	\begin{minipage}{0.45\textwidth}
 	\begin{lstlisting}[language=Stan,framexleftmargin=10pt,numbers=left]
-	data { int<lower=0> J; vector[J] y; vector<lower=0>[J]
-	sigma; }
-
-parameters { real mu; real<lower=0> tau; vector[J] eta; }
-
-transformed parameters { vector[J] theta; theta = mu + tau *
-	eta; }
-
-model { eta ~ normal(0, 1); y ~ normal(theta, sigma); }
+data {
+  int<lower=0> J;
+  vector[J] y;
+  vector<lower=0>[J] sigma;
+}
+
+parameters {
+  real mu;
+  real<lower=0> tau;
+  vector[J] eta;
+}
+
+transformed parameters {
+  vector[J] theta;
+  theta = mu + tau * eta;
+}
+
+model {
+  eta ~ normal(0, 1);
+  y ~ normal(theta, sigma);
+}
 \end{lstlisting}
 
-\centering a. Stan \end{minipage} \hfill
+\centering
+a. Stan
+	\end{minipage}
+	\hfill
 	\begin{minipage}{0.45\textwidth}
-	\begin{lstlisting}[framexleftmargin=10pt,numbers=left] type
-	Model struct { J          int Y          []float64 Sigma
-	[]float64 }
-
-func (m *Model) Observe(x []float64) float64 { mu := x[0] tau :=
-	math.Exp(x[1]) eta := x[2:]
+\begin{lstlisting}[framexleftmargin=10pt,numbers=left]
+type Model struct {
+  J          int
+  Y          []float64
+  Sigma      []float64
+}
+
+func (m *Model) Observe(x []float64) float64 {
+  mu := x[0]
+  tau := math.Exp(x[1])
+  eta := x[2:]
 
@@
 
-  for i, y := range m.Y { theta := mu + tau*eta[i] ll +=
-	Normal.Logp(theta, m.Sigma[i], y) }
-
-  return ll } \end{lstlisting} \centering b. Infergo
-\end{minipage} \caption{8 schools: Stan vs. Infergo. The Go
-implementation has a similar length and structure to the Stan
-model.} \label{fig:8-schools} \end{figure*}
+  for i, y := range m.Y {
+    theta := mu + tau*eta[i]
+    ll += Normal.Logp(theta, m.Sigma[i], y)
+  }
+
+  return ll
+}
+\end{lstlisting}
+\centering
+b. Infergo
+	\end{minipage}
+	\caption{8 schools: Stan vs. Infergo. The Go implementation
+	has a similar length and structure to the Stan model.}
+	\label{fig:8-schools}
+\end{figure*}
 
@@
 \begin{figure}
-	\begin{lstlisting}[framexleftmargin=10pt,numbers=left] type
-	Model struct { Data  [][]float64 }
-
-func (m *Model) Observe(x []float64) float64 { ll := 0.
-
-  alpha, beta := x[0], x[1] sigma := math.Exp(x[2])
-
-  for i := range m.Data { ll += Normal.Logp(
-	m.Simulate(m.Data[i][0], alpha, beta), sigma, m.Data[i][1])
-	} return ll
-
-// Simulate predicts y for x based // on inferred parameters.
-	func (m *Model) Simulate(x, alpha, beta float64) float64 { y
-	:= alpha +beta*x return y } \end{lstlisting} \caption{Linear
-	regression: the simulator is re-used in both inference and
-prediction.} \label{fig:linear-regression} \end{figure}
+\begin{lstlisting}[framexleftmargin=10pt,numbers=left]
+type Model struct {
+  Data  [][]float64
+}
+
+func (m *Model) Observe(x []float64)
+    float64 {
+  ll := 0.
+
+  alpha, beta := x[0], x[1]
+  sigma := math.Exp(x[2])
+
+  for i := range m.Data {
+    ll += Normal.Logp(
+      m.Simulate(m.Data[i][0], alpha, beta),
+      sigma, m.Data[i][1])
+  }
+  return ll
+
+// Simulate predicts y for x based
+// on inferred parameters.
+func (m *Model) Simulate(x, alpha, beta float64)
+    float64 {
+  y := alpha +beta*x
+  return y
+}
+\end{lstlisting}
+\caption{Linear regression: the simulator is re-used in both
+	inference and prediction.}
+\label{fig:linear-regression}
+\end{figure}
 
@@
 \begin{figure}
-	\begin{lstlisting}[framexleftmargin=10pt,numbers=left] type
-	Model struct { Data  []float64 // samples NComp int       //
-	number of components }
-
-func (m *Model) Observe(x []float64) float64 { ll := 0.0 mu :=
-	make([]float64, m.NComp) sigma := make([]float64, m.NComp)
-
-  // Fetch component parameters for j := 0; j != m.NComp; j++ {
-	  mu[j] = x[2*j] sigma[j] = math.Exp(x[2*j+1]) }
-
-  // Compute log likelihood of mixture // given the data for i
-	:= 0; i != len(m.Data); i++ { var l float64 for j := 0; j !=
-	m.NComp; j++ { lj := Normal.Logp(mu[j], sigma[j], m.Data[i])
-	if j == 0 { l = lj } else { l = mathx.LogSumExp(l, lj) } }
-	ll += l } return ll } \end{lstlisting} \caption{Gaussian
-	mixture: Infergo can handle models with dynamic control
-	structure.} \label{fig:gaussian-mixture} \end{figure}
+\begin{lstlisting}[framexleftmargin=10pt,numbers=left]
+type Model struct {
+  Data  []float64 // samples
+  NComp int       // number of components
+}
+
+func (m *Model) Observe(x []float64) float64 {
+  ll := 0.0
+  mu := make([]float64, m.NComp)
+  sigma := make([]float64, m.NComp)
+
+  // Fetch component parameters
+  for j := 0; j != m.NComp; j++ {
+    mu[j] = x[2*j]
+    sigma[j] = math.Exp(x[2*j+1])
+  }
+
+  // Compute log likelihood of mixture
+  // given the data
+  for i := 0; i != len(m.Data); i++ {
+    var l float64
+    for j := 0; j != m.NComp; j++ {
+      lj := Normal.Logp(mu[j], sigma[j],
+	                    m.Data[i])
+      if j == 0 {
+        l = lj
+      } else {
+        l = mathx.LogSumExp(l, lj)
+      }
+    }
+    ll += l
+  }
+  return ll
+}
+\end{lstlisting}
+\caption{Gaussian mixture: Infergo can handle models with
+	dynamic control structure.}
+\label{fig:gaussian-mixture}
+\end{figure}
 
@@
 
-{\smaller \begin{table}[H] \caption{Memory and running times for
-	1000 iterations of HMC with 10 leapfrog steps.}
-	\label{tab:memory-runtime} \begin{tabular}{r | c |  c | c}
-		{\it model}  & \multicolumn{2}{c|}{\it time} & {\it
-		memory} \\ & {\it compilation} & {\it execution} &
-		\\\hline 8 schools & 0.15s & 0.6s & 5.5MB \\ 10D normal,
-		100 points & 0.15s & 3.0s & 5.7MB \\ 50D normal, 100
-		points & 0.15s & 13.0s & 5.8MB \end{tabular}
+{\smaller
+\begin{table}[H]
+\caption{Memory and running times for 1000 iterations
+of HMC with 10 leapfrog steps.}
+\label{tab:memory-runtime}
+\begin{tabular}{r | c |  c | c}
+	{\it model}  & \multicolumn{2}{c|}{\it time} & {\it memory} \\ 
+	 & {\it compilation} & {\it execution} & \\\hline
+	8 schools & 0.15s & 0.6s & 5.5MB \\
+	10D normal, 100 points & 0.15s & 3.0s & 5.7MB \\
+	50D normal, 100 points & 0.15s & 13.0s & 5.8MB 
+\end{tabular}
 \end{table}}
@@
 added to most modern general-purpose programming languages, in
-particular, those used for implementation of large-scale
-software systems, making probabilistic programming inference
-more accessible in large scale server-side applications. We
-described Infergo implementation choices and techniques in
-intent to encourage broader adoption of probabilistic
-programming in production software systems and ease
-implementation of similar facilities in other languages and
-software environments.
+particular, those used for implementation of large-scale software
+systems, making probabilistic programming inference more
+accessible in large scale server-side applications. We described
+Infergo implementation choices and techniques in intent
+to encourage broader adoption of probabilistic programming in
+production software systems and ease implementation of similar
+facilities in other languages and software environments.
 
@@
 
-\clearpage \clearpage
+\clearpage
+\clearpage
 

### 2019-04-20 paper/paper.tex
highlights in progress

@@
 
-Infergo employs reverse-mode autodifferentiation via source code
-transformation~\cite{GW08}. The reverse mode is the default
-choice in machine learning applications, because it works best
-with differentiating a scalar return value with respect to
-multiple parameters. Automatic differentiation is usually
-implemented using either operator overloading or source code
-transformation. Operator overloading is often the method of
+Infergo employs reverse-mode automatic differentiation via
+source code transformation~\cite{GW08}. The reverse mode is the
+default choice in machine learning applications, because it
+works best with differentiating a scalar return value with
+respect to multiple parameters. Automatic differentiation is
+usually implemented using either operator overloading or source
+code transformation. Operator overloading is often the method of
 choice~\cite{Stan17,GXG18,PGC+17} due to relative simplicity,
@@
 numbers. Regardless of the issue of computational overhead, this
-restricts the language used to specify the model to operations on
-that special type and impairs interoperability: library
+restricts the language used to specify the model to operations
+on that special type and impairs interoperability: library
 functions cannot be directly called but have to use an adaptor,
@@
 
-Source code transformation allows to automatically differentiate
-program code operating on the built-in floating point type. This
-method involves generation of new code, for the function
-gradient or for the function itself, ahead of time or on the
-fly. To achieve the best performance, some implementations
-generate static code for computing the function
-gradient~\cite{MMW18,I18}. This requires a rather elaborated
-source code analysis, and although a significant progress has
-been made in this direction, not all language constructs are
-supported. Alternatively, the code of the function itself can be
-transformed by augmenting arithmetics and function calls on the
-floating point type by function calls that record the
-computations on the \textit{tape}, just like in the case of
-operator overloading. Then, the backward pass is identical to
-that of operator overloading. This method allows the
-use of the built-in floating point type along with
+\paragraph{Reverse mode source code transformation} Source code
+transformation allows to automatically differentiate program
+code operating on the built-in floating point type. This method
+involves generation of new code, for the function gradient or
+for the function itself, ahead of time or on the fly. To achieve
+the best performance, some implementations generate static code
+for computing the function gradient~\cite{MMW18,I18}. This
+requires a rather elaborated source code analysis, and although
+a significant progress has been made in this direction, not all
+language constructs are supported. Alternatively, the code of
+the function itself can be transformed by augmenting arithmetics
+and function calls on the floating point type by function calls
+that record the computations on the \textit{tape}, just like in
+the case of operator overloading. Then, the backward pass is
+identical to that of operator overloading. This method allows
+the use of the built-in floating point type along with
 straightforward virtually unrestricted language for model
@@
 differentiated model are available simultaneously in the calling
-code, so that invocations of model methods which do not require
-computation of gradients can be performed without the overhead
-of automatic differentiation.  However, the performance loss is
-rarely a major issue --- differentiated versions of model
-methods are only 2--4 times slower on models from the case
-studies~\ref{sec:case-studies}, and the backward pass, which
+code, so that model methods which do not require computation of
+gradients can be invoked without the overhead of automatic
+differentiation.  However, the performance loss is rarely a
+major issue --- differentiated versions of model methods are
+only 2--4 times slower on models from the case studies
+(Section~\ref{sec:case-studies}), and the backward pass, which
 computes the gradient, is roughly as fast as the original,
 undifferentiated version. At least partially, fast backward pass
-is due to careful implementation and thorough profiling of
-the tape.
+is due to careful implementation and thorough profiling of the
+tape.
+
+\paragraph{Automatic differentiation of models}
+It may seem desirable to implement or use an automatic
+differentiation library that is more general than needed for
+probabilistic programming. However the use of such library, in
+particular, with source code transformation, may make
+implementation restricted and less efficient: first, if a model
+comprises several functions, all functions must have a signature
+suitable for differentiation (that is, accept and return
+floating point scalars or vectors); second, the gradient of
+every nested call must be fully computed first and then used to
+compute the gradient of the caller. In Infergo, only the
+gradient of \lstinline{Observe} is needed, and any other
+differentiated model method can only be called from
+\lstinline{Observe} or another differentiated method.
+Consequently, only the gradient of \lstinline{Observe} is
+computed explicitly, and other differentiated methods may have
+arbitrary arguments. In particular, derivatives may propagate
+through fields of the model object and structured arguments.
+
+\subsection{Model composition}
+
+Just like it is possible to compose library and user-defined
+functions into more complex functions, it should be possible to
+compose models to build other models~\cite{THS+17,SSW+18}.
+Infergo supports model composition organically: a differentiated
+model method can be called from another differentiated method,
+not necessarily a method of the same model object or type.
+Derivatives can propagate through methods of different models
+without restriction. Each model is a building block for another
+model. For example, if there are two independent
+single-parameter models \lstinline{A} and \lstinline{B}
+describing the data, then the composite model \lstinline{AB} can
+be obtained by summation:
+\begin{lstlisting}
+func (model AB) Observe(x []float64) float64 {
+	return A{model.Data}.Observe(x[:1]) + 
+		B{model.Data}.Observe(x[1:])
+}
+\end{lstlisting}
+Combinators such as product, mixture, or hierarchy can be
+implemented as functions operating on Infergo models.
 

### 2019-04-21 paper/paper.tex
model composition

@@
 	benchmarks, comparing Infergo to dedicated inference-centric
-	probabilistic programming systems.
+	probabilistic programming frameworks.
 \end{abstract}
@@
 	variables. Inference algorithms are provided by the
-	probabilistic programming system, and each algorithm is
+	probabilistic programming framework, and each algorithm is
 	usually applicable to a wide class of probabilistic programs
@@
 	is an extension of a regular programming toolbox allowing
-	algorithms implemented in general-purpose programming languages to
-	have learnable parameters. In this view, statistical models
-	are integrated into the software system, and inference and
-	optimization take place during data processing, prediction,
-	and algorithmic decision making in production.
-	Probabilistic programming code runs unattended, and
-	inference results are an integral part of the algorithms.
+	algorithms implemented in general-purpose programming
+	languages to have learnable parameters. In this view,
+	statistical models are integrated into the software system,
+	and inference and optimization take place during data
+	processing, prediction, and algorithmic decision making in
+	production.  Probabilistic programming code runs unattended,
+	and inference results are an integral part of the
+	algorithms.
 
@@
 	programming as a flexible tool for ubiquitous statistical
-	inference. Most probabilistic programming systems are
+	inference. Most probabilistic programming frameworks are
 	suited, to a certain extent, to be used in either the
@@
 	proliferation of probabilistic programming languages and
-	systems\footnote{There are 45 probabilistic programming
+	frameworks \footnote{There are 45 probabilistic programming
 	languages listed on
@@
 	do not fall short from and sometimes exceed those of
-	probabilistic programming-centric systems with custom
+	probabilistic programming-centric frameworks with custom
 	languages and runtimes. Availability of such facilities
@@
 efficient inference in many probabilistic programming
-systems~\cite{Stan17,GXG18,ISF+18,BCJ+19,THS+17}. Different
+frameworks~\cite{Stan17,GXG18,ISF+18,BCJ+19,THS+17}. Different
 automatic differentiation techniques~\cite{GW08} allow 
@@
 particular application, probabilistic programming languages and
-systems are customarily compared by the expressive power
+frameworks are customarily compared by the expressive power
 (classes of probabilistic models that can be represented),
@@
 implementation of the simulation, on the other hand. Some
-probabilistic systems (Figaro~\cite{P09},
+probabilistic programming frameworks (Figaro~\cite{P09},
 Anglican~\cite{TMY+16}, Turing~\cite{GXG18}) are built with
@@
 of probabilistic programming tools. Probabilistic programming
-systems that are implemented or provide an interface in a
+frameworks that are implemented or provide an interface in a
 popular programming language, e.g.  Python
@@
 integrate and deploy, however the smaller the footprint of a
-probabilistic system, the easier is the adoption.
+probabilistic programming framework, the easier is the adoption.
 An obstacle in integration of probabilistic programming lies
@@
 Julia~\cite{Julia14} have been popular choices for development
-of probabilistic programming languages and systems integrated
+of probabilistic programming languages and frameworks integrated
 with general-purpose programming
@@
 single-parameter models \lstinline{A} and \lstinline{B}
-describing the data, then the composite model \lstinline{AB} can
-be obtained by summation:
-\begin{lstlisting}
+describing the data, then the product model \lstinline{AB} can
+be obtained by composition of \lstinline{A} and \lstinline{B}
+(Figure~\ref{fig:composition}).
+\begin{figure}
+	\begin{lstlisting}[framexleftmargin=10pt,numbers=left]
+type A struct {Data []float64}
+func (model A) Observe(x []float64) {
+	...
+}
+
+type B struct {Data []float64}
+func (model B) Observe(x []float64) {
+	...
+}
+
+type AB struct {Data []float64}
 func (model AB) Observe(x []float64) float64 {
@@
 	\end{lstlisting}
+	\caption{Model composition. Model \lstinline{AB} is the
+	product of \lstinline{A} and \lstinline{B}.}
+	\label{fig:composition}
+\end{figure}
 Combinators such as product, mixture, or hierarchy can be
@@
 
-% Elementals. Differentiation
-%of calls. Distributions. Composability. Nested inference.
+\paragraph{Distributions} A probabilistic programming facility
+should provide a library of distributions for defining and
+conditional random variables. To use a distribution in a
+differentiable model, the distribution's $\log$ probability
+density function must be differentiable by both the
+distribution's parameters and by the value of the random
+variable. Distributions are usually supplied as a part of the
+probabilistic programming framework~\cite{TMY+16,Stan17,GXG18},
+and, if at all possible, adding a user-defined distribution
+requires programming at a lower level of abstraction than while
+specifying a probabilistic model~\cite{S16,TMY+16,AMP17}.
+
+In Infergo, distributions are just models, and adding and using
+a custom distribution is indistinguishable from adding a model
+and calling the model's method from another model. By
+convention, library distributions also have methods
+\lstinline{Logp} and \lstinline{Logps} for the scalar and vector
+version of $\log$ probability density functions, for convenience
+of use in models. The distribution library in Infergo is just a
+model package which is automatically differentiated just like
+any other model. New distributions can be added in any package.
+Figure~\ref{fig:exponential-distribution} shows the definition of the
+exponential distribution.
+\begin{figure}
+	\begin{lstlisting}[framexleftmargin=10pt,numbers=left]
+// Exponential distribution.
+type expon struct{}
+
+// Exponential distribution, singleton instance.
+var Expon expon
+
+// Observe implements the Model interface.The
+// parameter vector is lambda, observations.
+func (dist expon) Observe(x []float64)
+	float64 {
+	lambda, y := x[0], x[1:]
+	if len(y) == 1 {
+		return dist.Logp(lambda, y[0])
+	} else {
+		return dist.Logps(lambda, y...)
+	}
+}
+
+// Logp computes the log pdf of a single
+// observation.
+func (_ expon) Logp(lambda float64, y float64)
+	float64 {
+	logl := math.Log(lambda)
+	return logl - lambda*y
+}
+
+// Logps computes the log pdf of a vector
+// of observations.
+func (_ expon) Logps(lambda float64, y ...float64)
+	float64 {
+	ll := 0.
+	logl := math.Log(lambda)
+	for i := range y {
+		ll += logl - lambda*y[i]
+	}
+	return ll
+}
+	\end{lstlisting}
+	\caption{Exponential distribution in Infergo.}
+	\label{fig:exponential-distribution}
+\end{figure}
 

### 2019-04-21 paper/paper.tex
model composition

@@
 \paragraph{Distributions} A probabilistic programming facility
-should provide a library of distributions for defining and
-conditional random variables. To use a distribution in a
+should provide a library of distributions for definition and
+conditioning of random variables. To use a distribution in a
 differentiable model, the distribution's $\log$ probability

### 2019-04-21 paper/paper.tex
stochasticity and streaming

@@
 \subsection{A Basic Example}
+\label{sec:basic-example}
 
@@
 
+Perhaps somewhat surprisingly, most probabilistic programs are
+semantically deterministic~\cite{SYW+16} --- probabilistic programs
+define distributions, and although they manipulate random
+variables, there is no randomization or non-determinism
+\textit{per se}. Randomization arises in some inference
+algorithms (notably of the Monte Carlo family) for tractable
+approximate inference, but this does not introduce
+non-determinism into the programs themselves. Non-deterministic
+probabilistic programs arise in applications. For example,
+\citet{MPT+16} explored using probabilistic programming for
+policy search in non-deterministic domains. To represent
+non-determinism, appropriate constructs had to be added to the
+probabilistic programming language used in the study.
+
+In Infergo, non-determinism can be easily introduced through
+data. In the basic example if Section~\ref{sec:basic-example},
+the data is encapsulated in the model as a vector, populated
+before inference. However, this is by no means required by
+Infergo.  Instead of being stored in a Go slice, the data can be
+read from a Go channel or network connection and generated by
+another goroutine or process.
+Figure~\ref{fig:basic-example-channel} shows a modified model
+from the basic example with the data read from a channel rather.
+\begin{figure}
+	\begin{lstlisting}[framexleftmargin=10pt,numbers=left]
+type Model struct {
+  Data chan float64  // data is a channel
+  N    int           // batch size
+}
+
+func (m *Model) Observe(x []float64) float64 {
+  ll := Normal.Logps(0, 1, x...)
+  // observe a batch of data from the channel
+  for i := 0; i != m.N; i++ {
+    ll += Normal.Logp(x[0], math.Exp(x[1]),
+		<- m.Data)
+  }
+  return ll
+}
+	\end{lstlisting}
+	\caption{The basic example with the data read from a channel.}
+	\label{fig:basic-example-channel}
+\end{figure}
+
+A similar approach can be applied to streaming data: the model
+may read the data gradually as it arrives, and the inference
+algorithm will emit samples from the time process inferred from
+the data. Work is still underway on algorithms ensuring robust
+and sound inference on streams, but Infergo models connected to
+data sources via channels are already used in practice.
+
 \section{Case Studies}

### 2019-04-21 paper/paper.tex
case studies in progress

@@
 
-%\linespread{1.0}
+\linespread{1.1}
 
@@
 
-// Observe implements the Model interface.The
+// Observe implements the Model interface. The
 // parameter vector is lambda, observations.
@@
 	\end{lstlisting}
-	\caption{Exponential distribution in Infergo.}
+	\caption{The exponential distribution in Infergo.}
 	\label{fig:exponential-distribution}
@@
 
-\subsection{8 schools}
+In this section, we present three case studies of probabilistic
+programs, each addressing a different aspect of Infergo. Eight
+schools (Section~\ref{sec:8-schools}) is a popular example  we
+use to compare model specification in Stan and Infergo. Linear
+regression (Session~\ref{sec:8-schools}) formulated as a
+probabilistic program lets us explore a model comprising
+multiple methods and illustrate the inference-simulation use
+case. Gaussian mixture (Section~\ref{sec:gaussian-mixture}) is
+the most complex model of the three, featuring nested loops and
+dynamic control structure.
+
+By no means the models presented in this section reflect 
+complexity or the size of real-world Infergo applications, the
+latter reaching hundreds of lines of code in length and
+performing inference on tens of thousands of data entries.
+Rather, their purpose is to provide a general picture of 
+probabilistic programming with Infergo.
+
+\subsection{Eight schools}
+\label{sec:8-schools}
+
+The eight schools problem~\cite{GCS+03}
 
@@
 	\end{minipage}
-	\caption{8 schools: Stan vs. Infergo. The Go implementation
+	\caption{Eight schools: Stan vs. Infergo. The Go implementation
 	has a similar length and structure to the Stan model.}
@@
 \subsection{Linear regression}
+\label{sec:linear-regression}
 
@@
 \subsection{Gaussian mixture}
+\label{sec:gaussian-mixture}
 

### 2019-04-21 paper/paper.tex
consistent indents
linear regression

@@
 \section{Challenges}
+\label{sec:challenges}
 
@@
 
-\paragraph{Simulation vs. inference} Probabilistic models
-often follow a design pattern of simulation-inference: a
-significant part of the model is a simulator, running an
-algorithm with fixed parameters; the optimal parameters, or
-their distribution, are to be inferred. The inferred parameters
-are then used by the software system to execute the simulation
-independently of inference for forecasting and decision making.
+\paragraph{Simulation vs. inference} Probabilistic models often
+follow the design pattern of simulation-inference: a significant
+part of the model is a simulator, running an algorithm with
+given parameters; the optimal parameters, or their distribution,
+are to be inferred. The inferred parameters are then used by the
+software system to execute the simulation independently of
+inference for prediction and decision making.
 
@@
 the data is encapsulated in the model as a vector, populated
-before inference. However, this is by no means required by
+before inference. However, this is not at all required by
 Infergo.  Instead of being stored in a Go slice, the data can be
@@
 programs, each addressing a different aspect of Infergo. Eight
-schools (Section~\ref{sec:8-schools}) is a popular example  we
+schools (Section~\ref{sec:eight-schools}) is a popular example  we
 use to compare model specification in Stan and Infergo. Linear
-regression (Session~\ref{sec:8-schools}) formulated as a
+regression (Session~\ref{sec:eight-schools}) formulated as a
 probabilistic program lets us explore a model comprising
-multiple methods and illustrate the inference-simulation use
-case. Gaussian mixture (Section~\ref{sec:gaussian-mixture}) is
+multiple methods and illustrate the simulation-inference pattern.
+Gaussian mixture (Section~\ref{sec:gaussian-mixture}) is
 the most complex model of the three, featuring nested loops and
@@
 
-By no means the models presented in this section reflect 
+By no means the models presented in this section reflect the
 complexity or the size of real-world Infergo applications, the
@@
 \subsection{Eight schools}
-\label{sec:8-schools}
-
-The eight schools problem~\cite{GCS+03}
-
+\label{sec:eight-schools}
+
+The eight schools problem~\cite{GCS+03} is an application of
+the hierarchical normal model that often serves as an introductory
+example of Bayesian statistical modeling. Without going into
+details, in this problem the effect of coaching programs on test
+outcomes is compared among eight schools.
+Figure~\ref{fig:eight-schools} shows specifications of the same
+model in Stan and Infergo, side-by-side.
 \begin{figure*}
@@
   has a similar length and structure to the Stan model.}
-	\label{fig:8-schools}
+  \label{fig:eight-schools}
 \end{figure*}
+
+Stan models are written in a probabilistic programming language
+tailored to statistical modeling. Infergo models are implemented
+in Go. One can see that the models have roughly the same length
+and structure. The definition of the model type in Go (lines
+1--5) corresponds to the \lstinline{data} section in Stan (lines
+1--5). Destructuring of the parameters of \lstinline{Observe} in
+Go (lines 9--11) names the parameters just like the
+\lstinline{parameters} section in Stan (lines 8--10).
+Introduction of \lstinline{theta} in Go (line 16) corresponds to
+the \lstinline{transformed parameters} section in Stan (lines
+13--16). The only essential difference is the explicit loop over
+the data in Go (lines 15--18) versus vectorized computations in
+Stan (lines 15 and 20), however vectorized computations are used
+in Stan to speed-up execution more than to improve readability;
+if repeated calls to \lstinline{Normal.Logp} (line 17) become
+the bottleneck, an optimized vectorized version can be
+implemented in Go in the same package. While subjective, our
+feeling is that the Go version is as readable as the Stan one.
+In addition, a Go programmer is less likely to resist adoption
+of statistical modeling if the models are implemented in Go.
 
@@
 
+The model in Figure~\ref{fig:linear-regression} specifies linear
+regression on unidimensional data: an improper uniform prior is
+placed on the parameters $\alpha$, $\beta$, and $\log \sigma$
+(lines 9--10), which are then conditioned on observations $x, y$
+such that
+\begin{equation}
+y \sim \mathrm{Normal}(\alpha + \beta x, \sigma).
+	\label{eqn:linear-regression-1}
+\end{equation}
 \begin{figure}
@@
       m.Simulate(m.Data[i][0], alpha, beta),
-      sigma, m.Data[i][1])
+	  sigma,
+	  m.Data[i][1])
   }
   return ll
-
-// Simulate predicts y for x based
-// on inferred parameters.
+}
+
+// Simulate simulates y based on x and 
+// regression parameters.
 func (m *Model) Simulate(x, alpha, beta float64)
     float64 {
-  y := alpha +beta*x
+  y := alpha + beta*x
   return y
@@
 \end{figure}
+If only the maximum \textit{a posteriori} is estimated, the
+parameter values will maximize their likelihood given the
+observations:
+\begin{equation}
+	(\alpha, \beta, \sigma)_{MAP} = \arg \max \prod\limits_i \mathrm{Normal}(\alpha + \beta x_i, \sigma)
+	\label{eqn:linear-regression-2}
+\end{equation}
+
+Once the values of $\alpha$ and $\beta$ are estimated, they are
+used to predict $y$ for unseen values of $x$. Hence, the linear
+transformation $\alpha + \beta x$ is computed in two different
+contexts: first, to compute the likelihood of the parameter
+values; then, to predict $y$ based on a given $x$. This is an
+example of the simulation-inference pattern
+(Section~\ref{sec:challenges}): method \lstinline{Simulate}
+simulates $y$ based on $x$ and is invoked both from
+\lstinline{Observe} and then directly for prediction.  Since
+both original and differentiated model are simultaneously
+available in the source code, $Simulate$ can be called
+for prediction on the original model, without any
+differentiation overhead.
+
+In this greatly simplified case study, the simulator is a
+one-line function. In a real-world scenario though, the
+simulator can be laborious to re-implement correctly and
+efficiently; for example, a simulator that executes a
+parameterized policy in an uncertain or dynamic domain, and
+depends on both parameter values and observations~\cite{MPT+16}.
+The ability to re-use a single implementation both as a part of
+the probabilistic program, where the simulator must be augmented
+for inference (e.g. by automatic differentiation), and for
+prediction or decision making, where the simulator may be
+executed repeatedly and under time pressure, saves the effort
+otherwise wasted on rewriting the simulator and removes the need
+to maintain consistency between different implementations.
 
@@
 
-{\smaller
-\begin{table}[H]
-\caption{Memory and running times for 1000 iterations
-of HMC with 10 leapfrog steps.}
+\begin{table*}
+\begin{tabular}{r | c |  c | c | c | c | c |}
+	{\it model}  & \multicolumn{4}{c|}{\it time} & \multicolumn{2}{c|}{} \\
+	& \multicolumn{2}{c|}{\it compilation} & \multicolumn{2}{c|}{\it execution} & \multicolumn{2}{c|}{\it memory} \\
+	& Infergo & Stan & Infergo & Stan & Infergo & Stan \\\hline
+	8 schools & 0.15s & 10s & 0.6s & 0.3s & 5.5MB & 16MB \\
+	10D normal, 100 points & 0.15s & 10s & 3.0s & 2.0s & 5.7MB & 20 MB \\
+	50D normal, 100 points & 0.15s & 10s & 13.0s & 4.0s & 5.8MB & 22 MB  
+\end{tabular}
+\caption{Memory and running times for 1000 iterations of HMC with 10 leapfrog steps.}
 \label{tab:memory-runtime}
-\begin{tabular}{r | c |  c | c}
-	{\it model}  & \multicolumn{2}{c|}{\it time} & {\it memory} \\ 
-	 & {\it compilation} & {\it execution} & \\\hline
-	8 schools & 0.15s & 0.6s & 5.5MB \\
-	10D normal, 100 points & 0.15s & 3.0s & 5.7MB \\
-	50D normal, 100 points & 0.15s & 13.0s & 5.8MB 
-\end{tabular}
-\end{table}}
+\end{table*}
 
@@
 
-\clearpage
-\clearpage
-
 \bibliography{refs}

### 2019-04-21 paper/paper.tex
in progress

@@
 
-\linespread{1.1}
+%\linespread{1.1}
 
@@
 
-\subsection{Model composition}
+\subsection{Model Composition}
 
@@
 
-\subsection{Eight schools}
+\subsection{Eight Schools}
 \label{sec:eight-schools}
@@
 
-\subsection{Linear regression}
+\subsection{Linear Regression}
 \label{sec:linear-regression}
@@
 
-\subsection{Gaussian mixture}
+\subsection{Gaussian Mixture Model}
 \label{sec:gaussian-mixture}
@@
   ll := 0.0
+
+  // Impose a prior on component parameters
+  ll += Normal.Logps(0, 1, x...)
+
+  // Fetch the parameters
   mu := make([]float64, m.NComp)
   sigma := make([]float64, m.NComp)
-
-  // Fetch component parameters
   for j := 0; j != m.NComp; j++ {
@@
 
-  // Compute log likelihood of mixture
+  // Compute log likelihood of the mixture
   // given the data
@@
       } else {
-        l = mathx.LogSumExp(l, lj)
+        l = logSumExp(l, lj)
       }
@@
 }
+
+// logSumExp computes log(exp(x)+exp(y)) robustly.
+func LogSumExp(x, y float64) float64 {
+  z := x
+  if y > z {
+    z = y
+  }
+  return z + math.Log(math.Exp(x-z)+math.Exp(y-z))
+}
+
+// logSumExp is an elemental, we must register
+// the gradient.
+func init() {
+  ad.RegisterElemental(logSumExp,
+    func(_ float64, params ...float64)
+      []float64 {
+      z := math.Exp(params[1] - params[0])
+      t := 1 / (1 + z)
+      return []float64{t, t * z}
+    })
+}
 \end{lstlisting}
-\caption{Gaussian mixture: Infergo can handle models with
-  dynamic control structure.}
+\caption{Gaussian mixture model: an Infergo model with
+  dynamic control structure and a user-defined elemental.}
 \label{fig:gaussian-mixture}

### 2019-04-22 paper/paper.tex
polishing

@@
 
-%\linespread{1.1}
+\linespread{1.1}
 
@@
 instead of using explicit calls or directives to denote parts of
-code which needs to be differentiated, we rely on the type
+code which need to be differentiated, we rely on the type
 system of Go to selectively differentiate the code relevant for
@@
 \textit{a posteriori} (MAP) estimation, which can be done using
-a gradient descent algorithm (Adam~\cite{KB15} in this case).
+a gradient descent algorithm (Adam~\cite{KB15} in this case):
 
@@
 statements, function and method calls, and recursion. Case
-studies~(Section~\ref{sec:case-studies}) provide elaborated
+studies~(Section~\ref{sec:case-studies}) provide more
 model examples.
@@
 the use of the built-in floating point type along with
-straightforward virtually unrestricted language for model
-specification, at the cost of slight performance loss.
+virtually unrestricted language for model specification, at the
+cost of slight performance loss.
 
@@
 Combinators such as product, mixture, or hierarchy can be
-implemented as functions operating on Infergo models.
+realized as functions operating on Infergo models.
 
@@
 and calling the model's method from another model. By
-convention, library distributions also have methods
-\lstinline{Logp} and \lstinline{Logps} for the scalar and vector
-version of $\log$ probability density functions, for convenience
-of use in models. The distribution library in Infergo is just a
-model package which is automatically differentiated just like
-any other model. New distributions can be added in any package.
-Figure~\ref{fig:exponential-distribution} shows the definition of the
-exponential distribution.
+convention, library distributions, in addition to
+\lstinline{Observe}, also have methods \lstinline{Logp} and
+\lstinline{Logps} for the scalar and vector version of $\log$
+probability density functions, for convenience of use in models.
+The distribution library in Infergo is just a model package
+which is automatically differentiated like any other model.  New
+distributions can be added in any package.
+Figure~\ref{fig:exponential-distribution} shows the Infergo
+definition of the exponential distribution.
 \begin{figure}
@@
 variables, there is no randomization or non-determinism
-\textit{per se}. Randomization arises in some inference
+\textit{per se}. Randomization arises in certain inference
 algorithms (notably of the Monte Carlo family) for tractable
@@
 Figure~\ref{fig:basic-example-channel} shows a modified model
-from the basic example with the data read from a channel rather.
+from the basic example with the data gradually read from a
+channel rather than passed as a whole in a field of the model
+object.
 \begin{figure}
@@
 the data in Go (lines 15--18) versus vectorized computations in
-Stan (lines 15 and 20), however vectorized computations are used
-in Stan to speed-up execution more than to improve readability;
-if repeated calls to \lstinline{Normal.Logp} (line 17) become
+Stan (lines 15 and 20). However, vectorized computations are used
+in Stan to speed-up execution more than to improve readability.
+If repeated calls to \lstinline{Normal.Logp} (line 17) become
 the bottleneck, an optimized vectorized version can be
@@
 \begin{equation}
-y \sim \mathrm{Normal}(\alpha + \beta x, \sigma).
+y \sim \mathrm{Normal}(\alpha + \beta x, \sigma)
 	\label{eqn:linear-regression-1}
 \end{equation}
+(lines 12--17).
 \begin{figure}
@@
 
+The Gaussian mixture model is a probabilistic model of a mixture
+of components, where each component comes from a Gaussian
+distribution. The model variant in
+Figure~\ref{fig:gaussian-mixture} accepts a vector of
+single-dimensional observations and the number of components,
+and infers the distribution of parameters of each component
+in the mixture. Following Stan User's Guide~\cite{SDT18},
+component memberships are summed out to make the model
+differentiable.
 \begin{figure}
@@
 type Model struct {
-  Data  []float64 // samples
+  Data  []float64 // observations
   NComp int       // number of components
@@
   ll := 0.0
-
   // Impose a prior on component parameters
@@
   // Compute log likelihood of the mixture
-  // given the data
   for i := 0; i != len(m.Data); i++ {
@@
 
-// logSumExp is an elemental, we must register
-// the gradient.
+// logSumExp gradient must be supplied.
 func init() {
@@
 \end{figure}
+The model contains most of the steps commonly found in Infergo
+models. First, a prior on the model parameters is imposed (lines
+8--9). Then, the parameter vector is destructured into a form
+convenient for formulation of conditioning (lines 11--17).
+Finally, the parameters are conditioned on the observations
+(lines 19--34).
+
+The conditioning of parameters involves a nested loop over all
+observations and over all mixture components for each
+observation.  Computing log-likelihood of the model parameters
+implies summing up likelihoods over components for each
+observation.  In the log-domain, a trick called `log-sum-exp' is
+required to avoid loss of precision. The trick is implemented
+as function \lstinline{logSumExp} and called for each
+observation and component (line 28).
+
+\lstinline{logSumExp} could have been implemented as a model
+method and differentiated automatically. However, this function
+is called on each iteration of the inner loop. Hence,
+efficiency of inference in general is affected by 
+an efficient implementation of \lstinline{logSumExp} (this can
+also be confirmed by profiling). To save on the cost of a
+differentiated call and speed up gradient computation,
+\lstinline{logSumExp} is implemented as an elemental (lines
+36--43). The hand-coded gradient is supplied for
+\lstinline{logSumExp} (lines 45--54) to make automatic
+differentiation work.
+
+While still a very simple model compared to real-world Infergo
+models, this case study illustrates common parts and coding
+patterns of an elaborated model. The model code may be quite
+complicated algorithmically and manipulate Go data structures
+freely. When efficiency becomes an issue, parts of the model
+can be implemented at a lower level of abstraction, but
+nonetheless in the same programming language as the rest of the
+model.
 
 \section{Performance}
-
-Table~\ref{tab:memory-runtime} provides memory consumption and
-running time measurements on basic models to illustrate
-Infergo's performance.  The measurements were obtained on a
-2.3GHz Intel Core 5 CPU with 8GB of memory for 1000 iterations
-of Hamiltonian Monte Carlo with 10 leapfrog steps. Note that
-log-likelihood computation for standard distributions is not
-optimized yet. Quite the opposite: since models in Infergo are
-fully composable, primitive distributions are themselves
-implemented as Infergo models and automatically differentiated.
 
@@
 
+Table~\ref{tab:memory-runtime} provides memory consumption and
+running time measurements on basic models to illustrate
+Infergo's performance.  The measurements were obtained on a
+2.3GHz Intel Core 5 CPU with 8GB of memory for 1000 iterations
+of Hamiltonian Monte Carlo with 10 leapfrog steps. Note that
+log-likelihood computation for standard distributions is not
+optimized yet. Quite the opposite: since models in Infergo are
+fully composable, primitive distributions are themselves
+implemented as Infergo models and automatically differentiated.
+
 \section{Conclusion}

### 2019-04-23 paper/paper.tex
work in progress

@@
 \begin{table*}
-\begin{tabular}{r | c |  c | c | c | c | c |}
-	{\it model}  & \multicolumn{4}{c|}{\it time} & \multicolumn{2}{c|}{} \\
-	& \multicolumn{2}{c|}{\it compilation} & \multicolumn{2}{c|}{\it execution} & \multicolumn{2}{c|}{\it memory} \\
-	& Infergo & Stan & Infergo & Stan & Infergo & Stan \\\hline
-	8 schools & 0.15s & 10s & 0.6s & 0.3s & 5.5MB & 16MB \\
-	10D normal, 100 points & 0.15s & 10s & 3.0s & 2.0s & 5.7MB & 20 MB \\
-	50D normal, 100 points & 0.15s & 10s & 13.0s & 4.0s & 5.8MB & 22 MB  
+\begin{tabular}{r | c |  c | c | c | c | c | }
+	{\it model}  & \multicolumn{6}{c|}{\it time} \\
+	& \multicolumn{3}{c|}{\it compilation} & \multicolumn{3}{c|}{\it execution} \\
+	& Infergo & Turing & Stan & Infergo & Turing & Stan \\\hline
+	8 schools              & 0.5s & & & 0.6s  & & \\
+	10D normal, 100 points & 0.5s & & & 3.0s  & & \\
+	50D normal, 100 points & 0.5s & & & 13.0s & & \\
+	LDA, 262 words, 25 docs
+	                       & 0.6s & & &       & & 
 \end{tabular}
-\caption{Memory and running times for 1000 iterations of HMC with 10 leapfrog steps.}
+\caption{Compilation and execution times for 1000 iterations 
+	of HMC with 10 leapfrog steps.}
 \label{tab:memory-runtime}

### 2019-04-23 paper/paper.tex
stan times

@@
 	\begin{tabular}{r | c |  c | c | c | c | c | }
-	{\it model}  & \multicolumn{6}{c|}{\it time} \\
+		{\it model}   & \multicolumn{6}{c|}{\it time, seconds} \\
 					  & \multicolumn{3}{c|}{\it compilation} & \multicolumn{3}{c|}{\it execution} \\
 					  & Infergo & Turing & Stan              & Infergo & Turing & Stan    \\\hline
-	8 schools              & 0.5s & & & 0.6s  & & \\
-	10D normal, 100 points & 0.5s & & & 3.0s  & & \\
-	50D normal, 100 points & 0.5s & & & 13.0s & & \\
-	LDA, 262 words, 25 docs
-	                       & 0.6s & & &       & & 
+		Eight schools & 0.5     &        & 50                & 0.6     &        & 0.12    \\
+		GMM           & 0.5     &        & 55                & 3.0     &        & 4.9     \\
+		LDA           & 0.5     &        & 55                &         &        & 3.7     \\ 
 	\end{tabular}

### 2019-04-23 paper/paper.tex
empirical

@@
 
-\linespread{1.1}
+\linespread{1.07}
 
@@
 implementation of a facility best suited for probabilistic
-inference in a porduction software system, we were looking
+inference in a production software system, we were looking
 for a language that \begin{itemize}
@@
 					        & Infergo & Turing & Stan              & Infergo  & Turing & Stan    \\\hline
-		Eight schools & 0.5     &        & 50                & 0.6     &        & 0.12    \\
-		GMM           & 0.5     &        & 55                & 3.0     &        & 4.9     \\
-		LDA           & 0.5     &        & 55                &         &        & 3.7     \\ 
+              Eight schools & 0.50    & -      & 50                & 0.60     & 2.8    & 0.12    \\
+     Gaussian mixture model & 0.50    & -      & 54                & 32       & 14     & 4.9     \\
+Latent Dirichlet allocation & 0.50    & -      & 54                & 8.9      & 12     & 3.7     \\ 
 	\end{tabular}
@@
 \end{table*}
-
-Table~\ref{tab:memory-runtime} provides memory consumption and
-running time measurements on basic models to illustrate
-Infergo's performance.  The measurements were obtained on a
-2.3GHz Intel Core 5 CPU with 8GB of memory for 1000 iterations
-of Hamiltonian Monte Carlo with 10 leapfrog steps. Note that
-log-likelihood computation for standard distributions is not
-optimized yet. Quite the opposite: since models in Infergo are
-fully composable, primitive distributions are themselves
-implemented as Infergo models and automatically differentiated.
+It is customary to include performance evaluation on benchmark
+problems in publications on new probabilistic programming
+frameworks~\cite{PW14,WVM14,TMY+16,GXG18}. We provide here a
+comparison of running times of Infergo and two other probabilistic
+programming frameworks: Turing~\cite{GXG18} and
+Stan~\cite{Stan17}. Turing is a Julia~\cite{Julia14} library
+and DSL offering exploratory statistical modeling  and
+inference composition. Stan is a compiled language and a library
+optimized for high performance.
+
+We use three models of different size and structure for the
+comparison: two of the models, Eight schools and Gaussian
+mixture model, were introduced as a part of the case
+studies~(Section~\ref{sec:case-studies}); the code for the third
+model, Latent Dirichlet allocation~\cite{BNG03}, is included in
+Appendix~\ref{app:lda}. The code and data for all models were
+originally included in Stan examples and translated for use
+with Turing and Infergo. \textit{Eight schools} is the smallest model.
+One may expect that most of the time execution time is spent in
+executing the boilerplate code rather than the model. \textit{Gaussian
+mixture model} is a more elaborated model which benefits greatly
+from vectorization. The greater the number of observations (1000
+data points were used in the evaluation), the more significant
+is the benefit of vectorized computations. \textit{Latent
+Dirichlet allocation} is, again, a relatively elaborated model,
+that is, however, not as easy to vectorize.
+
+We report both compilation time (for Stan and Infergo) and
+execution time. Compilation time is an important characteristic
+of a probabilistic programming framework --- statistical
+modeling often involves many cycles of model modifications and
+numerical experiments on subsampled data. The ability to modify
+and instantly re-run a model is crucial for exploratory data
+analysis and model design. Turing is based on Julia, which is
+a dynamic language with JIT compiler, and there is no separate
+compilation time. For Stan, the compilation time includes
+compiling the Stan model to C++, and then compiling the C++
+source into executable code for inference. For Infergo, the
+compilation time consists of automatic differentiation of the
+model, and then building a command-line executable that calls
+an inference algorithm on the model.
+
+Table~\ref{tab:memory-runtime} shows running time measurements
+on the models.  The measurements were obtained on a 1.25GHz
+Intel(R) Core(TM) i5 CPU with 8GB of memory for 1000 iterations
+of Hamiltonian Monte Carlo with 10 leapfrog steps and averaged
+over 10 runs.  In execution, Infergo is slower than Stan, which
+comes as little surprise --- Stan is a mature highly optimized
+inference and optimization library. However, Infergo and Turing
+show similar execution times, despite Turing relying on Julia's
+numerical libraries. The slowest relative execution time is on
+the Gaussian mixture model, where Stan is 6--7 times faster,
+apparently because Infergo model is not vectorized; the model
+could be made more efficient at the cost of verbosity and poorer
+readability, but we opted to use idiomatic Infergo models for
+the comparison. On the Latent Dirichlet allocation model, run on
+Stan example data with 25 documents and 262 word instances,
+Infergo is only 2.5 times slower than Stan, and faster than
+Turing, an illustration of the benefits of transformation-based
+differentiation and efficiency of Go.
+
+Infergo programs compile very fast, in particular in comparison
+to Stan models. Both automatic differentiation and code
+generation combined take a fraction of a second. Compilation time
+of Stan models is dominated by compilation of the generated C++
+code which takes close to a minute on our hardware. The ability
+to modify a model and then re-run inference almost instantly,
+and with reasonable performance, helps in exploratory model
+development with Infergo, which was conceived with server-side,
+unattended execution in mind, but fits the exploration and
+development stage just as well.
 
@@
 challenges, we proposed guidelines for implementation of a
-deployable probabilistic programming facility. Further on, we
-introduced Infergo, a probabilistic programming facility in Go,
-as a reference implementation in accordance with the guidelines.
-Infergo allows to program inference models in virtually
-unrestricted Go, and to perform inference on the model using
-either supplied or third-party algorithms. We demonstrated on
-case studies how Infergo overcomes the challenges by following
-the guidelines.
+deployable probabilistic programming facility. Based on the
+guidelines, we introduced Infergo, a probabilistic programming
+facility in Go, as a reference implementation in accordance with
+the guidelines.  Infergo allows to program inference models in
+virtually unrestricted Go, and to perform inference on the model
+using either supplied or third-party algorithms. We demonstrated
+on case studies how Infergo overcomes the challenges by
+following the guidelines.
 
@@
 
+\clearpage
+
+\appendix
+
+\begin{figure*}
+\section{Latent Dirichlet Allocation Model}
+	\label{app:lda}
+\vspace{\baselineskip}
+
+\begin{minipage}{0.49\textwidth}
+\centering
+\textbf{Infergo}
+
+\begin{lstlisting}[framexleftmargin=10pt,numbers=left]
+type Model struct {
+  K     int       // num topics
+  V     int       // num words
+  M     int       // num docs
+  N     int       // total word instances
+  Word  []int     // word n
+  Doc   []int     // doc for word n
+  Alpha []float64 // topic prior
+  Beta  []float64 // word prior
+}
+
+func (m *Model) Observe(x []float64) float64 {
+  ll := 0.0
+
+  // Destructure parameters
+  theta := make([][]float64, m.M)
+  m.simplices(&x, m.K, theta)
+  phi := make([][]float64, m.K)
+  m.simplices(&x, m.V, phi)
+
+  // Impose priors
+  ll += Dirichlet{m.K}.Logps(m.Alpha, theta...)
+  ll += Dirichlet{m.V}.Logps(m.Beta, phi...)
+
+  // Conditioning on observations
+  gamma := make([]float64, m.K)
+  for in := 0; in != m.N; in++ {
+    for ik := 0; ik != m.K; ik++ {
+      gamma[ik] = math.Log(theta[m.Doc[in]-1][ik]) +
+        math.Log(phi[ik][m.Word[in]-1])
+    }
+    ll += D.LogSumExp(gamma)
+  }
+  return ll
+}
+\end{lstlisting}
+\end{minipage}
+\hfill
+\begin{minipage}{0.49\textwidth}
+	\centering
+
+	\textbf{Stan}
+
+\begin{lstlisting}[language=Stan]
+data {
+  int<lower=2> K;               // num topics
+  int<lower=2> V;               // num words
+  int<lower=1> M;               // num docs
+  int<lower=1> N;           // total word instances
+  int<lower=1,upper=V> w[N];    // word n
+  int<lower=1,upper=M> doc[N];  // doc for word n
+  vector<lower=0>[K] alpha;     // topic prior
+  vector<lower=0>[V] beta;      // word prior
+}
+
+parameters {
+  simplex[K] theta[M]; // topic dist for doc m
+  simplex[V] phi[K];   // word dist for topic k
+}
+
+model {
+  for (m in 1:M)  
+    theta[m] ~ dirichlet(alpha);  // prior
+  for (k in 1:K)  
+    phi[k] ~ dirichlet(beta);     // prior
+
+  for (n in 1:N) {
+    real gamma[K];
+    for (k in 1:K) 
+      gamma[k] <- log(theta[doc[n],k]) 
+	  	+ log(phi[k,w[n]]);
+    increment_log_prob(log_sum_exp(gamma));
+  }
+}
+\end{lstlisting}
+	\vspace{2\baselineskip}
+\end{minipage}
+\end{figure*}
+
 \end{document}

### 2019-04-27 paper/paper.tex
language typo fixed

@@
 use to compare model specification in Stan and Infergo. Linear
-regression (Session~\ref{sec:eight-schools}) formulated as a
+regression (Session~\ref{sec:linear-regression}) formulated as a
 probabilistic program lets us explore a model comprising
@@
 
-By no means the models presented in this section reflect the
+By no means do the models presented in this section reflect the
 complexity or the size of real-world Infergo applications, the
@@
 
-  // Conditioning on observations
+  // Condition on observations
   gamma := make([]float64, m.K)

### 2019-04-29 paper/paper.tex
corrections in progress

@@
 
-\linespread{1.07}
+\linespread{1.05}
 
@@
   and inference results are an integral part of the
-  algorithms.
+  algorithms. Natural applications arise whenever the algorithm
+  is a generative model of a mental or physical process, in
+  particular when latent variables of the algorithm are used for
+  decision making~\cite{GT+16,WBD+19}; for example, inference
+  about user behavior in social networks, ongoing health
+  monitoring, or online motion planning of autonomous robotic
+  devices.
 
@@
 
+Infergo comprises a command-line tool that augments the source
+code of a model to facilitate inference, a collection of basic
+models (distributions) serving as building blocks for other
+models, and a library of inference algorithms.
+
 A probabilistic model in Infergo is an implementation of an
-interface requiring a single method \lstinline{Observe} which
-accepts a vector (a Go \textit{slice}) of floats, the parameters
-to infer, and returns a single float, interpreted as
-unnormalized log-likelihood of the posterior distribution.
+interface requiring a single method
+\lstinline{Observe}\footnote{The method name follows
+Venture~\cite{MSP14}, Anglican~\cite{TMY+16}, and other
+probabilistic programming languages in which \textit{observe} is
+the language construct for conditioning of the model on
+observations.} accepts a vector (a Go \textit{slice}) of floats,
+the parameters to infer, and returns a single float, interpreted
+as unnormalized log-likelihood of the posterior distribution.
 Implementation of model methods can be written in virtually
@@
 For inference, Infergo relies on automatic differentiation. The
-source code of the model is translated by a command-line tool
-provided by Infergo into an equivalent model with reverse-mode
-automatic differentiation of the log-likelihood with respect to
-the parameters applied. The differentiation operates on the
-built-in floating-point type and incurs only a small
-computational overhead. However, even this overhead is avoided
-when the model code is executed outside of inference algorithms:
-both the original and the differentiated model are
-simultaneously available to the rest of the program code, so the
-methods can be called on the differentiated model for inference,
-and on the original model for the most efficient execution with
-the inferred parameters.
+source code of the model is translated by the command-line tool
+into an equivalent model with reverse-mode automatic
+differentiation of the log-likelihood with respect to the
+parameters applied. The differentiation operates on the built-in
+floating-point type and incurs only a small computational
+overhead. However, even this overhead is avoided when the model
+code is executed outside of inference algorithms: both the
+original and the differentiated model are simultaneously
+available to the rest of the program code, so the methods can be
+called on the differentiated model for inference, and on the
+original model for the most efficient execution with the
+inferred parameters.
 
@@
 \begin{lstlisting}
-type Model struct {
+type ExampleModel struct {
   Data []float64
@@
 // x[1] is the log stddev of the distribution
-func (m *Model) Observe(x []float64) float64 {
+func (m *ExampleModel)
+     Observe(x []float64) float64 {
   // Our prior is a unit normal ...
@@
 // Data
-m := &Model{[]float64{
+m := &ExampleModel{[]float64{
   -0.854, 1.067, -1.220, 0.818, -0.749,
@@
   \begin{lstlisting}[framexleftmargin=10pt,numbers=left]
-type Model struct {
+type StreamingModel struct {
   Data chan float64  // data is a channel
@@
 
-func (m *Model) Observe(x []float64) float64 {
+func (m *StreamingModel)
+     Observe(x []float64) float64 {
   ll := Normal.Logps(0, 1, x...)
@@
 \begin{lstlisting}[framexleftmargin=10pt,numbers=left]
-type Model struct {
+type SchoolsModel struct {
   J          int
@@
 
-func (m *Model) Observe(x []float64) float64 {
+func (m *SchoolsModel)
+     Observe(x []float64) float64 {
   mu := x[0]
@@
 \begin{lstlisting}[framexleftmargin=10pt,numbers=left]
-type Model struct {
+type LRModel struct {
   Data  [][]float64
@@
 
-func (m *Model) Observe(x []float64)
+func (m *LRModel) Observe(x []float64)
     float64 {
@@
 // regression parameters.
-func (m *Model) Simulate(x, alpha, beta float64)
+func (m *LRModel) Simulate(x, alpha, beta float64)
     float64 {
@@
 \begin{lstlisting}[framexleftmargin=10pt,numbers=left]
-type Model struct {
+type GMModel struct {
   Data  []float64 // observations
@@
 
-func (m *Model) Observe(x []float64) float64 {
+func (m *GMModel) Observe(x []float64) float64 {
   ll := 0.0
@@
 \begin{lstlisting}[framexleftmargin=10pt,numbers=left]
-type Model struct {
+type LDAModel struct {
   K     int       // num topics
@@
 
-func (m *Model) Observe(x []float64) float64 {
+func (m *LDAModel) Observe(x []float64) float64 {
   ll := 0.0

### 2019-04-29 paper/paper.tex
some fixes

@@
 
-\linespread{1.05}
+\linespread{1.06}
 
@@
 as unnormalized log-likelihood of the posterior distribution.
+The model object usually encapsulates the observations, either
+as a data field or as a reference to a data source.
 Implementation of model methods can be written in virtually
@@
 \begin{figure*}
-  \begin{minipage}{0.45\textwidth}
+\begin{minipage}[t]{0.45\textwidth}
   \begin{lstlisting}[language=Stan,framexleftmargin=10pt,numbers=left]
@@
   \hfill
-  \begin{minipage}{0.45\textwidth}
+\begin{minipage}[t]{0.5\textwidth}
 \begin{lstlisting}[framexleftmargin=10pt,numbers=left]
@@
 
-func (m *SchoolsModel)
-     Observe(x []float64) float64 {
+func (m *SchoolsModel) Observe(x []float64) float64 {
   mu := x[0]
@@
 
-\begin{minipage}{0.49\textwidth}
+\begin{minipage}[t]{0.49\textwidth}
 \centering
@@
 
+  // Regularize the parameter vector
+  ll += Normal.Logps(0, 1, x...)
   // Destructure parameters
   theta := make([][]float64, m.M)
-  m.simplices(&x, m.K, theta)
+  m.FetchSimplices(&x, m.K, theta)
   phi := make([][]float64, m.K)
-  m.simplices(&x, m.V, phi)
+  m.FetchSimplices(&x, m.V, phi)
 
@@
 }
+
+func (m *Model) FetchSimplices(
+	px *[]float64,
+	k int,
+	simplices [][]float64,
+) {
+	for i := range simplices {
+		simplices[i] = make([]float64, k)
+		D.SoftMax(model.Shift(px, k), simplices[i])
+	}
+}
 \end{lstlisting}
@@
 \hfill
-\begin{minipage}{0.49\textwidth}
+\begin{minipage}[t]{0.49\textwidth}
 	\centering
@@
 \end{lstlisting}
-	\vspace{2\baselineskip}
 \end{minipage}

### 2019-04-29 paper/paper.tex
in progress

@@
   one as the language used to program the bulk of the system. 
-  We formulate guidelines for such implementation, and based
+  We formulate guidelines for such implementation, and, based
   on the guidelines, introduce a probabilistic programming
@@
 follow the design pattern of simulation-inference: a significant
-part of the model is a simulator, running an algorithm with
+part of the model is a simulator running an algorithm with
 given parameters; the optimal parameters, or their distribution,
@@
 small but expressive programming language with efficient
-implementation, which has recently become quite popular for
+implementation that has recently become quite popular for
 computation-intensive server-side programming. Infergo is used
@@
   Func func(x []float64) float64,
-  Grad func(grad []float64, x []float64)
-    []float64)
+  Grad func(grad []float64, x []float64))
 \end{lstlisting}
@@
 advanced versions of stochastic gradient descent, as well as
-algorithms of BFGS family, in particular LBFGS~\cite{LN89},
+algorithms of BFGS family, in particular L-BFGS~\cite{LN89},
 providing for a more efficient MAP estimation than stochastic

### 2019-04-29 paper/paper.tex
in progress

@@
 choices may be as well suited for the purpose as the ones we
-made, we feel it important to share our contemplations and
-justify decisions, to which the rest of this section is
-dedicated, from choosing the implementation language, through
-details of automatic differentiation, to support for inference
-on streaming and stochastic observations.
+made, we feel important to share our contemplations and justify
+decisions, to which the rest of this section is dedicated, from
+choosing the implementation language, through details of
+automatic differentiation, to support for inference on streaming
+and stochastic observations.
 
@@
     called in the Go ecosystem), also provided in the
-    standard library, augments parsing and allows to
+	standard library, augments parsing, and allows to
     selectively apply transformation-based automatic
@@
     devices to supercomputers. Go programs can be
-    cross-compiled, and Go programs can be reliably
-    developed and deployed in heterogeneous environments.
+	cross-compiled and reliably developed and deployed in
+	heterogeneous environments.
   \item Go offers efficient parallel execution as a
@@
 adopting Go, with performance-critical parts re-implemented and
-running faster, with better hardware and network utilization,
+running faster due to better hardware and network utilization,
 and simpler and more maintainable code for data processing,
@@
 
-This last method is used in Infergo: the model is written in Go
-and is placed in a separate Go package. A command-line utility
+This latter method is used in Infergo: the model is written in Go
+and is placed in a separate Go package. A command-line tool
 is run on the package to generate a sub-package containing a
@@
 differentiated model are available simultaneously in the calling
-code, so that model methods which do not require computation of
+code, so that model methods that do not require computation of
 gradients can be invoked without the overhead of automatic
@@
 probabilistic programming. However the use of such library, in
-particular, with source code transformation, may make
+particular with source code transformation, may make
 implementation restricted and less efficient: first, if a model
@@
 the most complex model of the three, featuring nested loops and
-dynamic control structure.
+conditionals.
 
@@
 added to most modern general-purpose programming languages, in
-particular, those used for implementation of large-scale software
+particular those used for implementation of large-scale software
 systems, making probabilistic programming inference more

### 2019-04-29 paper/paper.tex
a touch
convinced

@@
 choices may be as well suited for the purpose as the ones we
-made, we feel important to share our contemplations and justify
-decisions, to which the rest of this section is dedicated, from
-choosing the implementation language, through details of
-automatic differentiation, to support for inference on streaming
-and stochastic observations.
+made, we feel it is important to share our contemplations and
+justify decisions, to which the rest of this section is
+dedicated, from choosing the implementation language, through
+details of automatic differentiation, to support for inference
+on streaming and stochastic observations.
 
@@
 \begin{equation}
-	(\alpha, \beta, \sigma)_{MAP} = \arg \max \prod\limits_i \mathrm{Normal}(\alpha + \beta x_i, \sigma)
+	(\alpha, \beta, \sigma)_{MAP} = \arg \max \prod\limits_i \mathrm{pdf}_\mathrm{Normal}(\alpha + \beta x_i, \sigma)
 	\label{eqn:linear-regression-2}
@@
 Dirichlet allocation} is, again, a relatively elaborated model,
-that is, however, not as easy to vectorize.
+which is, however, not as easy to vectorize.
 

### 2019-06-18 paper/paper.tex
minor typos fixed in the paper

@@
 Based on the experience of developing and deploying solutions
-using different probabilistic environments, we came up with
-guidelines to implementation of a probabilistic programming
-facility for server-side applications. We believe that these
-guidelines, when followed, help easier integration of
-probabilistic programming inference into large-scale 
-software systems.
+using different probabilistic programming environments, we came
+up with guidelines to implementation of a probabilistic
+programming facility for server-side applications. We believe
+that these guidelines, when followed, help easier integration of
+probabilistic programming inference into large-scale software
+systems.
 
@@
 overhead. However, even this overhead is avoided when the model
-code is executed outside of inference algorithms: both the
-original and the differentiated model are simultaneously
-available to the rest of the program code, so the methods can be
-called on the differentiated model for inference, and on the
-original model for the most efficient execution with the
-inferred parameters.
+code is executed outside inference algorithms: both the original
+and the differentiated model are simultaneously available to the
+rest of the program code, so the methods can be called on the
+differentiated model for inference, and on the original model
+for the most efficient execution with the inferred parameters.
 

### 2019-06-19 paper/paper.tex
in prorgress
typo fixed, review removed

@@
-\documentclass[sigplan,review,10pt,anonymous]{acmart}
+\documentclass[sigplan,10pt]{acmart}
 \usepackage[utf8]{inputenc}
@@
 
-\linespread{1.06}
+\linespread{1.05}
 
@@
   we argue that a similar facility can be added to any modern
-  general-purpose purpose programming language, giving a
+  general-purpose programming language, giving a
   production system modeling and inference capabilities which

### 2019-07-01 paper/paper.tex
removed copyright
vector elementals

@@
 Functions are considered elementals (and must have a
-registered derivative) if their signature is of kind
-\lstinline{func (float64, float64*) float64}; that is, 
+registered derivative) if their signature is either of kind
+\lstinline{func (float64, float64*) float64}, that is, 
 one or more non-variadic \lstinline{float64} arguments and
-\lstinline{float64} return value. For example, function
-\lstinline{func (float64, float64, float64) float64}
-is considered elemental, while functions
+\lstinline{float64} return value, or \lstinline{func([]float64) float64}.
+For example, functions
 \begin{itemize}
-  \item \lstinline{func (...float64) float64}
-  \item \lstinline{func ([]float64) float64}
-  \item \lstinline{func (int, float64) float64}
+	\item \lstinline{func foo(float64, float64, float64) float64}
+	\item \lstinline{func bar([]float64) float64}
+\end{itemize}
+are considered elemental, while functions
+\begin{itemize}
+  \item \lstinline{func fee(...float64) float64}
+  \item \lstinline{func buz(int, float64) float64}
 \end{itemize}

### 2019-08-11 paper/paper.tex
typo fixed
up-to-date list of library elementals
basic rather than library
changes in progress before camera ready

@@
 program --- inferring parameters of a unidimensional Normal
-distribution. In this example, the model object holds the set of
-observations from which the distribution parameters are to be
-inferred:
+distribution given a set of data samples. The two parameters of
+a Normal distribution are the mean $\mu$ and the standard
+deviation $\sigma$. The solution can of course be obtained
+analytically, but the organization of implementation and
+inference possesses all of the elements of probabilistic
+programming for larger, more complicated problems, for which a
+closed form solution does not exist or difficult to derive.
+
+In this example, the model object holds the set of observations
+from which the distribution parameters are to be inferred:
 \begin{lstlisting}
@@
 parameters, and then conditions the posterior distribution on
-the observations:
+the observations\footnote{A call to \lstinline{Normal.Logps}
+computes the joint log-probability of observations from a Normal
+distribution.}:
 
@@
   ll := Normal.Logps(0, 1, x...)
-  // ... but the posterior is based 
-  // on data observations.
+  // ... but the posterior is based on 
+  // the data.
   ll += Normal.Logps(x[0], math.Exp(x[1]),
@@
 for i := 0; i != 5000; i++ {
-  x = <-samples
+  sample = <-samples
 }
@@
 
+Commonly, elementals are basic mathematical functions.
 Functions are considered elementals (and must have a
@@
 \end{itemize}
-are considered elemental, while functions
+are considered elementals, while functions
 \begin{itemize}
@@
 \lstinline{Exp}, \lstinline{Log}, \lstinline{Pow},
-\lstinline{Sin}, \lstinline{Cos}, \lstinline{Tan}). Auxiliary
-elemental functions with pre-defined gradients are provided in a
-separate package.
+\lstinline{Sin}, \lstinline{Cos}, \lstinline{Tan},
+\lstinline{Erf}, \lstinline{Erfc}). Auxiliary elemental
+functions with pre-defined gradients are provided in a separate
+package.
 
@@
 \end{lstlisting}
-registers the gradient for function
-\lstinline{Sigm}$(x)\coloneq\left(1+e^{-x}\right)^{-1}$.
+registers the gradient for the \textit{sigmoid} function:
+\begin{lstlisting}
+func Sigm(x float64) float64 {
+	return 1. / (1. + math.Exp(-x))
+}
+\end{lstlisting}
 
 \subsection{Inference}
+\label{sec:inference}
 
@@
 \subsection{Model Composition}
+\label{sec:model-composition}
 
@@
 
-\clearpage
-\clearpage
+\section*{Acknowledgements}
+
+The author thanks PUB+ (\url{http://pubplus.com/}) for
+supporting development of Infergo and application of
+probabilistic programming to critical decision-making in
+production enviroment.
 
 \bibliography{refs}
-
-\clearpage
 

### 2019-08-11 paper/paper.tex
intro example of prob prog

@@
 
+\lstdefinelanguage{WebPPL}{
+  keywords={typeof, new, true, false, catch, function, return,
+  null, catch, switch, var, if, in, while, do, else, case,
+  break, condition},
+  ndkeywords={class, export, boolean, throw, implements, import, this},
+  sensitive=false,
+  comment=[l]{//},
+  morecomment=[s]{/*}{*/},
+  morestring=[b]',
+  morestring=[b]"
+}
+
 \lstset{ % add your own preferences
@@
   optimization~\cite{Stan17,BCJ+19}, and others.
+
+\begin{figure}
+\begin{lstlisting}[language=WebPPL]
+var breastCancer = flip(0.01)
+var benignCyst = flip(0.2)
+var positiveMammogram = (breastCancer && flip(0.8))
+                      || (benignCyst && flip(0.5))
+condition(positiveMammogram)
+return breastCancer
+\end{lstlisting}
+	\caption{A probabilistic program in WebPPL~\cite{GS15}:
+	probability of breast cancer given a positive mammogram.}
+	\label{fig:mammo}
+\end{figure}
+  Consider, for example, the probabilistic program in
+  Figure~\ref{fig:mammo}. The program models dependency between
+  a positive mammogram and possible causes --- breast cancer
+  or a benign cyst.  The program defines three Bernoulli
+  random variables  --- \lstinline{breastCancer},
+  \lstinline{benignCyst}, \lstinline{positiveMammogram} ---
+  and conditions on one of the random variables,
+  \lstinline{positiveMammogram}, being true. Inference on the
+  program returns the probability of breast cancer given a
+  positive mammogram.
 
@@
 the hierarchical normal model that often serves as an introductory
-example of Bayesian statistical modeling. Without going into
-details, in this problem the effect of coaching programs on test
-outcomes is compared among eight schools.
+example of Bayesian statistical modeling.  In this problem,
+experiments were performed to estimate the effects of coaching
+programs in each of eight high schools. The outcome variable in
+each study was the increase in test score compared to the same
+test before the program. The average increase and the standard
+deviation of the increase were reported for each school. Based
+on that and on assumption of similarity between schools,
+expected effects are to be estimated.
 Figure~\ref{fig:eight-schools} shows specifications of the same
@@
 The model in Figure~\ref{fig:linear-regression} specifies linear
-regression on unidimensional data: an improper uniform prior is
-placed on the parameters $\alpha$, $\beta$, and $\log \sigma$
-(lines 9--10), which are then conditioned on observations $x, y$
-such that
+regression (approximating linear dependency between input $x$
+and output $y$) on unidimensional data. An improper uniform
+prior is placed on the parameters $\alpha$,
+$\beta$, and $\log \sigma$ (lines 9--10), which are then
+conditioned on observations $x, y$ such that
 \begin{equation}
@@
 of components, where each component comes from a Gaussian
-distribution. The model variant in
-Figure~\ref{fig:gaussian-mixture} accepts a vector of
-single-dimensional observations and the number of components,
-and infers the distribution of parameters of each component
-in the mixture. Following Stan User's Guide~\cite{SDT18},
-component memberships are summed out to make the model
-differentiable.
+distribution. Mixture models are used to represent a population
+consisting of different sub-populations, without indication in
+the data to which sub-population each observation belongs. The
+model variant in Figure~\ref{fig:gaussian-mixture} accepts a
+vector of single-dimensional observations and the number of
+components, and infers the distribution of parameters of each
+component in the mixture. Following Stan User's
+Guide~\cite{SDT18}, component memberships are summed out to make
+the model differentiable.
 \begin{figure}

### 2019-08-11 paper/paper.tex

@@
   we argue that a similar facility can be added to any modern
-  general-purpose programming language, giving a
+  general-purpose purpose programming language, giving a
   production system modeling and inference capabilities which

### 2019-08-12 paper/paper.tex
more introductory examples

--- 2019-08-11 paper/paper.tex
@@
 
+\lstdefinelanguage{Anglican}%
+{morekeywords={*,*1,*2,*3,*agent*,*allow-unresolved-vars*,*assert*,*clojure-version*,*command-line-args*,%
+		*compile-files*,*compile-path*,*e,*err*,*file*,*flush-on-newline*,*in*,*macro-meta*,%
+		*math-context*,*ns*,*out*,*print-dup*,*print-length*,*print-level*,*print-meta*,*print-readably*,%
+		*read-eval*,*source-path*,*use-context-classloader*,*warn-on-reflection*,+,-,->,->>,..,/,:else,%
+		<,<=,=,==,>,>=,@,accessor,aclone,add-classpath,add-watch,agent,agent-errors,aget,alength,alias,%
+		all-ns,alter,alter-meta!,alter-var-root,amap,ancestors,and,apply,areduce,array-map,aset,%
+		aset-boolean,aset-byte,aset-char,aset-double,aset-float,aset-int,aset-long,aset-short,assert,%
+		assoc,assoc!,assoc-in,associative?,atom,await,await-for,await1,bases,bean,bigdec,bigint,binding,%
+		bit-and,bit-and-not,bit-clear,bit-flip,bit-not,bit-or,bit-set,bit-shift-left,bit-shift-right,%
+		bit-test,bit-xor,boolean,boolean-array,booleans,bound-fn,bound-fn*,butlast,byte,byte-array,%
+		bytes,case,cast,char,char-array,char-escape-string,char-name-string,char?,chars,chunk,chunk-append,%
+		chunk-buffer,chunk-cons,chunk-first,chunk-next,chunk-rest,chunked-seq?,class,class?,%
+		clear-agent-errors,clojure-version,coll?,comment,commute,comp,comparator,compare,compare-and-set!,%
+		compile,complement,concat,cond,condp,conj,conj!,cons,constantly,construct-proxy,contains?,count,%
+		counted?,create-ns,create-struct,cycle,dec,decimal?,declare,def,defdist,definline,defm,defmacro,defmethod,%
+		defmulti,defn,defn-,defonce,defquery,defproc,defprotocol,defrecord,defstruct,deftype,delay,delay?,deliver,deref,derive,%
+		descendants,destructure,disj,disj!,dissoc,dissoc!,distinct,distinct?,do,do-template,doall,doc,%
+		dorun,doseq,dosync,dotimes,doto,double,double-array,doubles,drop,drop-last,drop-while,empty,empty?,%
+		ensure,enumeration-seq,eval,even?,every?,false,false?,ffirst,file-seq,filter,finally,find,find-doc,%
+		find-ns,find-var,first,float,float-array,float?,floats,flush,fm,fn,fn?,fnext,for,force,format,future,%
+		future-call,future-cancel,future-cancelled?,future-done?,future?,gen-class,gen-interface,gensym,%
+		get,get-in,get-method,get-proxy-class,get-thread-bindings,get-validator,hash,hash-map,hash-set,%
+		identical?,identity,if,if-let,if-not,ifn?,import,in-ns,inc,init-proxy,instance?,int,int-array,%
+		integer?,interleave,intern,interpose,into,into-array,ints,io!,isa?,iterate,iterator-seq,juxt,%
+		key,keys,keyword,keyword?,last,lazy-cat,lazy-seq,let,letfn,line-seq,list,list*,list?,load,load-file,%
+		load-reader,load-string,loaded-libs,locking,long,long-array,longs,loop,macroexpand,macroexpand-1,%
+		make-array,make-hierarchy,map,map?,mapcat,max,max-key,mem,memfn,memoize,merge,merge-with,meta,%
+		method-sig,methods,min,min-key,mod,monitor-enter,monitor-exit,name,namespace,neg?,new,newline,%
+		next,nfirst,nil,nil?,nnext,not,not-any?,not-empty,not-every?,not=,ns,ns-aliases,ns-imports,%
+		ns-interns,ns-map,ns-name,ns-publics,ns-refers,ns-resolve,ns-unalias,ns-unmap,nth,nthnext,num,%
+		number?,observe,odd?,or,parents,partial,partition,pcalls,peek,persistent!,pmap,pop,pop!,pop-thread-bindings,%
+		pos?,pr,pr-str,predict,prefer-method,prefers,primitives-classnames,print,print-ctor,print-doc,print-dup,%
+		print-method,print-namespace-doc,print-simple,print-special-doc,print-str,printf,println,println-str,%
+		prn,prn-str,promise,proxy,proxy-call-with-super,proxy-mappings,proxy-name,proxy-super,%
+		push-thread-bindings,pvalues,query,quot,rand,rand-int,range,ratio?,rational?,rationalize,re-find,%
+		re-groups,re-matcher,re-matches,re-pattern,re-seq,read,read-line,read-string,recur,reduce,ref,%
+		ref-history-count,ref-max-history,ref-min-history,ref-set,refer,refer-clojure,reify,%
+		release-pending-sends,rem,remove,remove-method,remove-ns,remove-watch,repeat,repeatedly,%
+		replace,replicate,require,reset!,reset-meta!,resolve,rest,resultset-seq,retrieve,reverse,reversible?,%
+		rseq,rsubseq,sample,second,select-keys,send,send-off,seq,seq?,seque,sequence,sequential?,set,set!,%
+		set-validator!,set?,short,short-array,shorts,shutdown-agents,slurp,some,sort,sort-by,sorted-map,%
+		sorted-map-by,sorted-set,sorted-set-by,sorted?,special-form-anchor,special-symbol?,split-at,%
+		split-with,store,str,stream?,string?,struct,struct-map,subs,subseq,subvec,supers,swap!,symbol,symbol?,%
+		sync,syntax-symbol-anchor,take,take-last,take-nth,take-while,test,the-ns,throw,time,to-array,%
+		to-array-2d,trampoline,transient,tree-seq,true,true?,try,type,unchecked-add,unchecked-dec,%
+		unchecked-divide,unchecked-inc,unchecked-multiply,unchecked-negate,unchecked-remainder,%
+		unchecked-subtract,underive,unquote,unquote-splicing,update-in,update-proxy,use,val,vals,%
+		var,var-get,var-set,var?,vary-meta,vec,vector,vector?,when,when-first,when-let,when-not,%
+		while,with-bindings,with-bindings*,with-in-str,with-loading-context,with-local-vars,%
+		with-meta,with-open,with-out-str,with-precision,xml-seq,zero?,zipmap
+	},%
+	   sensitive,% ???
+	      alsodigit=-,%
+		     morecomment=[l];,%
+			    morestring=[b]"%
+				  }[keywords,comments,strings]%
+
 \lstset{ % add your own preferences
@@
   propagation~\cite{MWG+10}, extensions of Sequential Monte
-  Carlo~\cite{WVM14,MYM+15,PWD+14,RNL+2016}, variational
+  Carlo~\cite{WVM14,MYM+15,PWD+14,RNL+2016,MS18}, variational
   inference~\cite{WW13,KTR+17}, gradient-based
@@
   probability of breast cancer given a positive mammogram.}
-	\label{fig:mammo}
+  \label{fig:mammo-webppl}
 \end{figure}
-  Consider, for example, the probabilistic program in
-  Figure~\ref{fig:mammo}. The program models dependency between
+  For example, the probabilistic program in
+  Figure~\ref{fig:mammo-webppl} models dependency between
   a positive mammogram and possible causes --- breast cancer
@@
   positive mammogram.
+
+  \begin{figure}
+    \begin{lstlisting}[language=Anglican]
+(defquery linearRegression [xs ys x*]
+  (let [m     (sample (normal 0 2))
+        b     (sample (normal 0 2))
+        sigma (sample (gamma 1 1))
+        f     (fn [x] (+ (* m x) b))]
+   (loop [data (map #(-> [%1 %2]) xs ys)]
+    (when (seq data)
+     (let [[[x y] & data] data] 
+       (observe (normal (f x) sigma) y)
+       (recur data))))
+   (predict (f x*))))
+    \end{lstlisting}
+    \caption{A probabilistic program in Anglican~\cite{TMY+16}:
+    linear regression.}
+    \label{fig:linreg-anglican}
+  \end{figure}
+  Probabilistic programs may contain loops, conditional
+  statements, recursion, and operate on built-in and user
+  defined data structures. The probabilistic program in
+  Figure~\ref{fig:linreg-anglican} implements linear regression
+  (approximating dependency between the input and the output by
+  a linear function) and contains loops
+  (\lstinline[language=Anglican]{loop/recur}), conditional
+  statements (\lstinline[language=Anglican]{when}), and
+  user-defined functions (\lstinline[language=Anglican]{fn});
+  the data is passed to the program as vectors of real numbers
+  (\lstinline[language=Anglican]{xs}, \lstinline[language=Anglican]{ys}).
 
@@
 program --- inferring parameters of a unidimensional Normal
-distribution given a set of data samples. The two parameters of
-a Normal distribution are the mean $\mu$ and the standard
-deviation $\sigma$. The solution can of course be obtained
-analytically, but the organization of implementation and
-inference possesses all of the elements of probabilistic
-programming for larger, more complicated problems, for which a
-closed form solution does not exist or difficult to derive.
+distribution given a set of data samples, or observations. The
+two parameters of a Normal distribution are the mean $\mu$ and
+the standard deviation $\sigma$. The solution can of course be
+obtained analytically, but the organization of implementation
+and inference possesses all of the elements of probabilistic
+programming for larger, more complex problems, for which a
+closed form solution does not exist or is difficult to derive.
 
@@
   sample = <-samples
+  // ...
 }
@@
          params ...float64) []float64)
-\end{lstlisting} For
-example, the call
+\end{lstlisting}
+\lstinline{RegisterElemental} receives both the parameters and
+the value of the function for efficient computation re-use, and
+the gradient can be computed using either the parameters, the
+value, or a combination of both~\cite{GW08}. For example, the
+call
 \begin{lstlisting}
@@
 \end{lstlisting}
-registers the gradient for the \textit{sigmoid} function:
+registers the gradient for the \textit{sigmoid} function
 \begin{lstlisting}
@@
 \end{lstlisting}
+and re-uses the function value to compute the gradient
+efficiently, avoiding an expensive call to \lstinline{math.Exp}.
 
@@
 
+\clearpage
 \appendix

### 2019-09-23 paper/paper.tex
preparing for camera ready

@@
-\documentclass[sigplan,10pt]{acmart}
+\documentclass[sigplan,10pt,screen]{acmart}
 \usepackage[utf8]{inputenc}
@@
 \usepackage{GoMono}
+\usepackage{flushend}
 
@@
 In line with the guidelines, we have implemented a probabilistic
-programming facility for the Go programming language, Infergo.
-%(\url{http://infergo.ml/}).
-We have chosen Go because Go is a
+programming facility for the Go programming language, Infergo
+(\url{http://infergo.org/}). We have chosen Go because Go is a
 small but expressive programming language with efficient

### 2019-09-24 paper/paper.tex
camera ready

--- 2019-08-12 paper/paper.tex
@@
-\documentclass[sigplan,10pt]{acmart}
+\documentclass[sigplan,screen]{acmart}
 \usepackage[utf8]{inputenc}
@@
 \lstset{ % add your own preferences
+    xrightmargin=1pt,
     frame=single,
@@
 
-\acmConference[Onward! 2019]{SPLASH Onward! 2019}{October 20-25, 2019}{Athens, Greece}
-
-% Copyright
-\setcopyright{none}
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\setcopyright{usgov}
-%\setcopyright{usgovmixed}
-%\setcopyright{cagov}
-%\setcopyright{cagovmixed}
+\setcopyright{acmlicensed}
+\acmPrice{15.00}
+\acmDOI{10.1145/3359591.3359727}
+\acmYear{2019}
+\copyrightyear{2019}
+\acmISBN{978-1-4503-6995-4/19/10}
+\acmConference[Onward! '19]{Proceedings of the 2019 ACM SIGPLAN International Symposium on New Ideas, New Paradigms, and Reflections on Programming and Software}{October 23--24, 2019}{Athens, Greece}
+\acmBooktitle{Proceedings of the 2019 ACM SIGPLAN International Symposium on New Ideas, New Paradigms, and Reflections on Programming and Software (Onward! '19), October 23--24, 2019, Athens, Greece}
+
+ \begin{CCSXML}
+<ccs2012>
+<concept>
+<concept_id>10010147.10010257</concept_id>
+<concept_desc>Computing methodologies~Machine learning</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+<concept>
+<concept_id>10010147.10010341</concept_id>
+<concept_desc>Computing methodologies~Modeling and simulation</concept_desc>
+<concept_significance>100</concept_significance>
+</concept>
+<concept>
+<concept_id>10010147.10010257.10010293</concept_id>
+<concept_desc>Computing methodologies~Machine learning approaches</concept_desc>
+<concept_significance>100</concept_significance>
+</concept>
+<concept>
+<concept_id>10011007.10011006.10011050.10011017</concept_id>
+<concept_desc>Software and its engineering~Domain specific languages</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+<concept>
+<concept_id>10011007.10011074.10011075</concept_id>
+<concept_desc>Software and its engineering~Designing software</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+</ccs2012>
+\end{CCSXML}
+
+\ccsdesc[300]{Computing methodologies~Machine learning}
+\ccsdesc[100]{Computing methodologies~Modeling and simulation}
+\ccsdesc[100]{Computing methodologies~Machine learning approaches}
+\ccsdesc[300]{Software and its engineering~Domain specific languages}
+\ccsdesc[500]{Software and its engineering~Designing software}
 
@@
 \begin{figure}
-  \begin{lstlisting}[framexleftmargin=10pt,numbers=left]
+  \begin{lstlisting}[xleftmargin=11pt,framexleftmargin=10pt,numbers=left]
 type A struct {Data []float64}
@@
 \begin{figure}
-  \begin{lstlisting}[framexleftmargin=10pt,numbers=left]
+  \begin{lstlisting}[xleftmargin=11pt,framexleftmargin=10pt,numbers=left]
 // Exponential distribution.
@@
 // observation.
-func (_ expon) Logp(lambda float64, y float64)
+func (expon) Logp(lambda float64, y float64)
   float64 {
@@
 // of observations.
-func (_ expon) Logps(lambda float64, y ...float64)
+func (expon) Logps(lambda float64, y ...float64)
   float64 {
@@
 \begin{figure}
-  \begin{lstlisting}[framexleftmargin=10pt,numbers=left]
+  \begin{lstlisting}[xleftmargin=11pt,framexleftmargin=10pt,numbers=left]
 type StreamingModel struct {
@@
 \begin{minipage}[t]{0.45\textwidth}
-  \begin{lstlisting}[language=Stan,framexleftmargin=10pt,numbers=left]
+  \begin{lstlisting}[language=Stan,xleftmargin=11pt,framexleftmargin=10pt,numbers=left]
 data {
@@
 \begin{minipage}[t]{0.5\textwidth}
-\begin{lstlisting}[framexleftmargin=10pt,numbers=left]
+\begin{lstlisting}[xleftmargin=11pt,framexleftmargin=10pt,numbers=left]
 type SchoolsModel struct {
@@
 
-func (m *SchoolsModel) Observe(x []float64) float64 {
+func (m *SchoolsModel) Observe(x []float64)
+    float64 {
   mu := x[0]
@@
 \begin{figure}
-\begin{lstlisting}[framexleftmargin=10pt,numbers=left]
+\begin{lstlisting}[xleftmargin=11pt,framexleftmargin=10pt,numbers=left]
 type LRModel struct {
@@
 // regression parameters.
-func (m *LRModel) Simulate(x, alpha, beta float64)
+func (m *LRModel) Simulate(x,alpha,beta float64)
     float64 {
@@
 \begin{figure}
-\begin{lstlisting}[framexleftmargin=10pt,numbers=left]
+\begin{lstlisting}[xleftmargin=11pt,framexleftmargin=10pt,numbers=left]
 type GMModel struct {
@@
 
-// logSumExp computes log(exp(x)+exp(y)) robustly.
+// logSumExp computes log(exp(x)+exp(y))
+// robustly.
 func LogSumExp(x, y float64) float64 {
@@
   }
-  return z + math.Log(math.Exp(x-z)+math.Exp(y-z))
+  return z+math.Log(math.Exp(x-z)+math.Exp(y-z))
 }
@@
 differentiation work.
+
+\begin{figure*}
+\begin{minipage}[t]{0.49\textwidth}
+\centering
+\textbf{Infergo}
+
+\begin{lstlisting}[xleftmargin=11pt,framexleftmargin=10pt,numbers=left]
+type LDAModel struct {
+  K     int       // num topics
+  V     int       // num words
+  M     int       // num docs
+  N     int       // total word instances
+  Word  []int     // word n
+  Doc   []int     // doc for word n
+  Alpha []float64 // topic prior
+  Beta  []float64 // word prior
+}
+
+func (m *LDAModel) Observe(x []float64) float64 {
+  ll := 0.0
+
+  // Regularize the parameter vector
+  ll += Normal.Logps(0, 1, x...)
+  // Destructure parameters
+  theta := make([][]float64, m.M)
+  m.FetchSimplices(&x, m.K, theta)
+  phi := make([][]float64, m.K)
+  m.FetchSimplices(&x, m.V, phi)
+
+  // Impose priors
+  ll += Dirichlet{m.K}.Logps(m.Alpha, theta...)
+  ll += Dirichlet{m.V}.Logps(m.Beta, phi...)
+
+  // Condition on observations
+  gamma := make([]float64, m.K)
+  for in := 0; in != m.N; in++ {
+    for ik := 0; ik != m.K; ik++ {
+      gamma[ik] = 
+        math.Log(theta[m.Doc[in]-1][ik]) +
+          math.Log(phi[ik][m.Word[in]-1])
+    }
+    ll += D.LogSumExp(gamma)
+  }
+  return ll
+}
+
+func (m *Model) FetchSimplices(
+  px *[]float64,
+  k int,
+  simplices [][]float64,
+) {
+  for i := range simplices {
+    simplices[i] = make([]float64, k)
+    D.SoftMax(model.Shift(px, k), simplices[i])
+  }
+}
+\end{lstlisting}
+\end{minipage}
+\hfill
+\begin{minipage}[t]{0.49\textwidth}
+  \centering
+
+  \textbf{Stan}
+
+\begin{lstlisting}[language=Stan]
+data {
+  int<lower=2> K;               // num topics
+  int<lower=2> V;               // num words
+  int<lower=1> M;               // num docs
+  int<lower=1> N;           // total word instances
+  int<lower=1,upper=V> w[N];    // word n
+  int<lower=1,upper=M> doc[N];  // doc for word n
+  vector<lower=0>[K] alpha;     // topic prior
+  vector<lower=0>[V] beta;      // word prior
+}
+
+parameters {
+  simplex[K] theta[M]; // topic dist for doc m
+  simplex[V] phi[K];   // word dist for topic k
+}
+
+model {
+  for (m in 1:M)
+    theta[m] ~ dirichlet(alpha);  // prior
+  for (k in 1:K)
+    phi[k] ~ dirichlet(beta);     // prior
+
+  for (n in 1:N) {
+    real gamma[K];
+    for (k in 1:K)
+      gamma[k] <- log(theta[doc[n],k])
+                + log(phi[k,w[n]]);
+    increment_log_prob(log_sum_exp(gamma));
+  }
+}
+\end{lstlisting}
+\end{minipage}
+\caption{Implementation of the Latent Dirichlet allocation model in Infergo and Stan.}
+\label{fig:lda}
+\end{figure*}
 
@@
 \end{table*}
+
 It is customary to include performance evaluation on benchmark
@@
 studies~(Section~\ref{sec:case-studies}); the code for the third
-model, Latent Dirichlet allocation~\cite{BNG03}, is included in
-Appendix~\ref{app:lda}. The code and data for all models were
+model, Latent Dirichlet allocation~\cite{BNG03}, is shown in
+Figure~\ref{fig:lda}. The code and data for all models were
 originally included in Stan examples and translated for use
@@
 \bibliography{refs}
-
-\clearpage
-\appendix
-
-\begin{figure*}
-\section{Latent Dirichlet Allocation Model}
-  \label{app:lda}
-\vspace{\baselineskip}
-
-\begin{minipage}[t]{0.49\textwidth}
-\centering
-\textbf{Infergo}
-
-\begin{lstlisting}[framexleftmargin=10pt,numbers=left]
-type LDAModel struct {
-  K     int       // num topics
-  V     int       // num words
-  M     int       // num docs
-  N     int       // total word instances
-  Word  []int     // word n
-  Doc   []int     // doc for word n
-  Alpha []float64 // topic prior
-  Beta  []float64 // word prior
-}
-
-func (m *LDAModel) Observe(x []float64) float64 {
-  ll := 0.0
-
-  // Regularize the parameter vector
-  ll += Normal.Logps(0, 1, x...)
-  // Destructure parameters
-  theta := make([][]float64, m.M)
-  m.FetchSimplices(&x, m.K, theta)
-  phi := make([][]float64, m.K)
-  m.FetchSimplices(&x, m.V, phi)
-
-  // Impose priors
-  ll += Dirichlet{m.K}.Logps(m.Alpha, theta...)
-  ll += Dirichlet{m.V}.Logps(m.Beta, phi...)
-
-  // Condition on observations
-  gamma := make([]float64, m.K)
-  for in := 0; in != m.N; in++ {
-    for ik := 0; ik != m.K; ik++ {
-      gamma[ik] = math.Log(theta[m.Doc[in]-1][ik]) +
-        math.Log(phi[ik][m.Word[in]-1])
-    }
-    ll += D.LogSumExp(gamma)
-  }
-  return ll
-}
-
-func (m *Model) FetchSimplices(
-  px *[]float64,
-  k int,
-  simplices [][]float64,
-) {
-  for i := range simplices {
-    simplices[i] = make([]float64, k)
-    D.SoftMax(model.Shift(px, k), simplices[i])
-  }
-}
-\end{lstlisting}
-\end{minipage}
-\hfill
-\begin{minipage}[t]{0.49\textwidth}
-  \centering
-
-  \textbf{Stan}
-
-\begin{lstlisting}[language=Stan]
-data {
-  int<lower=2> K;               // num topics
-  int<lower=2> V;               // num words
-  int<lower=1> M;               // num docs
-  int<lower=1> N;           // total word instances
-  int<lower=1,upper=V> w[N];    // word n
-  int<lower=1,upper=M> doc[N];  // doc for word n
-  vector<lower=0>[K] alpha;     // topic prior
-  vector<lower=0>[V] beta;      // word prior
-}
-
-parameters {
-  simplex[K] theta[M]; // topic dist for doc m
-  simplex[V] phi[K];   // word dist for topic k
-}
-
-model {
-  for (m in 1:M)
-    theta[m] ~ dirichlet(alpha);  // prior
-  for (k in 1:K)
-    phi[k] ~ dirichlet(beta);     // prior
-
-  for (n in 1:N) {
-    real gamma[K];
-    for (k in 1:K)
-      gamma[k] <- log(theta[doc[n],k])
-                + log(phi[k,w[n]]);
-    increment_log_prob(log_sum_exp(gamma));
-  }
-}
-\end{lstlisting}
-\end{minipage}
-\end{figure*}
-
 \end{document}

### 2019-09-24 paper/paper.tex
keywords
caption before

@@
 
+\keywords{probabilistic programming, algorithmic differentiation, bayesian modeling}
+
 \bibliographystyle{ACM-Reference-Format}
@@
 \begin{table*}
+  \caption{Compilation and execution times for 1000 iterations
+  of HMC with 10 leapfrog steps.}
   \begin{tabular}{r | c |  c | c | c | c | c | }
@@
   \end{tabular}
-\caption{Compilation and execution times for 1000 iterations
-  of HMC with 10 leapfrog steps.}
 \label{tab:memory-runtime}

### 2019-10-19 abstract/paper.tex
abstract in progress

@@
 
-\acmConference[LAFI'2019]{Languages for Inference}{January 15,
-2019}{Lisbon, Portugal}
+\acmConference[LAFI'2020]{Languages for Inference}{January 21,
+2020}{New Orleans, LA, USA}
 
@@
 
-\title{Server-Side Probabilistic Programming}
+\title{Probabilistic Programming around Gaussian Processes}
 \author{David Tolpin}
@@
 \begin{abstract}
-	We share our experiences from integrating probabilistic
-	programming into a server-side software system and
-	present our ongoing work on probabilistic programming
-	facility for Go, a modern programming language of choice for
-	server-side software development. We demonstrate how a
-	lightweight efficient probabilistic programming facility can
-	be added to an existing programming language.  Server-side
-	application of probabilistic programming poses challenges
-	for a probabilistic programming system. We discuss the
-	challenges and our experience in overcoming them, and
-	suggest guidelines that can help in a wider adoption of
-	probabilistic programming in server-side software systems.
+	We present GoGP, a library for probabilistic programming around
+	Gaussian processes. Kernels, beliefs about hyperparameters
+	and about observation inputs and outputs are
+	programmatically expressed using the same differentiable
+	probabilistic programming framework. One one hand, a basic
+	usage with maximum-likelihood hyperparameter estimates and
+	homoscedastic Gaussian noise requires as little coding as
+	with any Gaussian process library. On the other hand,
+	imposing prior beliefs on hyperparameters, handling input
+	uncertainty, heteroscedastic and non-Gaussian noise, or
+	change point detection can be organically added on top of a
+	Gaussian process kernel. Just like distributions in
+	probabilistic programs, kernels can be flexibly combined
+	using general code flow rather than through predefined
+	combination operators (usually addition and multiplication).
+	Implemented in Go, a general-purpose programming language 
+	popular for server-side programming, our library is well
+	suited for integration into data processing/machine learning
+	pipelines.  We demonstrate different uses of the library on
+	a variety of case studies inspired by applications.
 \end{abstract}
@@
 \begin{sloppypar}
+	\section{Gaussian Processes in Applications}
+
+	Gaussian processes~\cite{} possess properties that make
+	them the approach of choice in many applications:
+	\begin{itemize}
+		\item A Gaussian process works with as little or as much
+			data as available.
+		\item Non-uniformly sampled observations, missing
+			observations, and observation noise are handled
+			organically.
+		\item Libraries for inference with Gaussian processes
+			are readily available, both standalone and within
+			ubiquitous machine learning libraries, such that
+			Gaussian processes can be included into data
+			processing pipelines and combined with other machine
+			learning algorithms.
+	\end{itemize}
+
 \section{Challenges of Server-Side Probabilistic Programming}

### 2019-10-19 abstract/paper.tex
in progress

@@
 \begin{abstract}
-	We present GoGP, a library for probabilistic programming around
-	Gaussian processes. Kernels, beliefs about hyperparameters
-	and about observation inputs and outputs are
+	We present GoGP, a library for probabilistic programming
+	around Gaussian processes. Kernels, beliefs about
+	hyperparameters and about observation inputs and outputs are
 	programmatically expressed using the same differentiable
@@
 	Implemented in Go, a general-purpose programming language
-	popular for server-side programming, our library is well
-	suited for integration into data processing/machine learning
-	pipelines.  We demonstrate different uses of the library on
-	a variety of case studies inspired by applications.
+	popular for server-side programming, our library is both
+	well suited for integration into data processing/machine
+	learning pipelines and exploits language features such as
+	efficient light-weight computation parallelism for
+	computation efficiency.  We demonstrate different uses of
+	the library on a variety of case studies inspired by
+	applications. 
 \end{abstract}

### 2019-10-19 abstract/paper.tex
lafi abstract in progress

@@
 	programmatically expressed using the same differentiable
-	probabilistic programming framework. One one hand, a basic
+	probabilistic programming framework. On one hand, a basic
 	usage with maximum-likelihood hyperparameter estimates and
@@
 
-\begin{sloppypar}
 \section{Gaussian Processes in Applications}
@@
 		are readily available, both standalone and within
-			ubiquitous machine learning libraries, such that
-			Gaussian processes can be included into data
-			processing pipelines and combined with other machine
-			learning algorithms.
+		machine learning frameworks, such that Gaussian
+		processes can be included into data processing
+		pipelines and combined with other machine learning
+		algorithms.
 \end{itemize}
 
-\section{Challenges of Server-Side Probabilistic Programming}
+Gaussian processes are successfully deployed in many settings,
+in particular when 
+\begin{itemize}
+    \item a stationary kernel is sufficient;
+	\item kernel hyperparameters can be fixed \textit{a priori} or
+		reliably estimated from observations;
+	\item observation noise is constant.
+\end{itemize}
+Most basic cases are indeed covered by the above assumptions,
+and library implementations of Gaussian processes prioritize
+scalability with number of observations, rich libraries
+of kernel functions, and integration of Gaussian processes with
+machine learning pipelines.
 
-Incorporating a probabilistic program, or rather a probabilistic
-procedure, within a larger code body appears to be rather
-straightforward: one implements the model in the probabilistic
-programming language, fetches and preprocesses the data in the
-host programming language, passes the data and the model to an
-inference algorithm, and post-processes the results in the
-host programming language again to make algorithmic
-decisions based on inference outcomes. However, complex
-server-side software systems make integration of probabilistic
-inference challenging.
+There are cases though where the noise varies between
+observations, structured informative priors must be imposed on
+kernel hyperparameters, observation inputs are uncertain, or
+observations are best explained by presence of change
+points~\cite{}.
 
-\paragraph{Simulation vs. inference} Probabilistic models
-often follow a design pattern of simulation-inference: a
-significant part of the model is a simulator, running an
-algorithm with fixed parameters; the optimal parameters, or
-their distribution, are to be inferred. The inferred parameters
-are then used by the software system to execute the simulation
-independently of inference for forecasting and decision making.
+While many of the above cases are addressed in research
+literature, their diversity is beyond feasible support by a
+library, and the practitioner has to implement advanced Gaussian
+process algorithms on case-by-case basis. One, laborious, option
+is to implement the algorithms in a general-purpose programming
+language, along with probabilistic inference for uncertain
+quantities --- hyperparameters, observation inputs, location and
+number of change points.  This option may, in principle, result
+in an efficient and robust implementation, albeit at a high
+implementation cost. Another option is to implement the
+algorithms in a probabilistic programming framework~\cite{},
+which gains support for probabilistic inference on uncertain
+parameters but apparently limits implementation efficiency of
+Gaussian process-specific computations.
 
-This pattern suggests re-use of the simulator: instead of
-implementing the simulator twice, in the probabilistic model and
-in the host environment, the same code can serve both purposes.
-However to achieve this, the host language must coincide with
-the implementation language of the probabilistic model, on one
-hand, and allow a computationally efficient implementation of
-the simulation, on the other hand. Some probabilistic systems
-(Figaro~\cite{P09}, Anglican~\cite{TMY+16}, Turing~\cite{GXG18})
-are built with tight integration with the host environment in
-mind; more often than not though the probabilistic code is
-not trivial to re-use.
+In this work, we advocate a third approach, which combines
+advantages of the other two above. We present an architecture of
+a Gaussian process library that provides an efficient
+implementation of Gaussian process-related computations shared
+by all algorithms, but lets program, virtually without
+restrictions, both Gaussian process kernels and probabilistic
+models of uncertainty, and combine Gaussian processes and
+probabilistic models for inference using algorithms supported by
+the framework. Our implementation of probabilistic
+programming-enabled Gaussian processes, GoGP, is written in Go
+and makes extensive use of modelling and inference capabilities
+provided by Infergo~\cite{}, a Go library for differentiable
+probabilistic programming. In what follows we first outline GoGP
+design  and then illustrate on case studies, the use of GoGP in
+a number of case studies inspired by applications.
 
-\paragraph{Data interface} In a server-side application data
-for inference comes from a variety of sources: network,
-databases, distributed file systems, and in many different
-formats. Efficient inference depends on fast data access and
-updating. Libraries for data access and manipulation are
-available in the host environment. While the host environment
-can be used as a proxy retrieving and transforming the data,
-such as in the case of Stan~\cite{Stan17} integrations,
-sometimes direct access from the probabilistic code is the
-preferred option, for example when the data is streamed or
-retrieved conditionally.
+\section{GoGP --- library for programmable GP inference}
 
-\paragraph{Integration and deployment} Deployment of server-side
-software systems is a delicate process involving automatic
-builds and maintenance of dependencies. Adding  a component,
-which possibly introduces additional software dependencies or
-even a separate runtime, complicates deployment. Minimizing the
-burden of probabilistic programming on the integration
-and deployment process should be a major consideration in
-design or selection of probabilistic programming tools.
-Probabilistic programming systems that are implemented or 
-provide an interface in a popular programming language, e.g.
-Python (Edward~\cite{THS+17}, Pyro~\cite{Pyro18}) are easier
-to integrate and deploy, however the smaller the footprint
-of a probabilistic system, the easier is the adoption.
+\section{Case studies}
 
-\section{Probabilistic Programming Facility for Go}
+\paragraph{Barebones}
 
-Based on the experience of developing and deploying solutions
-using different probabilistic environments, we came up with
-guidelines to implementation of a probabilistic programming
-facility for server-side applications. We believe that these
-guidelines, when followed, help easier integration of
-probabilistic programming inference into large-scale server-side
-software systems.
+\paragraph{Priors on hyperparameters}
 
-\begin{enumerate}
-\item A probabilistic model should be programmed in the host
-programming language. The facility may impose a discipline on
-model implementation, such as through interface constraints, but
-otherwise supporting unrestricted use of the host language for
-implementation of the model.
+\paragraph{Uncertain observation inputs}
 
-\item Built-in and user-defined data structures and libraries
-should be accessible in the probabilistic programming model.
-Inference techniques relying on the code structure, such as
-those based on automatic differentiation, should support the
-use of common data structures of the host language.
+\paragraph{Non-Gaussian noise}
 
-\item The model code should be reusable between inference and
-simulation. The code which is not required solely for inference
-should be written once for both inference of parameters and use
-of the parameters in the host environment.  It should be
-possible to run simulation outside the probabilistic model without
-runtime or memory overhead imposed by inference needs.
-\end{enumerate}
-
-In line with the guidelines, we have implemented a probabilistic
-programming facility for the Go programming language,
-\texttt{infergo} (\url{http://infergo.ml/}). We have chosen Go
-because Go is a small but expressive programming language with
-efficient implementation, which has recently become quite
-popular for computation-intensive server-side programming. At
-the time of writing, the implementation is still at an early
-stage. However, this facility is already used in production
-environment for inference of mission-critical algorithm
-parameters.
-
-A probabilistic model in \texttt{infergo} is an implementation
-of the \texttt{Model} interface requiring a single method
-\texttt{Observe} which accepts a vector (a Go \textit{slice}) of
-floats, the parameters to infer, and returns a single float,
-interpreted as unnormalized log-likelihood of the posterior
-distribution. Implementation of model methods can be written
-in virtually unrestricted Go and use any Go libraries.
-
-For inference, \texttt{infergo} relies on automatic
-differentiation. The source code of the model is
-translated by a command-line tool provided by \texttt{infergo}
-into an equivalent model with reverse-mode automatic
-differentiation of the log-likelihood with respect 
-to the parameters applied. The differentiation operates
-on the built-in floating-point type and incurs only a small
-computational overhead. However, even this overhead is avoided
-when the model code is executed outside of inference algorithms:
-both the original and the differentiated model are
-simultaneously available to the rest of the program code, so
-the methods can be called on the differentiated model for
-inference, and on the original model for the most efficient
-execution with the inferred parameters.
-
-The Go programming language and development environment offer
-capabilities which made implementation of \texttt{infergo}
-affordable.
-
-\begin{itemize}
-    \item The Go parser and abstract syntax tree serializer are
-        a part of the standard library. Parsing, transforming,
-        and generating Go source code is straightforward and
-        effortless.
-	\item Type inference (or \textit{type checking} as it is
-		called in the Go ecosystem), also provided in the
-		standard library, augments parsing and allows to
-		selectively apply transformation-based automatic
-		differentiation  based on static expression types. 
-	\item Go compiles and runs fast. Fast compilation and
-		execution speeds allow to use the same facility for both
-		exploratory design of probabilistic models and for
-		inference in production environment.
-	\item Go offers efficient parallel execution as a
-		first-class feature, via so-called \textit{goroutines}.
-		Goroutines streamline implementation of sampling-based
-		inference algorithms. Sample generators and consumers
-		are run in parallel, communicating through channels. 
-		Inference is easy to parallelize in order to exploit
-		hardware multi-processing, and samples are retrieved
-		lazily for postprocessing. 
-\end{itemize}
-
-Table~\ref{tab:memory-runtime} provides memory consumption and
-running time measurements on basic models to
-illustrate  \texttt{infergo}'s performance.
-The measurements were obtained on a 2.3GHz Intel Core 5
-CPU with 8GB of memory for 1000 iterations of Hamiltonian Monte
-Carlo with 10 leapfrog steps. Note that log-likelihood computation 
-for standard distributions is not optimized yet. Quite the
-opposite: since models in \texttt{infergo} are fully composable,
-primitive distributions are themselves implemented as
-\texttt{infergo} models and automatically differentiated.
-
-{\smaller
-\begin{table}[H]
-	\caption{Memory and running times for 1000 iterations
-	of HMC with 10 leapfrog steps.}
-	\label{tab:memory-runtime}
-	\begin{tabular}{r | c |  c | c}
-		{\it model}  & \multicolumn{2}{c|}{\it time} & {\it memory} \\ 
-		 & {\it compilation} & {\it execution} & \\\hline
-		8 schools & 0.15s & 0.6s & 5.5MB \\
-		10D normal, 100 points & 0.15s & 3.0s & 5.7MB \\
-		50D normal, 100 points & 0.15s & 13.0s & 5.8MB 
-	\end{tabular}
-\end{table}}
-
-A lightweight probabilistic programming facility similar to
-\texttt{infergo} can be added to most modern general-purpose
-programming languages, in particular those used in implementing
-large-scale software systems, making probabilistic
-programming inference more accessible in server-side
-applications.
-\end{sloppypar}
+\section{Discussion}
 

### 2019-10-21 abstract/paper.tex
lafi abstract in progress

@@
 \usepackage{hyperref}
+\usepackage{listings}[language=Golang]
+\usepackage{GoMono}
 
-\linespread{1.03}
+\lstdefinelanguage{Golang}%
+  {morekeywords=[1]{package,import,func,type,struct,return,defer,panic,%
+     recover,select,var,const,iota,},%
+   morekeywords=[2]{string,uint,uint8,uint16,uint32,uint64,int,int8,int16,%
+     int32,int64,bool,float32,float64,complex64,complex128,byte,rune,uintptr,%
+     error,interface},%
+   morekeywords=[3]{map,slice,make,new,nil,len,cap,copy,close,true,false,%
+     delete,append,real,imag,complex,chan,},%
+   morekeywords=[4]{for,break,continue,range,go,goto,switch,case,fallthrough,if,%
+     else,default,},%
+   morekeywords=[5]{Println,Printf,Error,Print,},%
+   sensitive=true,%
+   morecomment=[l]{//},%
+   morecomment=[s]{/*}{*/},%
+   morestring=[b]',%
+   morestring=[b]",%
+   morestring=[s]{`}{`},%
+}
 
@@
 probabilistic programming. In what follows we first outline GoGP
-design  and then illustrate on case studies, the use of GoGP in
-a number of case studies inspired by applications.
+design  and then illustrate the use of GoGP on a number of case
+studies inspired by applications.
 
 \section{GoGP --- library for programmable GP inference}
+
+GoGP exports a type, \lstinline{GP}, implementing computations
+involved in any use of a Gaussian process. The type has four
+methods:
+
+\begin{itemize}
+	\item \lstinline{Absorb} --- adds (absorbs) observations;
+	\item \lstinline{Produce} --- predicts out distributions;
+	\item \lstlinline{Observe} --- computes log-likelihood of
+		kernel hyperparameters given observations;
+	\item \lstinline{Gradient} --- returns the gradient of
+		log-likelihood by hyperparameters, and optionally
+		by observations.
+\end{itemize}
+
+Methods \lstinline{Observe} and \lstinline{Gradient} also turn
+\lstinline{GP} into an Infergo model with a custom gradient,
+suitable for passing to Infergo inference algorithms.
+
+package gp // import "bitbucket.org/dtolpin/gogp/gp"
+
+TYPES
+
+type GP struct {
+	NDim                   int       // dimensions
+	Simil, Noise           Kernel    // kernel
+	ThetaSimil, ThetaNoise []float64 // kernel parameters
+
+	// inputs
+	X [][]float64 // inputs, for computing covariances
+	Y []float64   // outputs
+
+	// Has unexported fields.
+}
+    Type GP is the barebone implementation of GP.
+
+func (gp *GP) Absorb(x [][]float64, y []float64) (err error)
+    Absorb absorbs observations into the process.
+
+func (gp *GP) Gradient() []float64
+    Gradient computes the gradient of the log-likelihood with respect to the
+    parameters and the inputs (GPML:5.9):
+
+    ∇L = ½ tr((α α^⊤ - Σ^−1) ∂Σ/∂θ), where α = Σ^-1 y
+
+func (gp *GP) LML() float64
+    LML computes log marginal likelihood of the kernel given the absorbed
+    observations (GPML:5.8):
+
+    L = −½ log|Σ| − ½ y^⊤ α − n/2 log(2π), where α = Σ^-1 y
+
+func (gp *GP) Observe(x []float64) float64
+    Observe computes log marginal likelihood of the parameters given the
+    observations. The argument is contatenation of log-transformed
+    hyperparameters, inputs, and outputs.
+
+    Optionally, the input can be only log-transformed hyperparameters, and then
+    * only hyperparameters are inferred; * inputs must be assigned to fields X,
+    Y of gp.
+
+func (gp *GP) Produce(x [][]float64) (
+	mu, sigma []float64,
+	err error,
+)
+    Produce computes predictions.
+
+type Kernel interface {
+	model.Model
+	NTheta() int
+}
+    Type Kernel is the kernel interface, implemented by both covariance and
+    noise kernels.
+
+GoGP
 

### 2019-10-21 abstract/paper.tex
drafted, references missing

--- 2019-10-19 abstract/paper.tex
@@
 \usepackage{hyperref}
+\usepackage{listings}[language=Golang]
+\usepackage{GoMono}
 
-\linespread{1.03}
+\lstdefinelanguage{Golang}%
+  {morekeywords=[1]{package,import,func,type,struct,return,defer,panic,%
+     recover,select,var,const,iota,},%
+   morekeywords=[2]{string,uint,uint8,uint16,uint32,uint64,int,int8,int16,%
+     int32,int64,bool,float32,float64,complex64,complex128,byte,rune,uintptr,%
+     error,interface},%
+   morekeywords=[3]{map,slice,make,new,nil,len,cap,copy,close,true,false,%
+     delete,append,real,imag,complex,chan,},%
+   morekeywords=[4]{for,break,continue,range,go,goto,switch,case,fallthrough,if,%
+     else,default,},%
+   morekeywords=[5]{Println,Printf,Error,Print,},%
+   sensitive=true,%
+   morecomment=[l]{//},%
+   morecomment=[s]{/*}{*/},%
+   morestring=[b]',%
+   morestring=[b]",%
+   morestring=[s]{`}{`},%
+}
+
+\lstset{ % add your own preferences
+    xrightmargin=1pt,
+    frame=single,
+    basicstyle=\ttfamily\footnotesize,
+    numbers=none,
+    numbersep=4pt,
+    showstringspaces=false,
+    tabsize=4,
+    language=Golang % this is it !
+}
 
@@
 the framework. Our implementation of probabilistic
-programming-enabled Gaussian processes, GoGP, is written in Go
+programming-enabled Gaussian processes, GoGP
+(\url{http://bitbucket.org/dtolpin/gogp}), is written in Go
 and makes extensive use of modelling and inference capabilities
@@
 probabilistic programming. In what follows we first outline GoGP
-design  and then illustrate on case studies, the use of GoGP in
-a number of case studies inspired by applications.
+design  and then illustrate the use of GoGP on a number of case
+studies inspired by applications.
 
-\section{GoGP --- library for programmable GP inference}
+\section{GoGP --- Gaussian Process Library for Programmable Inference}
+
+GoGP is built around a dual view on the Gaussian process: as a
+stochastic process and as a probabilistic model with respect
+to kernel hyperparameters and, optionally, observations.
+\lstinline{GP}, the type encapsulating a Gaussian process,
+provides methods \lstinline{Absorb} and \lstinline{Produce} to 
+add observations to the process and predicting the distribution
+of outputs at given locations. In addition, \lstinline{GP}
+provides methods \lstinline{Observe} and \lstinline{Gradient}
+that turn it into an Infergo model (with a custom gradient).
+Inference can be performed on a \lstinline{GP} instance alone
+or on a model that combines a \lstinline{GP} instance with
+beliefs about hyperparameters and observations.  The
+log-likelihood gradient is computed with respect to
+hyperparameters only or with respect to both hyperparameters and
+observations, supporting both basic and advanced usage.
+
+Common computationally-intensive algorithms are implemented in 
+methods of \lstinline{GP} and rely on efficient parallelism of
+the Go language and a highly optimized library for mathematical 
+computations Gonum~\cite{}. Problem-specific code comprises two
+parts --- a kernel and optionally a probabilistic model --- that
+are both automatically differentiated by Infergo.  As a minimum,
+a \lstinline{GP} instance must be initialized with a kernel, an
+object of type implementing an Infergo model and computing the
+covariance between two locations. A few basic kernels are
+provided in GoGP as a convenience, and can be combined in derived
+kernels using virtually unrestricted Go code.
+
+% For example, here is an RBF (Normal) kernel as implemented in
+% GoGP (but could also be supplied by the library user):
+% \begin{lstlisting}
+% type normal struct{}
+% 
+% func (k normal) Observe(x []float64) float64 {
+% 	return k.Cov(x[0], x[1], x[2])
+% }
+% func (normal) NTheta() int { return 1 }
+% func (normal) Cov(l, xa, xb float64) float64 {
+% 	d := (xa - xb) / l
+% 	return math.Exp(-d * d / 2)
+% }
+% \end{lstlisting}
 
@@
 
-\paragraph{Barebones}
+GoGP includes case studies, illustrating, on simple examples,
+common patterns of GoGP use. We briefly summarize here some
+of the case studies.
+
+\paragraph{Basic usage}
+
+In the basic case, similar to that supported by many Gaussian
+process libraries, a GP directly serves as the model for
+inference on hyperparameters (or the hyperparameters can be just
+fixed).
+
+The library user specifies the kernel:
+\begin{lstlisting}
+type Basic struct{}
+func (Basic) Observe(x []float64) float64 {
+    return x[0] * kernel.Normal.Observe(x[1:])
+}
+func (Basic) NTheta() int { return 2 }
+\end{lstlisting}
+and initializes \lstinline{GP} with a kernel instance:
+\begin{lstlisting}
+gp := &gp.GP{
+    NDim:  1,
+    Simil: &Basic{},
+}
+\end{lstlisting}
+
+MLE inference on hyperparameters and prediction can then be performed
+through library functions.
 
@@
 
+If priors on hyperparameters are to be specified, the library
+user provides both the kernel and the model.  \lstinline{GP} is
+initialized with the kernel, and then \lstinline{GP} and the
+model are combined for inference in a derived model:
+\begin{lstlisting}
+type Model struct {
+    gp           *gp.GP
+    priors       *Priors
+    gGrad, pGrad []float64
+}
+func (m *Model) Observe(x []float64) float64 {
+    var gll, pll float64
+    gll, m.gGrad = \
+      m.gp.Observe(x), model.Gradient(m.gp)
+    pll, m.pGrad = \
+      m.priors.Observe(x),model.Gradient(m.priors)
+    return gll + pll
+}
+func (m *Model) Gradient() []float64 {
+	for i := range m.pGrad {
+		m.gGrad[i] += m.pGrad[i]
+	}
+	return m.gGrad
+}
+\end{lstlisting}
+In \lstinline{Model}, \lstinline{gp} holds a \lstinline{GP}
+instance and \lstinline{priors} holds an instance of the model
+expressing beliefs about hyperparameters. A \lstinline{Model}
+instance is used for inference on hyperparameters, a
+\lstinline{GP} instance --- for prediction.
+
 \paragraph{Uncertain observation inputs}
+
+When observation inputs are uncertain, beliefs about inputs can
+be specified, and the log-likelihood gradient can be computed
+with respect to both hyperparameters and observation inputs. For
+example, in a time series, one can assume that observation
+inputs come from a renewal process and let the inputs move 
+relative to each other. Then, forecasting can be performed 
+relative to posterior observation inputs.
 
@@
 
-\section{Discussion}
+In basic usage, the observation noise is assumed to be Gaussian.
+This usage is supported by initializing \lstinline{GP} with a noise
+kernel, along with a similarity kernel. When the noise is not
+Gaussian, an analytical solution for posterior Gaussian process
+inference does not always exist. However, non-Gaussian noise is
+straightforwardly supported by GoGP through supplying a
+model for beliefs on observation outputs:
+\begin{lstlisting}
+type Noise struct {
+	Y []float64 // noisy outputs
+}
+func (m *Noise) Observe(x []float64) float64 {
+	// Laplacian noise
+	for i := range m.Y {
+		ll += Expon.Logp(1/math.Exp(x[s]),
+			math.Abs(m.Y[i]-x[i0+n+i]))
+	}
+	return ll
+}
+\end{lstlisting}
 

### 2019-10-21 abstract/paper.tex
drafted, typo fixed

@@
     learning pipelines and exploits language features such as
-    efficient light-weight computation parallelism for
-    computation efficiency.  We demonstrate different uses of
-    the library on a variety of case studies inspired by
-    applications. 
+	efficient light-weight parallelism for computation
+	efficiency.  We demonstrate different uses of the library on
+	a variety of case studies inspired by applications. 
 \end{abstract}
@@
 scalability with number of observations, rich libraries
-of kernel functions, and integration of Gaussian processes with
-machine learning pipelines.
+of combinable kernel functions, and integration of Gaussian
+processes with machine learning pipelines.
 
-There are cases though where the noise varies between
+There are cases though where noise varies between
 observations, structured informative priors must be imposed on
@@
 provides methods \lstinline{Absorb} and \lstinline{Produce} to 
-add observations to the process and predicting the distribution
+add observations to the process and to predict the distribution
 of outputs at given locations. In addition, \lstinline{GP}
@@
     NDim:  1,
-    Simil: &Basic{},
+    Simil: Basic{},
 }
@@
 	for i := range m.Y {
-		ll += Expon.Logp(1/math.Exp(x[s]),
-			math.Abs(m.Y[i]-x[i0+n+i]))
+		ll += Expon.Logp(1/math.Exp(x[0]),
+			math.Abs(m.Y[i]-x[1+i]))
 	}

### 2019-10-22 abstract/paper.tex
with bibrefs

@@
 
-Gaussian processes~\cite{} possess properties that make
+Gaussian processes~\cite{RW05} possess properties that make
 them the approach of choice in many applications:
@@
 Most basic cases are indeed covered by the above assumptions,
-and library implementations of Gaussian processes prioritize
-scalability with number of observations, rich libraries
-of combinable kernel functions, and integration of Gaussian
-processes with machine learning pipelines.
+and library implementations~\cite{MWN+17,GPB+18,PVG+11} of
+Gaussian processes prioritize scalability with number of
+observations, rich libraries of combinable kernel functions, and
+integration of Gaussian processes with machine learning
+pipelines.
 
@@
 observations are best explained by presence of change
-points~\cite{}.
+points~\cite{ROE+13,SGR04,GOR09,MR11,STR10}.
 
@@
 implementation cost. Another option is to implement the
-algorithms in a probabilistic programming framework~\cite{},
+algorithms in a probabilistic programming
+framework~\cite{TMY+16,Stan17,GXG18},
 which gains support for probabilistic inference on uncertain
@@
 and makes extensive use of modelling and inference capabilities
-provided by Infergo~\cite{}, a Go library for differentiable
+provided by Infergo~\cite{T19}, a Go library for differentiable
 probabilistic programming. In what follows we first outline GoGP
@@
 the Go language and a highly optimized library for mathematical 
-computations Gonum~\cite{}. Problem-specific code comprises two
+computations Gonum~\cite{Gonum}. Problem-specific code comprises two
 parts --- a kernel and optionally a probabilistic model --- that
@@
 
-When observation inputs are uncertain, beliefs about inputs can
+When observation inputs are uncertain~\cite{MR11}, beliefs about inputs can
 be specified, and the log-likelihood gradient can be computed
@@
 kernel, along with a similarity kernel. When the noise is not
-Gaussian, an analytical solution for posterior Gaussian process
+Gaussian~\cite{MWN+17}, an analytical solution for posterior Gaussian process
 inference does not always exist. However, non-Gaussian noise is

### 2019-10-22 abstract/paper.tex
a couple typos fixed

@@
 }
-func (m *Noise) Observe(x []float64) float64 {
+func (m *Noise) Observe(x []float64) (ll float64){
 	// Laplacian noise

### 2019-10-24 abstract/paper.tex
better language

@@
 models of uncertainty, and combine Gaussian processes and
-probabilistic models for inference using algorithms supported by
-the framework. Our implementation of probabilistic
-programming-enabled Gaussian processes, GoGP
-(\url{http://bitbucket.org/dtolpin/gogp}), is written in Go
-and makes extensive use of modelling and inference capabilities
+probabilistic models for inference by means of inference
+algorithms supported by the framework. Our implementation of
+probabilistic programming-enabled Gaussian processes, GoGP
+(\url{http://bitbucket.org/dtolpin/gogp}), is written in Go and
+makes extensive use of modelling and inference capabilities
 provided by Infergo~\cite{T19}, a Go library for differentiable
